### 2019-01-20 paper.tex
outline of paper

\documentclass{IEEEtran}
\usepackage[hyphens]{url}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{colortbl} % for \rowcolor
\usepackage{caption}
\usepackage{subcaption}
\usepackage[bookmarks=false]{hyperref}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}

\newcommand{\todo}[1]{\marginpar{\parbox{18mm}{\flushleft\tiny\color{red}\textbf{TODO}:
      #1}}}
\definecolor{headcolor}{gray}{0.9}

\begin{document}
\title{SPA: An Apache Spark Pilot scheduler for HPC applications}
\author{
    \IEEEauthorblockN{
        Val\'erie Hayot-Sasson and Tristan Glatard
    }
    \IEEEauthorblockA{}
}
\maketitle

\begin{abstract}
    \todo{complete}
\end{abstract}

\section{Introduction}
\section{Materials and Methods}\label{sec:methods}
    \todo{walltime parameter of workflow}
    \subsection{Added value of pilot scheduling}
        \todo{execution of a batch cluster vs dynamic (pilot) cluster}
    \subsection{Robust masters}
        \todo{scala standalone vs our pyspark workaround. kill masters in experiments}
    \subsection{Checkpointing}
        \todo{metric for determining how often to checkpoint based on cluster size}
    \subsection{Job arrays}
        \todo{need to kill idle workers. may not want all workers to be running at once.}
    \subsection{Example application}
        \todo{incrementation with varying task durations}
\section{Discussion}\label{sec:discussion}
\section{Conclusion}\label{sec:conclusion}

\end{document}

### 2019-01-22 example/dummyscala/src/main/scala/IncrementApp.scala
start of scala incrementation program

@@ -1,0 +1,70 @@
+/* IncrementApp.scala */
+import org.apache.spark.SparkContext
+import org.apache.spark.SparkContext._
+import org.apache.spark.SparkConf
+import org.apache.spark.input.PortableDataStream
+import java.io.File
+import java.io.DataInputStream
+import com.ericbarnhill.niftijio.NiftiVolume
+import com.ericbarnhill.niftijio.NiftiHeader
+
+object IncrementApp {
+  val usage = """
+    Usage: increment bb_dir output_dir iterations [--delay] [--work_dir]
+  """
+
+  def readImg( filename:String, data:PortableDataStream ) : Byte = {
+    val niftibytes = data.open()
+    val nifti = NiftiHeader.read(niftibytes, filename)
+    niftibytes.close()
+    return data.toArray()(2)
+  }
+
+  def main(args: Array[String]) {
+    if (args.length == 0 || args.length < 3) { 
+      println(usage)
+      sys.exit(1)
+    }
+    val arglist = args.toList
+    type OptionMap = Map[Symbol, Any]
+
+    def nextArgument(map: OptionMap, list: List[String]) : OptionMap = {
+      list match {
+        case Nil => map
+        case "--work_dir" :: string :: tail =>
+                                nextArgument(map ++ Map('work_dir -> string), tail)
+        case "--delay" :: value :: tail =>
+                                nextArgument(map ++ Map('delay -> value.toFloat), tail)
+        case string :: tail if arglist.size - tail.size == 1  =>
+                                nextArgument(map ++ Map('bb_dir -> string), tail)
+        case string :: tail if arglist.size - tail.size == 2 =>
+                                nextArgument(map ++ Map('output_dir -> string), tail)
+        case string :: tail if arglist.size - tail.size == 3 =>
+                                nextArgument(map ++ Map('iterations -> string.toInt), tail)
+        case option :: tail => println("Unknown option "+option)
+                               sys.exit(1)
+      }
+    }
+    val options = nextArgument(Map(), arglist)
+    
+    val conf = new SparkConf().setAppName("Scala incrementation")
+    val sc = new SparkContext(conf)
+
+    val delay = if ((options get 'delay).isEmpty) 0 else (options get 'delay).get
+    val of = new File(options('output_dir).asInstanceOf[String])
+    val output_dir = of.getAbsolutePath()
+    of.mkdirs()
+    println(output_dir)
+
[…]

### 2019-02-24 example/standalone/hpc/hpc_batch_template.sh
added benchmarking

@@ -1,8 +1,7 @@
 #!/bin/bash
 
-spscript="--jars /home/vhayots/projects/def-glatard/vhayots/niftijio/target/scala-2.12/*.jar /home/vhayots/projects/def-glatard/vhayots/spa/example/scala_increment/target/scala-2.11/increment-app_2.11-1.0.jar /scratch/vhayots/splits /scratch/vhayots/scalaout 1"
-
 echo start $(date +%s.%N) > $mstr_bench
+echo $mstr_bench > singlenodes.out
 
 module load spark/2.3.0
 
@@ -10,6 +9,7 @@ export SPARK_IDENT_STRING=$SLURM_JOBID
 export SPARK_WORKER_DIR=$SLURM_TMPDIR
 export SLURM_SPARK_MEM_FLOAT=$(echo "${SLURM_MEM_PER_NODE} * 0.95" | bc)
 export SLURM_SPARK_MEM=${SLURM_SPARK_MEM_FLOAT%.*}
+echo ${SLURM_SPARK_MEM}
 
 start-master.sh
 while [ -z "$MASTER_URL" ]
@@ -19,7 +19,7 @@ do
 	sleep 5
 done
 
-NWORKERS=$((SLURM_NTASKS))
+NWORKERS=$((SLURM_NTASKS - 1))
 
 SPARK_NO_DAEMONIZE=1 srun -n ${NWORKERS} -N ${NWORKERS} --label --output=$SPARK_LOG_DIR/spark-%j-workers.out start-slave.sh -m ${SLURM_SPARK_MEM}M -c ${SLURM_CPUS_PER_TASK} ${MASTER_URL} &
 slaves_pid=$!

### 2019-02-24 example/standalone/hpc/hpc_pilot_template.sh

@@ -86,7 +86,7 @@ start_time=0
 
 while true; do
     cores_in_use=$([[ $(curl -s $WORKER_UI | jq -r ".coresused") == "0" ]] && { echo false; } || { echo true; })
-    executors_complete=$([[ $(curl -s $WORKER_UI | jq -r ".finishedexecutor" | jq "if length = 0 then 'true' else 'false' end") == "true" ]] && { echo false; } || { echo true; })
+    executors_complete=$([[ $(curl -s $WORKER_UI | jq -r ".finishedexecutor" | jq 'if length = 0 then "true" else "false" end') == "true" ]] && { echo false; } || { echo true; })
 
     echo 'executor running'
     echo "cores in use: " $cores_in_use

### 2019-02-24 pilotspark.py

@@ -89,12 +89,18 @@ def configure(conf, job_id, rand_hash):
 
 
 def submit_sbatch(template, conf):
-    print(time.time())
+
+    if "benchmark" in conf:
+        with open(conf["benchmark"], 'a+') as f:
+            f.write(os.linesep + datetime.now().isoformat() +
+                    '\t' + str(time.time()))
+
     submit_func = "sbatch"
     rand_hash = "" #gen_hash(template)
     job_id = '${SLURM_JOB_ID}'
     configure(conf, job_id, rand_hash)
     s = Slurm(conf["name"], conf["SLURM_CONF_GLOBAL"])
+    conf["DRIVER"]["mstr_bench"] = conf["COMPUTE"]["mstr_bench"]
     job_id = s.run(template, cmd_kwargs=conf["DRIVER"], _cmd=submit_func)
 
     job_id = str(job_id)
@@ -110,14 +116,17 @@ def submit_sbatch(template, conf):
         queue = [l.split(' ')[0] for l in out if l.split(' ') != '']
 
         condition = job_id in queue
+        time.sleep(5 * 60)
 
-    print(time.time())
+    if "benchmark" in conf:
+        with open(conf["benchmark"], 'a+') as f:
+            f.write('\t' + str(time.time()))
 
 
 def submit_locally(template, conf):
 
     submit_func = "bash"
-    rand_hash = gen_hash(template) 
+    rand_hash = gen_hash(template)
     job_id = ""
 
     configure(conf, job_id, rand_hash)
@@ -138,8 +147,11 @@ def submit_locally(template, conf):
 
 
 def submit_pilots(template, conf):
-
-    print(time.time())
+    
+    if "benchmark" in conf:
+        with open(conf["benchmark"], 'a+') as f:
+            f.write(os.linesep + datetime.now().isoformat() + '\t' + 
+                    str(time.time()))
     submit_func = "sbatch"
     rand_hash = gen_hash(template)
     slurm_job_id = '${SLURM_JOB_ID}'
@@ -237,15 +249,27 @@ def submit_pilots(template, conf):
                                            template, rand_hash, "sbatch", jobs)
 
             time.sleep(5 * 60)
-            r = requests.get(driver_rest)
-            driver_api = r.json()
-            print(driver_api)
 
-        p = Popen(["scancel"].extend(jobs), stdout=PIPE, stderr=PIPE)
-        (out, err) = p.communicate()
+            try:
+                r = requests.get(driver_rest)
+                driver_api = r.json()
[…]

### 2019-02-24 paper.tex
added abstract

@@
 \begin{abstract}
-    \todo{complete}
+    With the recent influx of neuroimaging Big Data, many researchers are moving
+    towards adopting Big Data processing engines such as Apache Spark 
+    to process the data. These processing engines, however, have been primarily 
+    designed to work on dedicated infrastructure and are not well designed to 
+    work on infrastructure typically available to researchers, such as 
+    High-Performance Computing (HPC) infrastructure. Typical mechanisms for 
+    launching Spark-based applications on HPC clusters involve batch-requesting
+    of the resources which can greatly increase queueing times of the Big Data 
+    applications. Pilot-scheduling techniques is a well known HPC scheduling
+    strategy that begins scheduling as compute resources are allocated to the 
+    user, thereby reducing HPC scheduling wait times. Here, we introduce SPA, an
+    open-source
+    pilot-job submission system designed for launching Apache Spark pipelines 
+    on Slurm-based clusters. SPA leverages optimizations available in Spark, 
+    such as robust workers, masters and driver, in addition to checkpointing, 
+    to ensure that processing occurs even when nodes are lost due to timeout. 
+    SPA also ensures that a user-specified number of pilots (nodes) have been 
+    requested on the cluster in order to ensure a constant level of parallelism.
+    Should pilots be idle for a certain amount of time, they are terminated by 
+    the system.
+
 \end{abstract}

### 2019-02-26 paper.tex

--- 2019-01-20 paper.tex
@@
 \hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
+
+\newcommand{\tristan}[1]{\color{red}\textbf{Note from Tristan}:
+      #1 \color{black}}
 
@@
 \begin{abstract}
-    \todo{complete}
+    Big Data processing engines such as Apache Spark are increasingly 
+    used to process scientific data. These processing engines, however, 
+    have been primarily designed to work on dedicated infrastructure 
+    and are not well designed to work on High-Performance Computing 
+    (HPC) systems commonly available to researchers. Typical mechanisms 
+    for launching Spark-based applications on HPC clusters involve 
+    batch-requesting of the resources which can greatly increase 
+    queueing times of the Big Data 
+    applications. Pilot-scheduling is a well known HPC scheduling
+    strategy that begins scheduling as compute resources are allocated to the 
+    user, thereby reducing HPC scheduling wait times. Here, we introduce SPA, an
+    open-source
+    pilot-job submission system designed for launching Apache Spark pipelines 
+    on Slurm-based clusters. SPA leverages optimizations available in Spark, 
+    such as robust workers, masters and driver, in addition to 
+    checkpointing, to ensure that processing occurs even when nodes are 
+    lost due to expired walltime. SPA also ensures that a 
+    user-specified number of pilots is maintained in the cluster in 
+    order to ensure a constant level of parallelism. Should pilots be 
+    idle for a certain amount of time, they are terminated by 
+    the system. \tristan{Add a sentence on results when you have them.}
+
 \end{abstract}

### 2019-02-26 paper.tex
some intro

--- 2019-02-24 paper.tex
@@
 \section{Introduction}
+
+In the recent years, neuroimaging Big Data has become more accessible, largely 
+attributed to the growing number of open-data initiatives. However, the
+processing of such large datasets using standard neuroimaging pipelines has 
+become unfeasable. As a result, many researchers are moving towards Big Data 
+engines, such as Apache Spark and Dask, to create their processing pipelines. 
+Unlike traditional neuroimaging engines, big data engines were designed to be 
+executed on commodity 
 \section{Materials and Methods}\label{sec:methods}

### 2019-02-27 paper.tex

--- 2019-02-26 paper.tex
@@
 \section{Introduction}
+
+In the recent years, neuroimaging Big Data has become more accessible, as a
+result of the growing number of open-data initiatives\cite{openneuro, hcp, ukbiobank}. However, the
+processing of such large datasets using standard neuroimaging pipelines has 
+become unfeasable. As a result, many researchers are moving towards Big Data 
+engines, such as Apache Spark and Dask\cite{spark, dask}, to create their processing pipelines. 
+Unlike traditional neuroimaging engines, big data engines were designed to be 
+executed on dedicated commodity infrastructures using Big Data schedulers. Although
+researchers may have access to clouds or local workstations, they still largely 
+rely on High Performance Computing (HPC) infrastructure to process their data.
+
+High-Performance Computing infrastructure differs from the commodity infrastructure 
+typically used by Big Data platforms in that it is not dedicated. That is, there 
+are multiple users running a variety of different applications and all using the 
+same resources. Schedulers used on HPC infrastructure include PBS, SGE, Slurm, HTCondor and 
+TORQUE\cite{schedulers}. The majority of these being batch submission schedulers
+in which the number of resources are requested by the user and the specified 
+program is executed on the infrastructure.
+
+Another popular HPC scheduling approach,
+as implemented by HTCondor, is known as pilot-scheduling. Rather than requesting
+all necessary resources in a single batch call, a pilot-scheduling system will request
+multiple instances of subsets of the required resources. Processing starts occuring 
+as soon as a minimal amount of resources are allocated, and can be replenished if 
+resources are lost. Such a scheduling approach
+decreases the time a user spends in the queue waiting for resources as less resources
+are being requested per instance.\todo{need citation}. Many scientific workflow
+engines are compatible with pilot-scheduling schedulers \cite{nipype and others} and
+some others exclusively use pilot-scheduling approaches \cite{Pegasus and PSOM}.
+
+Apache Spark is a popular Big Data framework, commonly used in both industrial
+and academic settings. Although it is a Scala-based framework, it also has APIs
+for Java, Python (PySpark) and R. What made it powerful over its predecessors is their main
+abstraction known as the Resilient Distributed Dataset (RDD), which enabled
+in-memory processing of pipelines by scheduling the tasks to where the data is 
+located. Through the use of RDDs, it also became possible to execute iterative 
+workflows -- something not easily doable in older frameworks such as Hadoop MapReduce.
+Schedulers for Spark include its built-in standalone schedule, Yet Another Resource Negotiator (YARN),
+and Mesos \cite{yarn, mesos}. Out of the three schedulers, only Mesos can be used
+as an HPC scheduler. However, Mesos is not frequently used as a scheduler in HPC
+environments.
+
+Although still not widely adopted possibly due to the difficulties in adopting Big Data frameworks on HPC \todo{better explain the difficulties...setting walltimes etc.
+also explain neuroimaging applications},
+projects, such as Thunder, a PySpark-compatible library of tools from processing
+images and time-series data of various fields, including neuroscience, exist. In addition,
+Spark has been used in various research projects to process neuroimaging data \cite{Boubela, ariel's paper, maybe the simulation one}.
+
+The recommended method for launching Spark applications on our HPC scheduler 
+involves batch requesting all the necessary resources and launching a Standalone
+Spark cluster once the resources have been requested. This could significantly increase
+the scheduling time of Spark-based applications as a large amount of resources 
+may be necessary to process Big Data. Rather than requesting all the resources 
+at once, a Spark cluster is launched with a subset of the resources and is expanded as 
+more resources get allocated. This, in turn, may reduce the overall processing of
+an application as less resources are requested at once and are therefore more likely
+to be scheduled faster.
+
+There have been some efforts to use pilot-scheduling approaches with Spark applications
+running on HPC \cite{jha and spark on pbs paper}. However, these projects are either
+not open-source and not necessarily compatible with Slurm.
+
+Here we present SPA, a pilot-scheduling library for launching Spark applications 
+on HPC infrastracture. This libary currently enables the launching of Spark applications
+on Slurm-based clusters by dividing the total amount of resources required and requesting
+multiple instances of subsets of the resources at a time. When a single instance is started,
+a Standalone Spark cluster is started. Built-in fault-tolerance of Spark masters 
+and workers is leveraged, in addition to Driver fault tolerance for Scala-based applications.
+When nodes are lost due to walltime expiration, additional pilot nodes are launched until
+the driver completes. Using this system, it is more feasible to run large neuroimaging Spark applications on
+HPC systems.
+
 \section{Materials and Methods}\label{sec:methods}

### 2019-02-27 paper.tex

--- 2019-02-26 paper.tex
@@
 \begin{document}
-\title{SPA: An Apache Spark Pilot scheduler for HPC applications}
+\title{SPA: An Apache Spark Pilot scheduler\\ for HPC applications}
 \author{
@@
     batch-requesting of the resources which can greatly increase 
-    queueing times of the Big Data 
-    applications. Pilot-scheduling is a well known HPC scheduling
-    strategy that begins scheduling as compute resources are allocated to the 
-    user, thereby reducing HPC scheduling wait times. Here, we introduce SPA, an
+    queuing times of the Big Data applications. Pilot scheduling is a 
+    well known strategy that decouples resource provisioning from task 
+    scheduling, thereby allowing applications to leverage dynamic 
+    pools of computing resources. Here, we introduce SPA, an 
     open-source
@@
     such as robust workers, masters and driver, in addition to 
-    checkpointing, to ensure that processing occurs even when nodes are 
-    lost due to expired walltime. SPA also ensures that a 
+    checkpointing, to ensure robustness against nodes lost due to
+    expired walltime. SPA also ensures that a 
     user-specified number of pilots is maintained in the cluster in 
@@
     idle for a certain amount of time, they are terminated by 
-    the system. \tristan{Add a sentence on results when you have them.}
+    the system. We evaluate SPA on neuroimaging pipelines, our main domain of interest. \tristan{Add a sentence on results when you have them.}
 
@@
 
-In the recent years, neuroimaging Big Data has become more accessible, largely 
-attributed to the growing number of open-data initiatives. However, the
+In the recent years, neuroimaging Big Data has become more accessible, as a
+result of the growing number of open-data initiatives~\cite{openneuro, hcp, ukbiobank}. However, the
 processing of such large datasets using standard neuroimaging pipelines has 
-become unfeasable. As a result, many researchers are moving towards Big Data 
-engines, such as Apache Spark and Dask, to create their processing pipelines. 
-Unlike traditional neuroimaging engines, big data engines were designed to be 
-executed on commodity 
+become cumbersome. As a result, many researchers are moving towards Big Data 
+engines, such as Apache Spark~\cite{spark} and Dask~\cite{sdask}, to 
+create their processing pipelines. Unlike traditional neuroimaging 
+engines, Big Data engines were designed to be executed on dedicated 
+commodity infrastructures using Big Data schedulers. However, 
+researchers still largely rely on High Performance Computing (HPC) 
+infrastructure to process their data, although they may have access to 
+clouds or local workstations.
+
+High-Performance Computing infrastructure differs from the commodity infrastructure 
+typically used by Big Data platforms in that it is not dedicated. That is, there 
+are multiple users running a variety of different applications and all using the 
+same resources. Schedulers used on HPC infrastructure include PBS, SGE, Slurm, HTCondor and 
+TORQUE~\cite{schedulers}. The majority of these are batch submission schedulers
+in which the number of resources are requested by the user and the specified 
+program is executed on the infrastructure.
+
+Another popular HPC scheduling approach,
+implemented notably by HTCondor~\cite{htcondor}, is known as pilot scheduling. Rather than requesting
+all necessary resources in a single batch call, a pilot-scheduling system will request
+multiple instances of subsets of the required resources. Processing starts 
+as soon as a minimal amount of resources are allocated, and can be replenished if 
+resources are lost. Such a scheduling approach decreases the time a 
+user spends in the queue waiting for resources as less resources are 
+being requested per instance.\todo{need citation} In addition, pilot 
+scheduling makes it easier for users to estimate the number of 
+cores required by their application, as a slight overestimation would not impact queuing 
+time. Many scientific workflow engines are compatible with 
+pilot-scheduling schedulers~\cite{nipype and others} and some others 
+exclusively use pilot-scheduling approaches~\cite{Pegasus and PSOM}. 
+\tristan{I'm not sure if Pegasus is exclusively pilot based.}
+
+Apache Spark is a popular Big Data framework, commonly used in both industrial
+and academic settings. Although it is a Scala-based framework, it also has APIs
+for Java, Python (PySpark) and R. Spark's Resilient Distributed Dataset (RDD) abstraction enabled
+in-memory processing of pipelines by co-locating tasks and data, which provided important performance improvements compared to its predecessors.
+ Through the use of RDDs, it also became possible to execute iterative 
+workflows -- something not easily doable in older frameworks such as Hadoop MapReduce.
+Schedulers for Spark include its built-in standalone schedule, Yet Another Resource Negotiator (YARN),
+and Mesos~\cite{yarn, mesos}. Out of the three schedulers, only Mesos can be used
+as an HPC scheduler, which is not frequently done. \tristan{why?}
+
+Projects such as Thunder~\cite{thunder}, a PySpark-compatible library 
+of image processing tools, have been developed in various fields 
+although they are still not widely adopted, possibly due to the 
+difficulties in adapting Big Data frameworks on HPC \todo{better 
+explain the difficulties...setting walltimes etc. also explain 
+neuroimaging 
+applications}. In addition, Spark has been used in 
+various research projects to process neuroimaging data \cite{Boubela, 
+ariel's paper, maybe the simulation one}. \tristan{in this paragraph you could refer
+to your pre-print for more information.}
+
+The recommended method for launching Spark applications on HPC schedulers 
+involves batch requesting all the necessary resources and launching a standalone
+Spark cluster once the resources have been requested. This could significantly increase
+the scheduling time of Spark-based applications as a large amount of resources 
+may be necessary to process Big Data. In a pilot-scheduling model, rather than requesting all the resources 
+at once, a Spark cluster is launched with a subset of the resources and is expanded as 
+more resources get allocated. This, in turn, may reduce the overall processing of
+an application as less resources are requested at once and are therefore more likely
+to be scheduled faster.
+
+There have been some efforts to use pilot-scheduling approaches with Spark applications
+running on HPC \cite{jha and spark on pbs paper}. However, these projects are either
+not open-source or not necessarily compatible with Slurm.
+
+Here we present SPA, a pilot-scheduling library for launching Spark applications 
+on HPC infrastructures. This library currently enables the launching of Spark applications
+on Slurm-based clusters by dividing the total amount of resources required and requesting
+multiple instances of subsets of the resources at a time. When a single 
+instance is started, a Standalone Spark cluster is started. Built-in 
+fault-tolerance of Spark masters and workers is leveraged, in addition 
+to Driver fault tolerance for Scala-based applications. When nodes are 
+lost due to walltime expiration, additional pilot nodes are launched 
+until the driver completes. SPA facilitates the execution of large 
+neuroimaging Spark applications on HPC systems.
+
 \section{Materials and Methods}\label{sec:methods}

### 2019-04-26 experiments/code/raw_to_json.py
new data with 4 dedicated nodes added

@@ -55,9 +55,17 @@ def get_jobs(fn, sj_benchmark_dir, s_logs, exec_mode):
                 sj_elem['id'] = sj
                 sj_elem['start_time'] = None
                 sj_elem['end_time'] = None
-                nodes, success = get_jobid_success(sj, s_logs)
+                nodes, success, worker_count = get_jobid_success(sj, s_logs)
                 sj_elem['nodes'] = list(nodes)
-                sj_elem['succeeded'] = success if success or driver_id is None else get_success(driver_path)
+
+                if success or driver_id is None:
+                    sj_elem['worker_count'] = worker_count
+                    sj_elem['succeeded'] = success
+                else:
+                    success, worker_count = get_success(driver_path)
+                    sj_elem['worker_count'] = worker_count
+                    sj_elem['succeeded'] = success
+
                 sjelems.append(sj_elem)
 
                 bench_dir = glob.glob(op.join(sj_benchmark_dir, '*benchmarks.{}.out'.format(sj)))
@@ -78,12 +86,16 @@ def get_jobs(fn, sj_benchmark_dir, s_logs, exec_mode):
 
 
 def get_success(fp):
+    success = False
+    worker_count = 0
     if op.isdir(fp):
         with open(op.join(fp, 'stderr'), 'r') as f:
             for line in f:
-                if 'Finished task' in line and '125/125' in line:
-                    return True
-    return False
+                if 'Executor added' in line:
+                    worker_count += 1
+                elif 'Finished task' in line and '125/125' in line:
+                    success = True
+    return success, worker_count
 
 
 def order_pilots(directory, sjids, exec_mode="batch"):
@@ -119,8 +131,10 @@ def order_pilots(directory, sjids, exec_mode="batch"):
                 dedicated = 'single'
             elif dedicated == 2:
                 dedicated = 'double'
-            else:
+            elif dedicated == 3:
                 dedicated = 'triple'
+            else:
+                dedicated = 'quadruple'
         
         abs_fn = op.abspath(op.join(directory, fn))
 
@@ -153,7 +167,12 @@ def order_pilots(directory, sjids, exec_mode="batch"):
 
     print(len(total_order))
     for idx, elem in enumerate(total_order):
+        elem['worker_count'] = sjids[idx][0]['worker_count'] if len(sjids) > 0 else 0
         elem['sid'] = sjids[idx]
+        
+        for sj in sjids[idx]:
+            sj.pop('worker_count')
+
[…]

### 2019-04-26 experiments/code/raw_to_json.py
added info to json

@@ -55,16 +55,15 @@ def get_jobs(fn, sj_benchmark_dir, s_logs, exec_mode):
                 sj_elem['id'] = sj
                 sj_elem['start_time'] = None
                 sj_elem['end_time'] = None
-                nodes, success, worker_count = get_jobid_success(sj, s_logs)
-                sj_elem['nodes'] = list(nodes)
+                nodes, success = get_jobid_success(sj, s_logs)
+                sj_elem['nodes'] = nodes
 
                 if success or driver_id is None:
-                    sj_elem['worker_count'] = worker_count
                     sj_elem['succeeded'] = success
                 else:
-                    success, worker_count = get_success(driver_path)
-                    sj_elem['worker_count'] = worker_count
+                    success, workers = get_success(driver_path, list(nodes.keys())[0] if len(nodes) > 0 else None)
                     sj_elem['succeeded'] = success
+                    sj_elem['nodes'] = workers
 
                 sjelems.append(sj_elem)
 
@@ -85,17 +84,18 @@ def get_jobs(fn, sj_benchmark_dir, s_logs, exec_mode):
     return job_ids
 
 
-def get_success(fp):
+def get_success(fp, node):
     success = False
-    worker_count = 0
-    if op.isdir(fp):
+    workers = set()
+    if op.isdir(fp) and node is not None:
         with open(op.join(fp, 'stderr'), 'r') as f:
             for line in f:
-                if 'Executor added' in line:
-                    worker_count += 1
+                if 'Executor added' in line and node in line:
+                    proc = line.split(' ')[-4].strip('(').strip(')').split(':')[1]
+                    workers.add(proc)
                 elif 'Finished task' in line and '125/125' in line:
                     success = True
-    return success, worker_count
+    return success, { node: list(workers) }
 
 
 def order_pilots(directory, sjids, exec_mode="batch"):
@@ -167,11 +167,9 @@ def order_pilots(directory, sjids, exec_mode="batch"):
 
     print(len(total_order))
     for idx, elem in enumerate(total_order):
-        elem['worker_count'] = sjids[idx][0]['worker_count'] if len(sjids) > 0 else 0
+        elem['worker_count'] = sum([len(el[1]) for sj in sjids[idx] for el in sj['nodes'].items()])
         elem['sid'] = sjids[idx]
         
-        for sj in sjids[idx]:
-            sj.pop('worker_count')
 
[…]

### 2019-04-26 experiments/code/raw_to_json.py
minor fixes

@@ -167,7 +167,15 @@ def order_pilots(directory, sjids, exec_mode="batch"):
 
     print(len(total_order))
     for idx, elem in enumerate(total_order):
-        elem['worker_count'] = sum([len(el[1]) for sj in sjids[idx] for el in sj['nodes'].items()])
+        
+        node_workers = {}
+        
+        for sj in sjids[idx]:
+            for k,v in sj['nodes'].items():
+                if k not in node_workers:
+                    node_workers[k] = v
+
+        elem['worker_count'] = sum([len(el[1]) for el in node_workers.items()])
         elem['sid'] = sjids[idx]
         
 

### 2019-04-27 experiments/code/raw_to_json.py
fixes to worker count

@@ -61,9 +61,8 @@ def get_jobs(fn, sj_benchmark_dir, s_logs, exec_mode):
                 if success or driver_id is None:
                     sj_elem['succeeded'] = success
                 else:
-                    success, workers = get_success(driver_path, list(nodes.keys())[0] if len(nodes) > 0 else None)
+                    success = get_success(driver_path, list(nodes.keys())[0] if len(nodes) > 0 else None)
                     sj_elem['succeeded'] = success
-                    sj_elem['nodes'] = workers
 
                 sjelems.append(sj_elem)
 
@@ -86,16 +85,12 @@ def get_jobs(fn, sj_benchmark_dir, s_logs, exec_mode):
 
 def get_success(fp, node):
     success = False
-    workers = set()
     if op.isdir(fp) and node is not None:
         with open(op.join(fp, 'stderr'), 'r') as f:
             for line in f:
-                if 'Executor added' in line and node in line:
-                    proc = line.split(' ')[-4].strip('(').strip(')').split(':')[1]
-                    workers.add(proc)
-                elif 'Finished task' in line and '125/125' in line:
+                if 'Finished task' in line and '125/125' in line:
                     success = True
-    return success, { node: list(workers) }
+    return success
 
 
 def order_pilots(directory, sjids, exec_mode="batch"):
@@ -169,13 +164,8 @@ def order_pilots(directory, sjids, exec_mode="batch"):
     for idx, elem in enumerate(total_order):
         
         node_workers = {}
-        
-        for sj in sjids[idx]:
-            for k,v in sj['nodes'].items():
-                if k not in node_workers:
-                    node_workers[k] = v
 
-        elem['worker_count'] = sum([len(el[1]) for el in node_workers.items()])
+        elem['worker_count'] = sum([len(el[1]) for sj in sjids[idx] for el in sj['nodes'].items()])
         elem['sid'] = sjids[idx]
         
 
@@ -198,10 +188,26 @@ def get_jobid_success(job_id, master_logs):
 
         with open(logfile[0], 'r') as f:
             for line in f:
-                if not batch and 'NODE: ' in line:
-                    executors[line.split(' ')[-1].strip('\n')] = []
-                elif '"finishedexecutors"' in line:
-                    return executors, False
+                if not batch and 'starting org.apache.spark.deploy.worker.Worker, logging to ' in line:
+                    wlogs = line.split('starting org.apache.spark.deploy.worker.Worker, logging to ')
+
+                    for wl in wlogs:
[…]

### 2019-04-27 experiments/notebooks/generate_figures.py
updated notebook

@@ -30,6 +30,7 @@ def all_succeeded(i):
 dedicated_1 = makespan_dict()
 dedicated_2 = makespan_dict()
 dedicated_3 = makespan_dict()
+dedicated_4 = makespan_dict()
 
 get_makespan = lambda x: x['end_time'] - x['start_time'] if x['success'] else 0
 is_same = lambda x, i: x in pilots8[i]['name'] and x in pilots16[i]['name']
@@ -47,6 +48,9 @@ def fill_dictionaries():
         elif 'triple' in batch[i]['name']:
             assert(is_same('3d', i))
             current_dict = dedicated_3
+        elif 'quadruple' in batch[i]['name']:
+            assert(is_same('4d', i))
+            current_dict = dedicated_4
 
         current_dict['batch'].append(get_makespan(batch[i]))
         current_dict['8pilots'].append(get_makespan(pilots8[i]))
@@ -93,11 +97,12 @@ def get_makespan_diff(d, num_pilots):
     return data
 
 
-def makespan_box(d1, d2, d3, num_pilots, system="Beluga", save=None):
+def makespan_box(d1, d2, d3, d4, num_pilots, system="Beluga", save=None):
     fig, ax = plt.subplots()
-    data = [get_makespan_diff(dedicated_1, num_pilots),
-            get_makespan_diff(dedicated_2, num_pilots),
-            get_makespan_diff(dedicated_3, num_pilots)]
+    data = [get_makespan_diff(d1, num_pilots),
+            get_makespan_diff(d2, num_pilots),
+            get_makespan_diff(d3, num_pilots),
+            get_makespan_diff(d4, num_pilots)]
 
     pos = np.array(range(len(data))) + 1
     bp = ax.boxplot(data, sym='k+', positions=pos)
@@ -153,14 +158,31 @@ def scatter_fig(x, y, num_pilots, xlabel, ylabel, system="Beluga", save=None):
         plt.savefig(save)
 
 
-def add_kv(d, k, v):
+def add_kv(d, k, v, w=True):
+
+    if len(v) > 0 and type(v[0]) == dict:
+        v = get_ld_keys(v)
     if k not in d:
         d[k] = set(v)
     else:
         d[k].update(v)
 
 
-def get_pilot_info(job, t):
+def get_ld_keys(l, w=False):
+    temp = []
+    if w:
+        for el in l:
+            temp.extend(["{0}:{1}".format(k, p) for k,v in el.items()
[…]

### 2019-04-27 experiments/code/launch_bench.py
added interexperiment randomization

@@ -96,13 +96,17 @@ while count < iterations :
         f_batch = NamedTemporaryFile(mode='w+', prefix='batch_', suffix='_{}'.format(exps['cond']))
         f_8pilot = NamedTemporaryFile(mode='w+', prefix='pilot8_', suffix='_{}'.format(exps['cond']))
         f_16pilot = NamedTemporaryFile(mode='w+', prefix='pilot16', suffix='_{}'.format(exps['cond']))
-        x = Popen(exps['batch'], bufsize=1, stdout=f_batch, stderr=f_batch)
-        y = Popen(exps['8pilot'], bufsize=1, stdout=f_8pilot, stderr=f_8pilot)
-        z = Popen(exps['16pilot'], bufsize=1, stdout=f_16pilot, stderr=f_16pilot)
+        options = [('batch', f_batch, batch_out),
+                   ('8pilot', f_8pilot, pilot8_out),
+                   ('16pilot', f_16pilot, pilot16_out)]
+        shuffle(options)
+        x = Popen(exps[options[0][0]], bufsize=1, stdout=options[0][1], stderr=options[0][1])
+        y = Popen(exps[options[1][0]], bufsize=1, stdout=options[1][1], stderr=options[1][1])
+        z = Popen(exps[options[2][0]], bufsize=1, stdout=options[2][1], stderr=options[2][1])
 
-        get_results(x, f_batch, "batch", exps['cond'], batch_out)
-        get_results(y, f_8pilot, "8pilot", exps['cond'], pilot8_out)
-        get_results(z, f_16pilot, "16pilot", exps['cond'], pilot16_out)
+        get_results(x, options[0][1], options[0][0], exps['cond'], options[0][2])
+        get_results(y, options[1][1], options[1][0], exps['cond'], options[1][2])
+        get_results(z, options[2][1], options[2][0], exps['cond'], options[2][2])
 
         try:
             shutil.rmtree(batch_out, ignore_errors=True)

### 2019-04-29 paper.tex
modified abstract

@@
 \begin{abstract}
-    Big Data processing engines such as Apache Spark are increasingly 
-    used to process scientific data. These processing engines, however, 
-    have been primarily designed to work on dedicated infrastructure 
-    and are not well designed to work on High-Performance Computing 
-    (HPC) systems commonly available to researchers. Typical mechanisms 
-    for launching Spark-based applications on HPC clusters involve 
-    batch-requesting of the resources which can greatly increase 
-    queuing times of the Big Data applications. Pilot scheduling is a 
-    well known strategy that decouples resource provisioning from task 
-    scheduling, thereby allowing applications to leverage dynamic 
-    pools of computing resources. Here, we introduce SPA, an 
-    open-source
-    pilot-job submission system designed for launching Apache Spark pipelines 
-    on Slurm-based clusters. SPA leverages optimizations available in Spark, 
-    such as robust workers, masters and driver, in addition to 
-    checkpointing, to ensure robustness against nodes lost due to
-    expired walltime. SPA also ensures that a 
-    user-specified number of pilots is maintained in the cluster in 
-    order to ensure a constant level of parallelism. Should pilots be 
-    idle for a certain amount of time, they are terminated by 
-    the system. We evaluate SPA on neuroimaging pipelines, our main domain of interest. \tristan{Add a sentence on results when you have them.}
+    Big Data has become prominent throughout many scientific fields and, as a
+    result, scientific communities have sought out Big Data frameworks to 
+    accelerate the processing of their increasingly data-intensive pipelines.
+    However, while scientific communities typically rely on High-Performance 
+    Computing (HPC) clusters for the parallelization of their pipelines, many 
+    popular Big Data frameworks such as Hadoop and Spark were primarily designed
+    to be executed on dedicated commodity infrastructure. Differences found between
+    these infrastructures and the policies that surround them limit the advantages
+    of deploying Big Data pipelines on HPC infrastructure. For instance, HPC
+    clusters typically employ a batch job submission system requiring users to
+    specify details on number of cores, nodes, amount of memory and duration.
+    Big Data engines, such as Spark, have no built-in mechanism to interact with
+    HPC batch job submission systems. Morever, the details required by the HPC
+    schedulers are not necessarily initially known to users of Big Data frameworks.
 
+    Pilot scheduling strategies have been developed to address the limitations 
+    of traditional HPC batch job schedulers. Pilot schedulers decouple resource
+    provisioning from task scheduling, thereby enabling efficient resource
+    utilization through dynamic scheduling. This paper evaluates the benefits 
+    pilot-scheduling strategies over traditional batch submission
+    on HPC clusters with overlay Apache Spark clusters. We first evaluate the
+    overall speedup brought on by employing pilot-scheduling strategies. We then
+    examine the robustness to master and driver failures, which may be frequent 
+    when underestimating pilot walltime. Spark built-in checkpointing is also 
+    investigated in relation to pilot expiration.
+
+    Our results show that\ldots TBD.
 \end{abstract}
@@
 \section{Materials and Methods}\label{sec:methods}
+    SPA consists of a Python-based driver with bash-templates that are called 
+    using the Slurmpy~\cite{slurmpy} library. In order to commence pilot-scheduling,
+    the user must supply the bash template of choice and a JSON configuration 
+    file which contains details on pilot node requirements and the desired application
+    to be launched. Currently, there only exists a single pilot template for launching
+    Spark jobs on a Slurm-based HPC cluster. The pseudocode for that template can 
+    be seen in ~\ref{Figure:pilot-temp}.
     \todo{walltime parameter of workflow}

### 2019-04-29 paper.tex
Comments on abstract and intro

@@
       #1 \color{black}}
+\newcommand{\TG}[1]{\tristan{#1}}
 
@@
     accelerate the processing of their increasingly data-intensive pipelines.
+    \TG{I'd write the previous sentence in continuous present tense, since I think
+    this is an on-going process.}
     However, while scientific communities typically rely on High-Performance 
@@
     popular Big Data frameworks such as Hadoop and Spark were primarily designed
-    to be executed on dedicated commodity infrastructure. Differences found between
+    to be executed on dedicated commodity infrastructures. Differences found between
     these infrastructures and the policies that surround them limit the advantages
@@
     HPC batch job submission systems. Morever, the details required by the HPC
-    schedulers are not necessarily initially known to users of Big Data frameworks.
+    schedulers \TG{be more specific than ``details'': do you mean ``scheduling policies'',
+    ``application resource requirerments?''
+    } are not necessarily initially known to users of Big Data frameworks.
+    \TG{The previous paragraph is a bit long: it could be summarized to 2-3
+    sentences.}
 
     Pilot scheduling strategies have been developed to address the limitations 
-    of traditional HPC batch job schedulers. Pilot schedulers decouple resource
+    of traditional HPC batch job schedulers \TG{You could give examples of pilot schedulers as you did with Spark and Hadoop}.
+    Pilot schedulers decouple resource
     provisioning from task scheduling, thereby enabling efficient resource
@@
 
-In the recent years, neuroimaging Big Data has become more accessible, as a
+In the recent years, neuroimaging \TG{If you want to put neuroimaging first, then you need to explain it. Just say that it's an example among many fields}
+Big Data has become more accessible, as a
 result of the growing number of open-data initiatives~\cite{openneuro, hcp, ukbiobank}. However, the
@@
 involves batch requesting all the necessary resources and launching a standalone
-Spark cluster once the resources have been requested. This could significantly increase
+Spark cluster once the resources have been requested \TG{Add link to Compute Canada's instructions?}. This could significantly increase
 the scheduling time of Spark-based applications as a large amount of resources 
@@
 neuroimaging Spark applications on HPC systems.
+
+\TG{The introduction is still very much geared toward the presentation of 
+the SPA software. Instead, I think it should introduce the question whether pilot jobs
+are useful at all to overlay schedulers, explain why they may be, and why they may not (there is already an overlay scheduler,
+and resource heterogeneity is limited). You could explain that the traditional benefits of pilot-job systems are:
+\begin{itemize}
+    \item Address worker node heterogeneity
+    \item Address worker node failures (software, incl walltime expirations, or hardware)
+    \item Address variable queuing times
+    \item Tolerant to resource requirement overestimation, in particular walltime and CPU
+    \item Increase workload distribution (use more nodes)
+\end{itemize}
+Among these, (1) and to some extent (2) are already addressed by an overlay scheduler.
+
+The title should also reflect this.
+}
 

### 2019-04-29 example/standalone/hpc/hpc_pilot_template.sh
modified templates for benchmarking individual tasks

@@ -80,6 +80,9 @@ pilot_program(){
         		MASTER_URI=${MASTER_URI/%????/$REST_SERVER_PORT} # cluster mode requires REST port
 			echo 'Cluster deploy master: ' $MASTER_URI
 
+			app_log_dir=${SLURM_JOBID}-applogs
+			mkdir $app_log_dir
+
 			eval $driver_prog > output 2>&1
 			cat output
 			driverid=`cat output | grep submissionId | grep -Po 'driver-\d+-\d+'`

### 2019-05-01 paper.tex
some fixes to intro

@@
 \begin{document}
-\title{SPA: An Apache Spark Pilot scheduler\\ for HPC applications}
+\title{Evaluation of dynamic resource allocation strategies for overlay Apache
+       Spark clusters on HPC}
 \author{
@@
 
-In the recent years, neuroimaging \TG{If you want to put neuroimaging first, then you need to explain it. Just say that it's an example among many fields}
-Big Data has become more accessible, as a
-result of the growing number of open-data initiatives~\cite{openneuro, hcp, ukbiobank}. However, the
-processing of such large datasets using standard neuroimaging pipelines has 
-become cumbersome. As a result, many researchers are moving towards Big Data 
-engines, such as Apache Spark~\cite{spark} and Dask~\cite{sdask}, to 
-create their processing pipelines. Unlike traditional neuroimaging 
-engines, Big Data engines were designed to be executed on dedicated 
-commodity infrastructures using Big Data schedulers. However, 
-researchers still largely rely on High Performance Computing (HPC) 
-infrastructure to process their data, although they may have access to 
-clouds or local workstations.
+With the increasing volume of data being collected and shared, Big Data has become
+an important area of focus to many scientific domains. As is the case with neuroimaging,
+dataset available publicly now reach up to Petabytes in size~\cite{UKBiobank, hcp}. Average workstations found in research lab would not be
+able to support the processing of such large dataset, leaving researchers to look 
+into cloud or High-Performance Computing~(HPC) solutions. Whereas both options are
+suitable for the processing of scientific Big Data, HPC clusters remain the more 
+sought after solution, potentially due to its cost-effectiveness.
 
-High-Performance Computing infrastructure differs from the commodity infrastructure 
-typically used by Big Data platforms in that it is not dedicated. That is, there 
-are multiple users running a variety of different applications and all using the 
-same resources. Schedulers used on HPC infrastructure include PBS, SGE, Slurm, HTCondor and 
-TORQUE~\cite{schedulers}. The majority of these are batch submission schedulers
-in which the number of resources are requested by the user and the specified 
-program is executed on the infrastructure.
+Scientific workflows may be quite compute intensive, requiring more than a few
+hours to complete (\cite{freesurfer, fmriprep, a paper which mentions duration}).
+The coupling of large datasets with compute-intensive workflows may result in 
+unreasonably long processing times. Such limitations restrict the amount data 
+utilized in scientific experiments. To reduce the effects of Big Data on processing
+times, researchers have begun to move towards employing Big Data frameworks for their
+analyses. Examples of Big Data frameworks including Hadoop MapReduce~\cite{Hadoop},
+Apache Spark~\cite{Spark} and Dask~\cite{Dask}. However, these frameworks were 
+designed with dedicated commodity infrastructure in mind, and, with the exception
+of Dask, do not support traditional HPC schedulers.
 
-Another popular HPC scheduling approach,
-implemented notably by HTCondor~\cite{htcondor}, is known as pilot scheduling. Rather than requesting
-all necessary resources in a single batch call, a pilot-scheduling system will request
-multiple instances of subsets of the required resources. Processing starts 
-as soon as a minimal amount of resources are allocated, and can be replenished if 
-resources are lost. Such a scheduling approach decreases the time a 
-user spends in the queue waiting for resources as less resources are 
-being requested per instance.\todo{need citation} In addition, pilot 
-scheduling makes it easier for users to estimate the number of 
-cores required by their application, as a slight overestimation would not impact queuing 
-time. Many scientific workflow engines are compatible with 
-pilot-scheduling schedulers~\cite{nipype and others} and some others 
-exclusively use pilot-scheduling approaches~\cite{Pegasus and PSOM}. 
-\tristan{I'm not sure if Pegasus is exclusively pilot based.}
+Within High-Performance Computing clusters, many users processing a variety of different
+applications, share resources. These shared resources include storage (e.g. Parallel file system) and 
+compute node resources (memory, cpu, local disk space, etc.). In order for users to 
+gain access to shared compute resources, HPC clusters typically employ schedulers for 
+batch resource provisioning. Examples of such schedulers include PBS, SGE, Slurm and 
+TORQUE~\cite{schedulers}. These schedulers require that the user specify the amount
+of resources required and the duration that they require the resources for. The users 
+are then placed in a queue at a position based on the user's priority and resource 
+availability.
+
+There are several limitations to running applications on batch HPC schedulers, 
+particularly in the face of Big Data. For instance, users may not necessarily 
+know the exact requirements of the applications. With applications that are 
+both data- and compute-intensive, this may result in the user overestimating the
+resources required and the duration that these resources are required for. Frequent
+high overestimates may create bottlenecks in the system, preventing other users
+from occupying allocated idle resources. In contrast, users may underestimate required 
+resources. This may subsequently result in many failures within the application leading to
+additional re-executions of the same application. This is problematic for the user
+as their priority may be affected by the cluster usage. Additionally, the user may 
+require more time on the cluster to relaunch non-checkpointed and interrupted 
+tasks. Furthermore, Big Data frameworks, such as Apache Spark, do not interact naturally
+with HPC schedulers and typically rely on other schedulers for resource provisioning
+and task scheduling. Therefore, to run Big Data applications on HPC schedulers, 
+it is also necessary to start an overlay cluster.
+
+In contrast to static resource provisioning, as is the case with batch HPC schedulers, 
+dynamic resource provisioning, known as pilot job or glide-in scheduling, exists. 
+This dynamic resource provisioning allows the user to specify a minimum and maximum 
+amount of resources, and the cluster can expand and contract depending on application requirements.
+Each pilot may consist of the minimal amount of resources required to launch the 
+application. As each pilot would request allocation of fewer resources than that of a batch
+request, Pilot Jobs would typically spend less time on the resource allocation 
+queue. Furthermore, although individual pilots may expire due to walltimes, the dynamic
+nature of pilots allows the entire application to surpass walltime limitations, as
+additional pilots may be added to the pilot queue until application completion. 
+Examples of HPC schedulers employing pilot strategies include HTCondor~\cite{htcondor}
+and DIRAC~\cite{DIRAC}. \todo{maybe mention PSOM and Pegasus here}
 
@@
 Schedulers for Spark include its built-in standalone schedule, Yet Another Resource Negotiator (YARN),
-and Mesos~\cite{yarn, mesos}. Out of the three schedulers, only Mesos can be used
-as an HPC scheduler, which is not frequently done. \tristan{why?}
+and Mesos~\cite{yarn, mesos}.
 
@@
 
-There have been some efforts to use pilot-scheduling approaches with Spark applications
-running on HPC \cite{jha and spark on pbs paper}. However, these projects are either
+Pilot scheduling has been applied to scientific workflow engines including Pegasus,
+PSOM and RADICAL-Pilot. While there have been some efforts on implementing pilot scheduling
+approaches for \cite{jha and spark on pbs paper}. However, these projects are either
 not open-source or not necessarily compatible with Slurm.

### 2019-05-02 paper.tex
updated cedar figures, bug fixes to figure generation script and changes to intro

@@
 
-Pilot scheduling has been applied to scientific workflow engines including Pegasus,
-PSOM and RADICAL-Pilot. While there have been some efforts on implementing pilot scheduling
-approaches for \cite{jha and spark on pbs paper}. However, these projects are either
-not open-source or not necessarily compatible with Slurm.
+Pilot scheduling techniques have already been applied to scientific workflow engines, such as
+Pegasus, PSOM and RADICAL-Pilot~\cite{}. While there have been some efforts on
+implementing pilot scheduling approaches for Apache Spark\cite{jha and spark on pbs paper},
+research is limited and none of the them detail the effects of an overlay cluster
+on overall performance and robustness.
 
+In this paper, we will analyse the effects of pilot scheduling strategies with an 
+overlay Spark cluster. Our evaluation will be broken down into two parts: 1)
+the overall added value of pilots, and 2), tolerance to failures. Part 1 will 
+investigate the effect of queuing times on pilot applications. 
 Here we present SPA, a pilot-scheduling library for launching Spark applications 

### 2019-05-02 experiments/notebooks/generate_figures.py

@@ -32,7 +32,9 @@ dedicated_2 = makespan_dict()
 dedicated_3 = makespan_dict()
 dedicated_4 = makespan_dict()
 
-get_makespan = lambda x: x['end_time'] - x['start_time'] if x['success'] else 0
+get_makespan = lambda x: (x['end_time'] - x['start_time']
+                          if x['success'] and x['end_time'] is not None
+                          else 0)
 is_same = lambda x, i: x in pilots8[i]['name'] and x in pilots16[i]['name']
 
 
@@ -121,16 +123,19 @@ def makespan_box(d1, d2, d3, d4, num_pilots, system="Beluga", save=None):
 
 def queuing_time(bd, pd):
     return [bd[i]['sid'][0]['start_time'] -
-            min([p['start_time'] for p in pd[i]['sid']])
+            min([p['start_time'] for p in pd[i]['sid']
+                 if p['start_time'] is not None])
             for i in range(len(bd))
-            if bd[i]['success'] and pd[i]['success']]
+            if bd[i]['success'] and bd[i]['end_time'] is not None
+            and pd[i]['success'] and pd[i]['end_time'] is not None]
 
 
 def makespan_all(bd, pd):
     return [(bd[i]['end_time'] - bd[i]['start_time']) - 
             (pd[i]['end_time'] - pd[i]['start_time'])
             for i in range(len(bd))
-            if bd[i]['success'] and pd[i]['success']]
+            if bd[i]['success'] and bd[i]['end_time'] is not None
+            and pd[i]['success'] and pd[i]['end_time'] is not None]
 
 
 def makespan_queue(num_pilot):
@@ -162,10 +167,12 @@ def add_kv(d, k, v, w=True):
 
     if len(v) > 0 and type(v[0]) == dict:
         v = get_ld_keys(v)
-    if k not in d:
-        d[k] = set(v)
-    else:
-        d[k].update(v)
+
+    if k is not None:
+        if k not in d:
+            d[k] = set(v)
+        else:
+            d[k].update(v)
 
 
 def get_ld_keys(l, w=False):
@@ -207,7 +214,8 @@ def get_pilot_info(job, t, w=False):
     # Second passage
     for sid in job['sid']:
         for k,v in makespan_dict.items():
-            if k > sid['start_time'] and k < sid['end_time']:
+            if (sid['end_time'] is not None and sid['start_time'] is not None
+                and k > sid['start_time'] and k < sid['end_time']):
[…]

### 2019-05-02 paper.tex

@@
 overlay Spark cluster. Our evaluation will be broken down into two parts: 1)
-the overall added value of pilots, and 2), tolerance to failures. Part 1 will 
-investigate the effect of queuing times on pilot applications. 
-Here we present SPA, a pilot-scheduling library for launching Spark applications 
-on HPC infrastructures. This library currently enables the launching of Spark applications
-on Slurm-based clusters by dividing the total amount of resources required and requesting
-multiple instances of subsets of the resources at a time. When a single 
-instance is started, a Standalone Spark cluster is started. Built-in 
-fault-tolerance of Spark masters and workers is leveraged, in addition 
-to Driver fault tolerance for Scala-based applications. When nodes are 
-lost due to walltime expiration, additional pilot nodes are launched 
-until the driver completes. SPA facilitates the execution of large 
-neuroimaging Spark applications on HPC systems.
+the added value of pilots, and 2), tolerance to failures. For determining the 
+added value of pilots, we will investigate the effect of queuing times on 
+pilot applications as compared to a batch requests. As much of the application's
+tolerance to failures are features brought on by the overlay Spark cluster, in
+Part 2 we will introduce failures that would be typical to pilot scheduling 
+approaches (e.g. node failures due to walltime expiration). This part will be 
+broken down into master failures, driver failures and checkpointing. As pilots 
+may need to be replenished during execution, we will also consider the effects of 
+job queues \todo{reword}.
 

### 2019-05-03 experiments/notebooks/generate_figures.py
fixed to figure generation script and updates to notebook

@@ -280,8 +280,9 @@ get_batch_avgpilots = lambda x:(((len(set(x['sid'][0]['nodes']))
                                  / (x['end_time'] - x['start_time'])),
                                 x['end_time'] - x['start_time'])
 
-get_batch_avgw = lambda x:(((len(set(['{0}:{1}'.format(k, v)
-                                      for k,v in x['sid'][0]['nodes'].items()]))
+get_batch_avgw = lambda x:(((len(set(['{0}:{1}'.format(k, p)
+                                      for k,v in x['sid'][0]['nodes'].items()
+                                      for p in v]))
                                  * (x['sid'][0]['end_time']
                                     - x['sid'][0]['start_time']))
                                 / (x['end_time'] - x['start_time'])),

### 2019-05-03 paper.tex

@@
 \usepackage{colortbl} % for \rowcolor
+\usepackage{ulem} % for \uwave
 \usepackage{caption}
@@
 \title{Evaluation of dynamic resource allocation strategies for overlay Apache
-       Spark clusters on HPC}
+       Spark clusters on HPC \tristan{``Overlay'' and ``Apache Spark'' are redundant}}
 \author{
@@
 an important area of focus to many scientific domains. As is the case with neuroimaging,
-dataset available publicly now reach up to Petabytes in size~\cite{UKBiobank, hcp}. Average workstations found in research lab would not be
-able to support the processing of such large dataset, leaving researchers to look 
+datasets available publicly now reach up to Petabytes in size~\cite{UKBiobank, hcp}. Average workstations found in research labs would not be
+able to support the processing of such large datasets, leaving researchers to look 
 into cloud or High-Performance Computing~(HPC) solutions. Whereas both options are
 suitable for the processing of scientific Big Data, HPC clusters remain the more 
-sought after solution, potentially due to its cost-effectiveness.
+sought after solution, potentially due to their cost-effectiveness.
 
@@
 The coupling of large datasets with compute-intensive workflows may result in 
-unreasonably long processing times. Such limitations restrict the amount data 
+unreasonably long processing times, which restricts the amount of data 
 utilized in scientific experiments. To reduce the effects of Big Data on processing
 times, researchers have begun to move towards employing Big Data frameworks for their
-analyses. Examples of Big Data frameworks including Hadoop MapReduce~\cite{Hadoop},
+analyses, such as Hadoop MapReduce~\cite{Hadoop},
 Apache Spark~\cite{Spark} and Dask~\cite{Dask}. However, these frameworks were 
@@
 
-Within High-Performance Computing clusters, many users processing a variety of different
-applications, share resources. These shared resources include storage (e.g. Parallel file system) and 
-compute node resources (memory, cpu, local disk space, etc.). In order for users to 
+Within HPC clusters, many users share resources to process a variety of different
+applications. These shared resources include storage (e.g. parallel file system) and 
+compute node resources (memory, CPU, local disk space, etc.). In order for users to 
 gain access to shared compute resources, HPC clusters typically employ schedulers for 
@@
 from occupying allocated idle resources. In contrast, users may underestimate required 
-resources. This may subsequently result in many failures within the application leading to
+resources, leading the batch scheduler to kill applications due to exceeded resource quotas.
+This may subsequently result in many failures within the application leading to
 additional re-executions of the same application. This is problematic for the user
@@
 require more time on the cluster to relaunch non-checkpointed and interrupted 
-tasks. Furthermore, Big Data frameworks, such as Apache Spark, do not interact naturally
+tasks. \uwave{Furthermore, Big Data frameworks, such as Apache Spark, do not interact naturally
 with HPC schedulers and typically rely on other schedulers for resource provisioning
 and task scheduling. Therefore, to run Big Data applications on HPC schedulers, 
-it is also necessary to start an overlay cluster.
+it is also necessary to start an overlay cluster.} \tristan{The previous 2 sentences are a bit unclear. You should explain
+what an overlay cluster means here. ``Interact with other schedulers'' is vague. This point should end up explaining
+what static resource provisioning means for Spark, i.e., the Compute Canada approach. It might even be a separate paragraph.
+You could merge here the paragraph that currently explains that later.}
 
@@
 dynamic resource provisioning, known as pilot job or glide-in scheduling, exists. 
-This dynamic resource provisioning allows the user to specify a minimum and maximum 
+Dynamic resource provisioning allows the user to specify a minimum and maximum 
 amount of resources, and the cluster can expand and contract depending on application requirements.
+\tristan{Add one sentence to define what's a pilot}. 
 Each pilot may consist of the minimal amount of resources required to launch the 
@@
 Examples of HPC schedulers employing pilot strategies include HTCondor~\cite{htcondor}
-and DIRAC~\cite{DIRAC}. \todo{maybe mention PSOM and Pegasus here}
+and DIRAC~\cite{DIRAC}. \todo{maybe mention PSOM and Pegasus here \tristan{Yes for PSOM as it's neuroimaging. Pegasus
+only uses pilots through condor afaik.}}
 
-Apache Spark is a popular Big Data framework, commonly used in both industrial
+\tristan{I think this could be the 3rd paragraph.} Apache Spark is a popular Big Data framework, commonly used in both industrial
 and academic settings. Although it is a Scala-based framework, it also has APIs
@@
 Schedulers for Spark include its built-in standalone schedule, Yet Another Resource Negotiator (YARN),
-and Mesos~\cite{yarn, mesos}.
+and Mesos~\cite{yarn, mesos}. \tristan{You could cite Ariel's and your paper to say that Spark is sometimes useful in 
+neuroimaging on a cluster.}
 
@@
 ariel's paper, maybe the simulation one}. \tristan{in this paragraph you could refer
-to your pre-print for more information.}
+to your pre-print for more information.} \tristan{I would summarize that in 1 or 2 sentences at the end of the previous
+paragraph.}
 
@@
 on overall performance and robustness. 
+\tristan{This paragraph should write the scientific question of your paper more clearly: is it useful to 
+use a pilot job framework to run Spark workflows on HPC? Also, this comes a bit late in the intro, it would be good
+to precise the question early on, and the explain it in length as you do.}
 
@@
 
+\tristan{Your text now mentions points 3 and 4 below, but doesn't talk about 1 and 2 and the fact that
+an overlay cluster already deals with that. Something that wasn't mentioned in my initial list was that pilot jobs
+also reduce the number of tasks seen by the cluster (this is also addressed by an overlay scheduler). It would be interesting
+to explain that many of the interests of pilot jobs are already addressed by overlay scheduling, hence the question whether
+pilot jobs are useful at all in this context.}
 \TG{The introduction is still very much geared toward the presentation of 

### 2019-05-03 paper.tex
some methods text

--- 2019-05-02 paper.tex
@@
 \section{Materials and Methods}\label{sec:methods}
-    SPA consists of a Python-based driver with bash-templates that are called 
-    using the Slurmpy~\cite{slurmpy} library. In order to commence pilot-scheduling,
-    the user must supply the bash template of choice and a JSON configuration 
-    file which contains details on pilot node requirements and the desired application
-    to be launched. Currently, there only exists a single pilot template for launching
-    Spark jobs on a Slurm-based HPC cluster. The pseudocode for that template can 
-    be seen in ~\ref{Figure:pilot-temp}.
+
+    In order to compare and contrast pilot over batch scheduling for scientific 
+    computing, we first created a synthetic application that would process the 
+    Big Brain~\cite{bigBrain}, a 76G 3D histological image of a human brain. The algorithm
+    is described in~\ref{Figure:incrementation}. This algorithm was selected as 
+    it would give us the flexibility to easily vary computing time as needed.
+
+    For the batch submission script, we reused the once recommended on the
+    compute canada wiki ~\cite{wiki}.
+
     \todo{walltime parameter of workflow}

### 2019-05-06 paper.tex

@@
 \hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
+\usepackage{algorithm}
+\usepackage{algorithmicx} % Doc is at http://tug.ctan.org/macros/latex/contrib/algorithmicx/algorithmicx.pdf
+\usepackage{algpseudocode}
+\algnewcommand\algorithmicforeach{\textbf{for each}}
+\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
+\algblock{Input}{EndInput}
+\algnotext{EndInput}
+\algblock{Output}{EndOutput}
+\algnotext{EndOutput}
+\newcommand{\Desc}[2]{\State \makebox[2em][l]{#1}#2}
 
@@
 
-    In order to compare and contrast pilot over batch scheduling for scientific 
-    computing, we first created a synthetic application that would process the 
-    Big Brain~\cite{bigBrain}, a 76G 3D histological image of a human brain. The algorithm
-    is described in~\ref{Figure:incrementation}. This algorithm was selected as 
-    it would give us the flexibility to easily vary computing time as needed.
+    The application, templates, configuration files, benchmarks and analysis
+    scripts are publicly available and can be found at:
+    \href{https://github.com/big-data-lab-team/spa}{https://github.com/big-data-lab-team/spa}.
+    Links to the processing engines and processed data are provided in the text.
 
-    For the batch submission script, we reused the once recommended on the
-    compute canada wiki ~\cite{wiki}.
+    \subsection{Infrastructure}
+    All experiments were conducted on the Cedar and B\'eluga HPC computing clusters
+    made available by Compute Canada. The Cedar cluster has a total of 1542 nodes
+    with a total of 58,416 CPU cores. Available memory on a Cedar node can range
+    from 125 to 3022~G. Standard nodes are equipped with either 2x Intel E5-2683 v4
+    Broadwell @ 2.1~Ghz (32 cores total) or 2x Intel Platinum 8160F Skylake @
+    2.1Ghz (48 cores total) CPUs and 
+    2 x 480~G SSD. All nodes and temporary storage on Cedar is connected by an 
+    Intel OmniPath (version 1) interconnect with 100Gbit/s bandwidth.
+
+    B\'eluga, on the other hand, is a smaller cluster with 872 available nodes. 
+    Node memory can range between 92 to 752~G, with the most common node type having 
+    186G of available memory. All nodes contain a total of 2 x Intel Gold 6148
+    Skylake @ 2.4~Ghz (40 cores/node) CPU and are connected to each other with a
+    100~Gb/s Mellanox Infiniband EDR network. Each non-GPU node type contains one 
+    480~G SSD.
+
+    \subsection{Templates}
+
+    Two different templates were used for our experiments: the batch submission
+    template and the pilot submission template. The batch submission template was
+    inspired by the template provided by Compute Canada to launch Spark applications 
+    on Slurm \href{https://docs.computecanada.ca/wiki/Apache\_Spark/en}{available here}.
+    The template operates as follows: Certain resource requirements are requested by
+    the user (e.g. walltime, amount of memory per node, number of CPUs per task, number 
+    of nodes and number of tasks per node). Once these resources are allocated, a 
+    master is started on one of the requested node resources. Then, after the master
+    has successfully started, the workers are started on all nodes. Multiple worker
+    instances are started on a single node by setting the \textit{SPARK\_WORKER\_INSTANCES}
+    environment variable to the number of tasks per node. Each worker is given as many as 
+    many cores as specified by the user in the SLURM resource allocation request.
+    After both the masters and the workers have successfully started, the driver is finally
+    started. The amount of memory given to each executor corresponds to 95\% of the 
+    available memory on the node, to allow for offheap space. 
+    The deploy-mode selected for the batch template is client-mode.
+
+    The pilot submission template is similar to that of the batch template, however,
+    each pilot will start its own Spark master and worker. However, only one pilot,
+    the first pilot to attempt to launch the driver, will do so. The reason for 
+    which each pilot starts its own master is to ensure the fault tolerance of the 
+    masters. In this sort of configuration, should the active master be killed, one
+    of the stand-by masters will takeover and the application will be able to 
+    resume. Such a configuration is particularly favourable in pilot scheduling 
+    scenarios as node failures may be more frequent do to walltime expiration.
+    Additionally, the deploy mode of the driver was selected to be cluster deploy mode.
+    This would not only allow the driver to be executed directly on one of the workers,
+    but also allow us to make the driver fault-tolerant through the \textit{supervise} 
+    mode, which is only available in cluster deploy. As with the masters, it is
+    particularly important to have a fault-tolerant driver in pilot-scheduling scenarios due walltime
+    expiration. Should pilots be idle for a certain duration, the pilots will terminate
+    themselves such as to not hog resources.
+
+    Due to the differences in deploy modes between batch and pilot submission, 
+    batch will always inevitably have one more worker than pilot. This is because
+    in cluster deploy, which pilot uses, the driver occupies a worker, whereas in
+    client deploy, the driver is separate from any worker.
+
+    Both of these Slurm templates are launched within a Python application called
+    \textit{SPA}. The templates are used in conjunction with JSON configuration 
+    file and passed to the SlurmPy~\cite{slurmpy} library within \textit{SPA}.
+    The \textit{SPA} application, all the while ensures that all is preconfigured
+    correctly before passing it to SlurmPy. It also ensures that enough pilots are
+    launched, maintains track of the running/queued pilots, and launches additional
+    pilots if there are less pilots than requested by the user in the Slurm queue.
+
+    \subsection{Application}
+        \begin{algorithm}\caption{Incrementation}\label{alg:incrementation}              
+            \begin{algorithmic}[1]                                                       
+                \Input                                                                       
+                    \Desc{$x$}{a sleep delay in seconds}                                         
+                    \Desc{$n$}{a number of iterations}                                           
+                    \Desc{$C$}{a set of image chunks}                                            
+                \EndInput                                                                    
+                \ForEach{$chunk \in C$}                                                      
+                    \State read $chunk$                                        
+                    \For{$i \in [1, n]$}                                                         
+                        \State $chunk\gets chunk+1$                                              
+                        \State sleep $x$                                                         
+                    \EndFor                                                                      
+                    \State write $chunk$                                            
+                \EndFor                                                                      
+            \end{algorithmic}                                                                
+        \end{algorithm}
+    To determine the added value of pilot scheduling over batch scheduling of Spark
+    applications, we required a Spark application operating on a large dataset with
+    an important processing time to emulate a what would be the average requirements
+    of a scientific Spark application. For this, we created a synthetic application 
+    that would process the Big Brain~\cite{bigBrain}, a 76G 3D histological
+    image of a human brain. The algorithm is a chain of map transformations that
+    at each transformation increment the voxels of the image by 1 (see Figure~\ref{alg:incrementation}.
+    We chose such a synthetic algorithm
+    as the focus of our experiments is pilot scheduling and not the application in
+    itself. Furthermore, this algorithm enabled us to have control over the task duration
+    which was representative of scientific applications. Additionally, it was important
+    that the overall application duration did not vary between the different levels of
+    parallelism within our experiments. Being able to adjust the task duration based on
+    level of parallelism allowed us to achieve this.
+
+    Investigating the fault-tolerance provided by the overlay cluster is
+    crucial in determining the suitable of pilot scheduling with overlay clusters.
+    Fault-tolerance of the driver is only possible in cluster deploy mode, however,
+    when using Spark's standalone scheduler, this mode is not available for Python 
+    applications. It is for this reason that our synthetic application is written
+    in Scala. Nevertheless, cluster deploy mode is possible for Python application
+    using YARN or Mesos schedulers.
 
     \todo{walltime parameter of workflow}
+    \subsection{Model}
+
+    The makespan of an application can be defined as the total duration of the 
+    application including any scheduler queueing time. It can be defined by
+    the following expression:
+
+    \begin{equation}
+        M = \frac{C}{W} \label{eq:mcw}
+    \end{equation}
+    where:
+    \begin{itemize}
+        \item $M$ is the makespan of the application
+        \item $C$ is the total CPU time of the application
+        \item $W$ is the average number of workers
+    \end{itemize}
+
+    The average number of workers allows us to obtain a more accurate makespan
+    figure as the average number of workers will decrease with increasing 
+    queuing times. Should the average number of workers equal the number of workers
+    requested, the application was not subjected to any scheduling queueing time.
+
+    Furthermore, should the total CPU time of the application be fixed, the relation
+    between batch and pilot jobs can be represented as:
+
+    \begin{equation}
+        \frac{M_{batch}}{M_{pilot}} = f(\frac{W_{pilot}}{W_{batch}})\label{eq:makespancomp}
+    \end{equation}
+    where:
+    \begin{itemize}
+        \item $M_{batch}$ is the makespan of the batch application
+        \item $M_{pilot}$ is the makespan of the pilot application
+        \item $W_{pilot}$ is the average number of workers of the pilot application
+        \item $W_{batch}$ is the average number of workers of the batch application
+    \end{itemize}
+
+    \todo{perhaps expand a bit here}
+
     \subsection{Added value of pilot scheduling}
+        To determine if there are any performance benefits of using pilot over 
+        batch scheduling, we needed to compare both strategies given various resource
+        requirements. It is expected that large batch requests will stay in the 
+        resource queue longer than multiple pilot requests that ultimately use up
+        the same amount of resources as each individual pilot requests less resources at
+        a time. We therefore used four resource configurations to investigate this
+        hypothesis (Table~\ref{table:dedicatednodes}).
+
+        \begin{table}                                                                    
+\centering                                                                       
+\begin{tabular}{c|c|c}                                                             
+\rowcolor{headcolor}                                                             
+Configuration & RAM (GB) & Tasks \\                               
+\hline                                                                           
+1 & 112 & 16 \\                                               
+2 & 224 & 32 \\                                               
+3 & 336 & 48 \\
+4 & 448 & 64 \\
+\end{tabular}                                                                    
+\setlength{\belowcaptionskip}{-10pt}                                             
+\caption{Resource configurations}                                                    
+\label{table:dedicatednodes}                                                            
+\end{table} 
         \todo{execution of a batch cluster vs dynamic (pilot) cluster}
@@
         \todo{incrementation with varying task durations}
+\section{Results}\label{sec:Results}
 \section{Discussion}\label{sec:discussion}

### 2019-05-06 paper.tex
Comments and notes on Section II

@@
 
-    The application, templates, configuration files, benchmarks and analysis
-    scripts are publicly available and can be found at:
+    The application, templates, configuration files, benchmarks and
+    analysis scripts are publicly available and can be found in our Spark
+    Pilot-job scheduler for HPC Applications (SPA) software at:
     \href{https://github.com/big-data-lab-team/spa}{https://github.com/big-data-lab-team/spa}.
-    Links to the processing engines and processed data are provided in the text.
+    Links to the processing engines and processed data are provided in the
+    text.
 
@@
     All experiments were conducted on the Cedar and B\'eluga HPC computing clusters
-    made available by Compute Canada. The Cedar cluster has a total of 1542 nodes
+    made available by Compute Canada through WestGrid and Calcul Qu\'ebec \TG{add links}. Both 
+    clusters are accessible through the Slurm batch scheduler and Lustre parallel file system.
+    The Cedar cluster has a total of 1542 nodes
     with a total of 58,416 CPU cores. Available memory on a Cedar node can range
@@
     2.1Ghz (48 cores total) CPUs and 
-    2 x 480~G SSD. All nodes and temporary storage on Cedar is connected by an 
-    Intel OmniPath (version 1) interconnect with 100Gbit/s bandwidth.
+    2 x 480~G SSD. All nodes and temporary storage on Cedar are connected by an 
+    Intel OmniPath (version 1) with 100Gbit/s bandwidth.
 
@@
     Node memory can range between 92 to 752~G, with the most common node type having 
-    186G of available memory. All nodes contain a total of 2 x Intel Gold 6148
+    186G. All nodes contain 2 x Intel Gold 6148
     Skylake @ 2.4~Ghz (40 cores/node) CPU and are connected to each other with a
-    100~Gb/s Mellanox Infiniband EDR network. Each non-GPU node type contains one 
+    100~Gb/s Mellanox Infiniband EDR network \TG{expand EDR}. Each non-GPU node type contains one 
     480~G SSD.
 
-    \subsection{Templates}
+    \subsection{Job Templates}
 
-    Two different templates were used for our experiments: the batch submission
+    Two different job templates were used to implement the two main
+    conditions compared in our experiments: the batch submission
     template and the pilot submission template. The batch submission template was
@@
     instances are started on a single node by setting the \textit{SPARK\_WORKER\_INSTANCES}
-    environment variable to the number of tasks per node. Each worker is given as many as 
-    many cores as specified by the user in the SLURM resource allocation request.
+    environment variable to the number of tasks per node. Each worker is given as many
+     cores as specified by the user in the Slurm resource allocation request.
     After both the masters and the workers have successfully started, the driver is finally
@@
     available memory on the node, to allow for offheap space. 
-    The deploy-mode selected for the batch template is client-mode.
+    The Spark deploy-mode selected for the batch template is client-mode.
 
     The pilot submission template is similar to that of the batch template, however,
-    each pilot will start its own Spark master and worker. However, only one pilot,
+    each pilot will start its own Spark master and worker. Only one pilot,
     the first pilot to attempt to launch the driver, will do so. The reason for 
@@
     resume. Such a configuration is particularly favourable in pilot scheduling 
-    scenarios as node failures may be more frequent do to walltime expiration.
-    Additionally, the deploy mode of the driver was selected to be cluster deploy mode.
+    scenarios as node failures may be more frequent due to walltime expiration.
+    Additionally, the Spark deploy mode of the driver was selected to be cluster deploy mode.
     This would not only allow the driver to be executed directly on one of the workers,
-    but also allow us to make the driver fault-tolerant through the \textit{supervise} 
-    mode, which is only available in cluster deploy. As with the masters, it is
-    particularly important to have a fault-tolerant driver in pilot-scheduling scenarios due walltime
-    expiration. Should pilots be idle for a certain duration, the pilots will terminate
-    themselves such as to not hog resources.
+    but also allow us to make the driver fault-tolerant through the
+    \textit{supervise} mode, which is only available in cluster deploy
+    \TG{A brief section on Spark to explain these modes and features would
+    be useful}. As with the masters, it is particularly important to have a
+    fault-tolerant driver in pilot-scheduling scenarios due to possible
+    walltime expiration \TG{It should be explained before that one of the interests of pilots
+    is to ``play'' with shorter walltimes}. \TG{It would also be nice to have runs where pilots 
+    have shorter walltimes, to see if queuing times are shorter too.}
+    Should pilots be idle for a certain duration, the
+    pilots will terminate themselves such as to not hog resources.
 
@@
                         \State $chunk\gets chunk+1$                                              
-                        \State sleep $x$                                                         
+                        \State sleep $x$ \TG{Don't you sleep by x minus the duration of the previous step?}                                                        
                     \EndFor                                                                      
@@
     of a scientific Spark application. For this, we created a synthetic application 
-    that would process the Big Brain~\cite{bigBrain}, a 76G 3D histological
+    that would process the Big Brain~\cite{bigBrain}, a 76~G 3D histological
     image of a human brain. The algorithm is a chain of map transformations that
-    at each transformation increment the voxels of the image by 1 (see Figure~\ref{alg:incrementation}.
+    at each transformation increment the voxels of the image by 1 (see Algorithm~\ref{alg:incrementation}).
     We chose such a synthetic algorithm
@@
     Investigating the fault-tolerance provided by the overlay cluster is
-    crucial in determining the suitable of pilot scheduling with overlay clusters.
+    crucial in determining the suitability of pilot scheduling with overlay clusters.
     Fault-tolerance of the driver is only possible in cluster deploy mode, however,
@@
 
-    The makespan of an application can be defined as the total duration of the 
-    application including any scheduler queueing time. It can be defined by
-    the following expression:
+    The makespan of an application can be defined as the total duration
+    between the submission time of the first application task, and the
+    completion time of the last application time. It includes any
+    scheduling time, queueing time, data transfer time, and any other
+    overhead.
 
+    Assuming a divisible load, i.e., the application can be divided in any
+    number of tasks, the makespan can be expressed using the following
+    expression, which holds for both batch and pilot execution modes:
     \begin{equation}
@@
         \item $C$ is the total CPU time of the application
-        \item $W$ is the average number of workers
+        \item $W$ is the average number of Spark workers throughout the execution
     \end{itemize}
+    The average number of workers $W$ allows us to obtain a more accurate makespan
+    figure as the average number of workers will decrease with increasing 
+    queuing times. It is computed as follows:
+    $$
+    W = \frac{1}{M}\int_0^M{w(t)dt},
+    $$
+    where $w(t)$ is the number of workers available at time $t$. When the
+    application is not subject to any scheduling or queuing time, the
+    average number of workers equal the number of workers requested.
 
-    The average number of workers allows us to obtain a more accurate makespan
-    figure as the average number of workers will decrease with increasing 
-    queuing times. Should the average number of workers equal the number of workers
-    requested, the application was not subjected to any scheduling queueing time.
-
-    Furthermore, should the total CPU time of the application be fixed, the relation
+    Therefore, assuming a fixed total CPU time, the relation
     between batch and pilot jobs can be represented as:
-
     \begin{equation}
-        \frac{M_{batch}}{M_{pilot}} = f(\frac{W_{pilot}}{W_{batch}})\label{eq:makespancomp}
+        \frac{M_{batch}}{M_{pilot}} = \frac{W_{pilot}}{W_{batch}}\label{eq:makespancomp}
     \end{equation}
@@
     \end{itemize}
+    We will use this relation to discuss our results later on. It
+    corresponds to an ideal case where no data or other overhead is
+    present: only queuing times are included.
 
-    \todo{perhaps expand a bit here}
+    \todo{perhaps expand a bit here \TG{I have a formal proof for eq (1) but I don't think
+    it's required here.}}
 
@@
 \section{Discussion}\label{sec:discussion}
+
+SPA vs other frameworks.
+
+Technical overhead of using pilots + an overlay cluster over a cluster.
+
 \section{Conclusion}\label{sec:conclusion}

### 2019-05-08 experiments/code/launch_bench.py
added no interleaving option

@@ -6,6 +6,7 @@ from random import shuffle
 from tempfile import TemporaryFile, NamedTemporaryFile
 import time
 import shutil
+import sys
 
 
 iterations = 5
@@ -88,32 +89,51 @@ def get_results(sp, tmp_file, launch, cond, out_dir):
 
 count = 0 
 while count < iterations :
-    shuffle(experiments)
-    for exps in experiments:
-        print(exps['batch'])
-        print(exps['8pilot'])
-        print(exps['16pilot'])
-        f_batch = NamedTemporaryFile(mode='w+', prefix='batch_', suffix='_{}'.format(exps['cond']))
-        f_8pilot = NamedTemporaryFile(mode='w+', prefix='pilot8_', suffix='_{}'.format(exps['cond']))
-        f_16pilot = NamedTemporaryFile(mode='w+', prefix='pilot16', suffix='_{}'.format(exps['cond']))
-        options = [('batch', f_batch, batch_out),
-                   ('8pilot', f_8pilot, pilot8_out),
-                   ('16pilot', f_16pilot, pilot16_out)]
+    if sys.argv[1] == "--no_interleave":
+        print("running without interleaving")
+        exps = experiments[2]
+
+        options = [('batch', batch_out), ('8pilot', pilot8_out), ('16pilot', pilot16_out)]
         shuffle(options)
-        x = Popen(exps[options[0][0]], bufsize=1, stdout=options[0][1], stderr=options[0][1])
-        y = Popen(exps[options[1][0]], bufsize=1, stdout=options[1][1], stderr=options[1][1])
-        z = Popen(exps[options[2][0]], bufsize=1, stdout=options[2][1], stderr=options[2][1])
+        
+        for opt in options:
+            f = NamedTemporaryFile(mode='w+', prefix='{}_'.format(opt[0]), suffix='_{}'.format(exps['cond']))
+            x = Popen(exps[opt[0]], bufsize=1, stdout=f, stderr=f)
+            get_results(x, f, opt[0], exps['cond'], opt[1])
 
-        get_results(x, options[0][1], options[0][0], exps['cond'], options[0][2])
-        get_results(y, options[1][1], options[1][0], exps['cond'], options[1][2])
-        get_results(z, options[2][1], options[2][0], exps['cond'], options[2][2])
+            try:
+                shutil.rmtree(opt[1], ignore_errors=True)
 
-        try:
-            shutil.rmtree(batch_out, ignore_errors=True)
-            shutil.rmtree(pilot8_out, ignore_errors=True)
[…]

### 2019-05-08 experiments/code/prep_benchmark_scripts.sh
added script for task benchmark cleanup

@@ -1,0 +1,16 @@
+#!/usr/bin/bash
+
+for i in $(find . -regextype "egrep" -iregex './[0-9][0-9][0-9]+.*applogs'); do
+	slurm_id=$(echo $i | sed -e 's/-applogs//g' -e 's/\.\///g')
+	log=$(find . -regextype "egrep" -iregex "./logs/.*$slurm_id\.out")
+	task=$(echo $log | sed -e 's/\.\/logs\///g' -e 's/-.*//g')
+	out="$task-$slurm_id.txt"
+	out_new="$task-$slurm_id.csv"
+
+	echo "slurm_id: $slurm_id \t task: $task \t outfile: $out"
+	grep -H '' $i/* > $out
+
+	sed -e 's/:/,/g' -e 's/\.int.*txt//g' -e 's/.*applogs\///g' <$out >$out_new
+	rm $out
+
+done

### 2019-05-08 paper.tex

@@
     has successfully started, the workers are started on all nodes. Multiple worker
-    instances are started on a single node by setting the \textit{SPARK\_WORKER\_INSTANCES}
+    instances are started on a single node by setting the \texttt{SPARK\_WORKER\_INSTANCES}
     environment variable to the number of tasks per node. Each worker is given as many
@@
     but also allow us to make the driver fault-tolerant through the
-    \textit{supervise} mode, which is only available in cluster deploy
+    \texttt{supervise} mode, which is only available in cluster deploy
     \TG{A brief section on Spark to explain these modes and features would
@@
     \subsection{Added value of pilot scheduling}
-        To determine if there are any performance benefits of using pilot over 
-        batch scheduling, we needed to compare both strategies given various resource
-        requirements. It is expected that large batch requests will stay in the 
-        resource queue longer than multiple pilot requests that ultimately use up
-        the same amount of resources as each individual pilot requests less resources at
-        a time. We therefore used four resource configurations to investigate this
-        hypothesis (Table~\ref{table:dedicatednodes}).
-
         \begin{table}                                                                    
@@
         \end{table}
-        \todo{execution of a batch cluster vs dynamic (pilot) cluster}
-    \subsection{Robust masters}
+
+        \begin{table*}                                                                   
+        \centering                                                                       
+        \begin{tabular}{c|cccccc}                                                   
+          \rowcolor{headcolor}                                                           
+          \multicolumn{7}{c}{Configuration 1}\\                      
+          \hline                                                                         
+          \rowcolor{headcolor}                                                           
+          execution mode & nodes/job & RAM (G) & CPUs per task & tasks/node & walltime & task delay (s) \\                             
+          \hline
+          Batch & 1 & 112 & 1 & 16 & 2h30 & 45 \\
+          8 pilots & 1 & 14 & 1 & 2 & 2h30 & 45 \\
+          16 pilots & 1 & 7 & 1 & 1 & 2h30 & 45 \\
+
+          \hline                                                                           
+          \multicolumn{7}{c}{}\\
+
+          \rowcolor{headcolor}                                                           
+          \multicolumn{7}{c}{Configuration 2}\\                      
+          \hline                                                                         
+          \rowcolor{headcolor}                                                           
+          execution mode & nodes/job & RAM (G) & CPUs per task & tasks/node & walltime & task delay (s) \\                             
+          \hline
+          Batch & 2 & 112 & 1 & 16 & 2h30 & 90 \\
+          8 pilots & 1 & 28 & 1 & 4 & 2h30 & 90 \\
+          16 pilots & 1 & 14 & 1 & 2 & 2h30 & 90 \\
+
+          \hline                                                                           
+          \multicolumn{7}{c}{}\\
+
+          \rowcolor{headcolor}                                                           
+          \multicolumn{7}{c}{Configuration 3}\\                      
+          \hline                                                                         
+          \rowcolor{headcolor}                                                           
+          execution mode & nodes/job & RAM (G) & CPUs per task & tasks/node & walltime & task delay (s) \\                             
+          \hline
+          Batch & 3 & 112 & 1 & 16 & 2h30 & 120 \\
+          8 pilots & 1 & 42 & 1 & 6 & 2h30 & 120 \\
+          16 pilots & 1 & 21 & 1 & 3 & 2h30 & 120 \\
+
+          \hline                                                                           
+          \multicolumn{7}{c}{}\\
+
+          \rowcolor{headcolor}                                                           
+          \multicolumn{7}{c}{Configuration 4}\\                      
+          \hline                                                                         
+          \rowcolor{headcolor}                                                           
+          execution mode & nodes/job & RAM (G) & CPUs per task & tasks/node & walltime & task delay (s) \\                             
+          \hline
+          Batch & 4 & 112 & 1 & 16 & 2h30 & 180 \\
+          8 pilots & 1 & 56 & 1 & 8 & 2h30 & 180 \\
+          16 pilots & 1 & 28 & 1 & 4 & 2h30 & 180 \\
+
+          \hline                                                                           
+          \multicolumn{7}{c}{}\\                                                        
+        \end{tabular}                                                                    
+        \setlength{\belowcaptionskip}{-10pt}                                             
+        \caption{Experimental conditions}
+        \label{table:conditions}                                                        
+        \end{table*}                                                                                   
+        To determine if there are any performance benefits to using pilot over 
+        batch scheduling, we needed to compare both strategies given various resource
+        requirements. It is expected that large batch requests will stay in the 
+        resource queue longer than multiple pilot requests that ultimately use up
+        the same amount of resources as each individual pilot requests less resources at
+        a time. We therefore used four resource configurations to investigate this
+        hypothesis (Table~\ref{table:dedicatednodes}).
+
+       For batch, we requested 1 to 4 dedicated nodes, depending on resource 
+       configuration, each with 112~G of RAM, 1 core per task and 16 tasks per 
+       node. On the other hand, for all configurations, we ran our experiments
+       with 8 and 16 pilots. Experimental conditions can be seen in Table~\ref{table:conditions}.
+
+       As the focus of our experiments here is to measure the impact pilot scheduling
+       has on queueing times, we wanted to ensure that our walltime estimates would
+       be consistent for each experimental condition. Therefore, we adjusted task 
+       delay based on the maximum level of parallelism to ensure that walltimes would
+       not need to be readjusted for each configuration. Given utmost parallelism,
+       the sleep delay added would amount to 1 hour of processing time regardless
+       of configuration. The real processing time of the application, however, 
+       would differ between configurations as we did not account for task duration
+       in the sleep delay. However, task duration was not expected to signicantly impact
+       overall processing time, but nevertheless, we set application walltime to 
+       2 hours and 30 minutes for all configurations to account for additional
+       processing time that may be incurred in applications with lower levels of 
+       parallelism.
+
+       For each configuration, all three execution modes were executed in parallel
+       to ensure that the status of the overall cluster was the same when the 
+       different execution modes were launched. Furthermore, the order of the execution modes
+       for each configuration was randomized to ensure that SLURM job request
+       order could not have affected results.
+
+       There were 15 repetitions in total for each configuration, and the order
+       in which the configurations were launched was also randomized. This was to 
+       account for any system variability that can occur, particularly in production
+       HPC clusters.
+
+       Different clusters may have different resource configurations, number of 
+       users and scheduling policies. Therefore, we executed all 15 repetitions on
+       both Cedar and B\'eluga to determine how much our results differ between
+       two distinct clusters.
+
+    \subsection{Pipeline robustness}
+        While nodes failures can occur for various reasons, underestimation of 
+        walltime may lead to early node termination in pilot scheduling systems.
+        Therefore, it is crucial for the overlay cluster within the pilot system
+        to be robust to node failures. In this series of experiments, we investigate
+        Spark's fault-tolerance with respect to worker, master and driver termination.
+
+        In our configuration, the driver is running in cluster deploy-mode, resulting
+        in it being executed directly on a worker. This feature alone should not make
+        it fault-tolerant to failures. However, using this feature coupled with the 
+        \texttt{--supervise} flag, will.
+
         \todo{scala standalone vs our pyspark workaround. kill masters in experiments}

### 2019-05-09 example/standalone/hpc/hpc_pilot_template.sh
added api retries and modifies pilot idle duration

@@ -128,7 +128,7 @@ pilot_program(){
 			sleep 5
     		fi
 
-    		[ $elapsed_time -gt 120 ] && break
+    		[ $elapsed_time -gt 1000000 ] && break
 
 	done
 	$SPARK_HOME/sbin/stop-slave.sh

### 2019-05-09 pilotspark.py

@@ -31,6 +31,17 @@ from copy import deepcopy
 import glob
 
 
+def query_api(url, retries=5):
+    for i in range(retries):
+        logging.info("Attempting to query Spark logs... Attempt #%d", i)
+        try:
+            return requests.get(url)
+        except Exception as e:
+            logging.error(str(e))
+            time.sleep(10)
+
+
+
 def write_bench_start(bench):
     with open(bench, 'a+') as f:
         f.write(",".join([os.linesep + datetime.now().isoformat(),
@@ -298,7 +309,7 @@ def submit_pilots(template, conf):
             time.sleep(5)
 
         try:
-            r = requests.get(driver_rest)
+            r = query_api(driver_rest)
             driver_api = r.json()
             logging.debug("Driver status: %s", driver_api)
             driver_id = driver_api["submissionId"]
@@ -311,7 +322,7 @@ def submit_pilots(template, conf):
             time.sleep(5)
 
             try:
-                r = requests.get(driver_rest)
+                r = query_api(driver_rest)
                 driver_api = r.json()
                 logging.debug(driver_api)
                 driver_id = driver_api["submissionId"]
@@ -349,7 +360,7 @@ def submit_pilots(template, conf):
             time.sleep(5 * 60)
 
             try:
-                r = requests.get(driver_rest)
+                r = query_api(driver_rest)
                 driver_api = r.json()
 
                 result = driver_api["driverState"]

### 2019-05-09 example/standalone/hpc/hpc_pilot_template.sh
prep for testing cluster failures

@@ -26,6 +26,7 @@ term_handler()
 trap 'term_handler' TERM
 
 pilot_program(){
+	echo "Starting Master..."
 	$SPARK_HOME/sbin/start-master.sh
 	if [ ! -f $mstr_log ]; then
     		lockfile -r 0 $mstr_lock

### 2019-05-09 pilotspark.py

@@ -187,6 +187,22 @@ def submit_sbatch(template, conf):
         write_bench_end(conf["benchmark"])
         write_bench_result(conf["benchmark"], result)
 
+def get_queued(jobs):
+
+    p = Popen(["squeue", "-j", ",".join(jobs)], stdout=PIPE, stderr=PIPE)
+    (out, err) = p.communicate()
+    out = str(out, 'utf-8')
+    err = str(err, 'utf-8')
+    logging.debug("stdout: %s", out)
+    logging.debug("stderr: %s", err)
+
+    if 'send/recv' in out or 'send/recv' in err:
+        return ['']
+
+    running_jobs = out.split(os.linesep)[1:]
+    running_jobs = [l.strip().split(' ')[0] for l in running_jobs if l.strip().split(' ')[0] != '']
+    return running_jobs
+
 
 def submit_locally(template, conf):
 
@@ -288,13 +304,17 @@ def submit_pilots(template, conf):
 
         while not op.isfile(conf["COMPUTE"]["drvr_log"]):
             logging.warning("Driver log %s not created. Sleeping for 5 seconds",
-                         conf["COMPUTE"]["drvr_log"])
+                            conf["COMPUTE"]["drvr_log"])
+            if len(get_queued(jobs)) < 1:
+                break
             time.sleep(5)
 
-        while "http" not in driver_rest:
+        while "http" not in driver_rest and op.isfile(conf["COMPUTE"]["drvr_log"]):
             with open(conf["COMPUTE"]["drvr_log"], 'r') as f:
                 driver_rest = f.readline().strip(os.linesep)
                 logging.info('Driver URL at: %s', driver_rest)
+            if len(get_queued(jobs)) < 1:
+                break
             time.sleep(5)
 
         try:
@@ -308,6 +328,8 @@ def submit_pilots(template, conf):
         while (driver_api is not None and "driverState" in driver_api
                and driver_api["driverState"] == "SUBMITTED"):
             logging.warning('Driver in SUBMITTED state. Sleeping for 5 seconds')
+            if len(get_queued(jobs)) < 1:
+                break
             time.sleep(5)
 
             try:
@@ -322,19 +344,8 @@ def submit_pilots(template, conf):
         while (driver_api is not None and "driverState" in driver_api
                and driver_api["driverState"] == "RUNNING"):
             logging.info('Driver currently in RUNNING state')
-            p = Popen(["squeue", "-j", ",".join(jobs)], stdout=PIPE, stderr=PIPE)
[…]

### 2019-05-11 experiments/code/raw_to_json.py
latest results cedar

@@ -3,6 +3,7 @@
 import argparse
 from os import listdir, path as op, linesep
 from time import strptime
+from datetime import datetime as dt
 import json
 import glob
 
@@ -49,20 +50,23 @@ def get_jobs(fn, sj_benchmark_dir, s_logs, exec_mode):
 
             if driver_id is not None:
                 driver_path = op.join(worker_logdir, driver_id)
+                #driver_path = op.join('sworker_logs', driver_id)
 
             for sj in sjids:
                 sj_elem = {}
                 sj_elem['id'] = sj
                 sj_elem['start_time'] = None
                 sj_elem['end_time'] = None
-                nodes, success = get_jobid_success(sj, s_logs)
+                nodes, success, job_start = get_jobid_success(sj, s_logs)
                 sj_elem['nodes'] = nodes
 
                 if success or driver_id is None:
                     sj_elem['succeeded'] = success
+                    sj_elem['job_start'] = job_start
                 else:
-                    success = get_success(driver_path, list(nodes.keys())[0] if len(nodes) > 0 else None)
+                    success, job_start = get_success(driver_path, list(nodes.keys())[0] if len(nodes) > 0 else None)
                     sj_elem['succeeded'] = success
+                    sj_elem['job_start'] = job_start
 
                 sjelems.append(sj_elem)
 
@@ -85,12 +89,17 @@ def get_jobs(fn, sj_benchmark_dir, s_logs, exec_mode):
 
 def get_success(fp, node):
     success = False
+    job_start = None
+    print(fp)
     if op.isdir(fp) and node is not None:
         with open(op.join(fp, 'stderr'), 'r') as f:
             for line in f:
                 if 'Finished task' in line and '125/125' in line:
                     success = True
-    return success
+                elif 'Starting job: collect at IncrementApp.scala' in line:
+                    job_start = dt.timestamp(dt.strptime(linesep.join(line.split(' ')[:2]), "%y/%m/%d %H:%M:%S"))
+
+    return success, job_start
 
 
 def order_pilots(directory, sjids, exec_mode="batch"):
@@ -180,6 +189,7 @@ def get_jobid_success(job_id, master_logs):
     logfile = glob.glob(op.join(op.abspath(master_logs), '*.{}.out'.format(job_id)))
     batch = False
     executors = {}
+    job_start = None
 
     if len(logfile) > 0:
         if 'batch' in logfile[0]:
@@ -216,15 +226,18 @@ def get_jobid_success(job_id, master_logs):
                         executors[node].add(proc)
                     else:
[…]

### 2019-05-13 example/standalone/hpc/hpc_pilot_template.sh
fixed driver output file name

@@ -84,9 +84,11 @@ pilot_program(){
 			app_log_dir=${SLURM_JOBID}-applogs
 			mkdir $app_log_dir
 
-			eval $driver_prog > output 2>&1
-			cat output
-			driverid=`cat output | grep submissionId | grep -Po 'driver-\d+-\d+'`
+			out_file=output-$SLURM_JOBID
+
+			eval $driver_prog > $out_file 2>&1
+			cat $out_file
+			driverid=`cat $out_file | grep submissionId | grep -Po 'driver-\d+-\d+'`
 			echo 'Spark driver ID: ' $driverid
 
 			DRIVER_REST=${MASTER_URI/spark/http}
@@ -94,6 +96,7 @@ pilot_program(){
 			echo 'REST API Url: ' $DRIVER_REST
 			curl $DRIVER_REST
 			echo $DRIVER_REST > $drvr_log
+			rm $out_file
     		fi
 	fi
 

### 2019-05-15 experiments/code/raw_to_json.py
fixed raw_to_json code and beluga data

@@ -30,7 +30,7 @@ def get_jobs(fn, sj_benchmark_dir, s_logs, exec_mode):
     log_files = glob.glob(op.join(fn, '*{}*.out'.format(exec_mode)))
 
     for f in sorted(log_files, key=lambda x: op.basename(x).split('-')[0]):
-        print(f)
+        print('logfile', f)
         with open(f, 'r') as logfile:
             sjids = []
             sjelems = []
@@ -47,8 +47,10 @@ def get_jobs(fn, sj_benchmark_dir, s_logs, exec_mode):
                     worker_logdir = line.split(' ')[-1].strip(linesep)
                 elif "'submissionId': " in line:
                     driver_id = line.split("'submissionId': ")[1].split(',')[0].strip("'")
+                elif "GET /v1/submissions/status/" in line:
+                    driver_id = line.split("GET /v1/submissions/status/")[1].split(' ')[0]
 
-            if driver_id is not None:
+            if driver_id is not None and 'driver' in driver_id:
                 driver_path = op.join(worker_logdir, driver_id)
                 #driver_path = op.join('sworker_logs', driver_id)
 
@@ -60,7 +62,7 @@ def get_jobs(fn, sj_benchmark_dir, s_logs, exec_mode):
                 nodes, success, job_start = get_jobid_success(sj, s_logs)
                 sj_elem['nodes'] = nodes
 
-                if success or driver_id is None:
+                if success or driver_path is None:
                     sj_elem['succeeded'] = success
                     sj_elem['job_start'] = job_start
                 else:
@@ -90,7 +92,6 @@ def get_jobs(fn, sj_benchmark_dir, s_logs, exec_mode):
 def get_success(fp, node):
     success = False
     job_start = None
-    print(fp)
     if op.isdir(fp) and node is not None:
         with open(op.join(fp, 'stderr'), 'r') as f:
             for line in f:

### 2019-05-15 paper.tex
Refactored introduction and cut through abstract

@@
 \begin{document}
-\title{Evaluation of dynamic resource allocation strategies for overlay Apache
-       Spark clusters on HPC \tristan{``Overlay'' and ``Apache Spark'' are redundant}}
+\title{Evaluation of pilot jobs for Apache Spark\\ applications on HPC clusters}
 \author{
     \IEEEauthorblockN{
-        Val\'erie Hayot-Sasson and Tristan Glatard
+        Val\'erie Hayot-Sasson and Tristan Glatard \\
+        Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada
     }
@@
 \begin{abstract}
-    Big Data has become prominent throughout many scientific fields and, as a
-    result, scientific communities have sought out Big Data frameworks to 
-    accelerate the processing of their increasingly data-intensive pipelines.
-    \TG{I'd write the previous sentence in continuous present tense, since I think
-    this is an on-going process.}
-    However, while scientific communities typically rely on High-Performance 
-    Computing (HPC) clusters for the parallelization of their pipelines, many 
-    popular Big Data frameworks such as Hadoop and Spark were primarily designed
-    to be executed on dedicated commodity infrastructures. Differences found between
-    these infrastructures and the policies that surround them limit the advantages
-    of deploying Big Data pipelines on HPC infrastructure. For instance, HPC
-    clusters typically employ a batch job submission system requiring users to
-    specify details on number of cores, nodes, amount of memory and duration.
-    Big Data engines, such as Spark, have no built-in mechanism to interact with
-    HPC batch job submission systems. Morever, the details required by the HPC
-    schedulers \TG{be more specific than ``details'': do you mean ``scheduling policies'',
-    ``application resource requirerments?''
-    } are not necessarily initially known to users of Big Data frameworks.
-    \TG{The previous paragraph is a bit long: it could be summarized to 2-3
-    sentences.}
-
-    Pilot scheduling strategies have been developed to address the limitations 
-    of traditional HPC batch job schedulers \TG{You could give examples of pilot schedulers as you did with Spark and Hadoop}.
-    Pilot schedulers decouple resource
-    provisioning from task scheduling, thereby enabling efficient resource
-    utilization through dynamic scheduling. This paper evaluates the benefits 
-    pilot-scheduling strategies over traditional batch submission
-    on HPC clusters with overlay Apache Spark clusters. We first evaluate the
-    overall speedup brought on by employing pilot-scheduling strategies. We then
-    examine the robustness to master and driver failures, which may be frequent 
-    when underestimating pilot walltime. Spark built-in checkpointing is also 
-    investigated in relation to pilot expiration.
-
-    Our results show that\ldots TBD.
+    Big Data has become prominent throughout many scientific fields and, as
+    a result, scientific communities have sought out Big Data frameworks to
+    accelerate the processing of their increasingly data-intensive
+    pipelines. However, while scientific communities typically rely on
+    High-Performance Computing (HPC) clusters for the parallelization of
+    their pipelines, many popular Big Data frameworks such as Hadoop and
+    Apache Spark were primarily designed to be executed on dedicated
+    commodity infrastructures. This paper evaluates the benefits of pilot
+    jobs over traditional batch submission on HPC clusters with overlay
+    Apache Spark clusters. We first evaluate the overall speedup brought on
+    by employing pilot jobs. We then examine the robustness to master and
+    driver failures, which may be frequent when underestimating pilot
+    walltime. Spark built-in checkpointing is also investigated in relation
+    to pilot expiration. Surprisingly, our results show that the speed-up
+    provided by pilot jobs over batch scheduling remains moderate (\TG{X} on
+    average) despite the presence of long queuing times. In addition, pilot
+    jobs provide an extra layer of scheduling that complexifies debugging
+    and deployment. We conclude that traditional batch scheduling should
+    remain the default strategies to deploy Apache Spark applications on
+    HPC clusters.
 \end{abstract}
@@
 
-With the increasing volume of data being collected and shared, Big Data has become
-an important area of focus to many scientific domains. As is the case with neuroimaging,
-datasets available publicly now reach up to Petabytes in size~\cite{UKBiobank, hcp}. Average workstations found in research labs would not be
-able to support the processing of such large datasets, leaving researchers to look 
-into cloud or High-Performance Computing~(HPC) solutions. Whereas both options are
-suitable for the processing of scientific Big Data, HPC clusters remain the more 
-sought after solution, potentially due to their cost-effectiveness.
+Pilot jobs, also known as dynamic resource provisioning or glide-in
+scheduling, are a widely-used technique to address infrastructure
+heterogeneity, variable task queuing times, fine task granularity, and node
+failures on distributed computing infrastructures~\cite{jakubsmodel}. Made popular by software
+projects such as Condor~\cite{thain2005distributed} and DIRAC~\cite{casajus2010dirac}, they
+became critical to grid computing, greatly improved the performance of HPC
+clusters, and enabled multi-cloud executions.
 
-Scientific workflows may be quite compute intensive, requiring more than a few
-hours to complete (\cite{freesurfer, fmriprep, a paper which mentions duration}).
-The coupling of large datasets with compute-intensive workflows may result in 
-unreasonably long processing times, which restricts the amount of data 
-utilized in scientific experiments. To reduce the effects of Big Data on processing
-times, researchers have begun to move towards employing Big Data frameworks for their
-analyses, such as Hadoop MapReduce~\cite{Hadoop},
-Apache Spark~\cite{Spark} and Dask~\cite{Dask}. However, these frameworks were 
-designed with dedicated commodity infrastructure in mind, and, with the exception
-of Dask, do not support traditional HPC schedulers.
+In contrast to static resource provisioning, pilot jobs are submitted to
+the infrastructure separately from the application, to provision resources
+on which application tasks will eventually be scheduled. Large pools of
+resources can thus be created, shielding applications from the underlying
+queuing times, failures, and other idiosyncracies. A variety of frameworks
+now rely on pilot jobs, including RADICAL-Pilot~\cite{merzky2015radical},
+and recent versions of the Pipeline System for Octave and Matlab
+(PSOM)~\cite{bellec2012pipeline}. The survey in~\cite{turilli2018comprehensive} reviews
+the current pilot-job systems.
 
-Within HPC clusters, many users share resources to process a variety of different
-applications. These shared resources include storage (e.g. parallel file system) and 
-compute node resources (memory, CPU, local disk space, etc.). In order for users to 
-gain access to shared compute resources, HPC clusters typically employ schedulers for 
-batch resource provisioning. Examples of such schedulers include PBS, SGE, Slurm and 
-TORQUE~\cite{schedulers}. These schedulers require that the user specify the amount
-of resources required and the duration that they require the resources for. The users 
-are then placed in a queue at a position based on the user's priority and resource 
-availability.
+% Goal of the paper
+In this paper, we study the use of pilot jobs to deploy Apache
+Spark~\cite{zaharia2016apache} applications on shared HPC clusters. The method
+currently recommended for this purpose involves batch requesting all the
+necessary resources and launching a standalone Spark cluster once the
+resources have been allocated. This is for instance the method used in
+Compute Canada (see details \href{https://docs.computecanada.ca/wiki/Apache_Spark/en}{here}),
+ our national computing infrastructure. With pilot
+jobs, rather than requesting all the resources at once, a Spark cluster is
+launched with a subset of the resources and is expanded as more resources
+get allocated. Consistently with previous examples of pilot job
+deployments, we hypothesize that this strategy would reduce queuing times
+by (1) fragmenting resource requirements, and (2) allowing shorter walltime
+estimates. While there have been some efforts on implementing pilot jobs
+for Apache Spark~\cite{jha-spark-pbs}, research is limited and
+none of the them detail their quantitative effect.
 
-There are several limitations to running applications on batch HPC schedulers, 
-particularly in the face of Big Data. For instance, users may not necessarily 
-know the exact requirements of the applications. With applications that are 
-both data- and compute-intensive, this may result in the user overestimating the
-resources required and the duration that these resources are required for. Frequent
-high overestimates may create bottlenecks in the system, preventing other users
-from occupying allocated idle resources. In contrast, users may underestimate required 
-resources, leading the batch scheduler to kill applications due to exceeded resource quotas.
-This may subsequently result in many failures within the application leading to
-additional re-executions of the same application. This is problematic for the user
-as their priority may be affected by the cluster usage. Additionally, the user may 
-require more time on the cluster to relaunch non-checkpointed and interrupted 
-tasks. \uwave{Furthermore, Big Data frameworks, such as Apache Spark, do not interact naturally
-with HPC schedulers and typically rely on other schedulers for resource provisioning
-and task scheduling. Therefore, to run Big Data applications on HPC schedulers, 
-it is also necessary to start an overlay cluster.} \tristan{The previous 2 sentences are a bit unclear. You should explain
-what an overlay cluster means here. ``Interact with other schedulers'' is vague. This point should end up explaining
-what static resource provisioning means for Spark, i.e., the Compute Canada approach. It might even be a separate paragraph.
-You could merge here the paragraph that currently explains that later.}
+Apache Spark is a popular Big Data framework, commonly used in both
+industrial and academic settings. Although it is a Scala-based framework,
+it also has APIs for Java, Python (PySpark) and R. Spark's Resilient
+Distributed Dataset (RDD) abstraction enabled in-memory processing of
+pipelines by co-locating tasks and data, which provided important
+performance improvements compared to its predecessor Hadoop
+MapReduce~\cite{dean2008mapreduce}. Through the use of RDDs, it also became possible
+to execute iterative workflows -- something not easily doable in older
+frameworks. Schedulers for Spark include its built-in standalone schedule,
+Yet Another Resource Negotiator (YARN~\cite{apache13yet}), and Mesos~\cite{hindman2011mesos}. As a
+result of the sustained increase in data volumes, Apache Spark and other
+Big Data engines are increasingly used for scientific applications,
+including in neuroimaging, our primary field of
+interest~\cite{boubela2016big,mehta2017comparative,maybethesimulationone,freeman2014mapping}.
 
-In contrast to static resource provisioning, as is the case with batch HPC schedulers, 
-dynamic resource provisioning, known as pilot job or glide-in scheduling, exists. 
-Dynamic resource provisioning allows the user to specify a minimum and maximum 
-amount of resources, and the cluster can expand and contract depending on application requirements.
-\tristan{Add one sentence to define what's a pilot}. 
-Each pilot may consist of the minimal amount of resources required to launch the 
-application. As each pilot would request allocation of fewer resources than that of a batch
-request, Pilot Jobs would typically spend less time on the resource allocation 
-queue. Furthermore, although individual pilots may expire due to walltimes, the dynamic
-nature of pilots allows the entire application to surpass walltime limitations, as
-additional pilots may be added to the pilot queue until application completion. 
-Examples of HPC schedulers employing pilot strategies include HTCondor~\cite{htcondor}
-and DIRAC~\cite{DIRAC}. \todo{maybe mention PSOM and Pegasus here \tristan{Yes for PSOM as it's neuroimaging. Pegasus
-only uses pilots through condor afaik.}}
+Big Data frameworks were designed with dedicated commodity infrastructure
+in mind, and, with the exception of Dask~\cite{rocklin2015dask}, do not support batch
+HPC schedulers such as PBS, SGE, Slurm and TORQUE~\cite{schedulers}, which
+are commonly available to scientists. Therefore, to run Big Data
+applications on HPC schedulers, it is also necessary to start an overlay
+cluster that will schedule application tasks on resources provisioned
+through batch schedulers. Our experiments quantify the effect of pilot jobs
+when combined with such an overlay cluster. Our evaluation will be broken
+down into two parts:
+1) the added value of pilots, and 2), tolerance to failures.
 
-\tristan{I think this could be the 3rd paragraph.} Apache Spark is a popular Big Data framework, commonly used in both industrial
-and academic settings. Although it is a Scala-based framework, it also has APIs
-for Java, Python (PySpark) and R. Spark's Resilient Distributed Dataset (RDD) abstraction enabled
-in-memory processing of pipelines by co-locating tasks and data, which provided important performance improvements compared to its predecessors.
- Through the use of RDDs, it also became possible to execute iterative 
-workflows -- something not easily doable in older frameworks such as Hadoop MapReduce.
-Schedulers for Spark include its built-in standalone schedule, Yet Another Resource Negotiator (YARN),
-and Mesos~\cite{yarn, mesos}. \tristan{You could cite Ariel's and your paper to say that Spark is sometimes useful in 
-neuroimaging on a cluster.}
-
-Projects such as Thunder~\cite{thunder}, a PySpark-compatible library 
-of image processing tools, have been developed in various fields 
-although they are still not widely adopted, possibly due to the 
-difficulties in adapting Big Data frameworks on HPC \todo{better 
-explain the difficulties...setting walltimes etc. also explain 
-neuroimaging 
-applications}. In addition, Spark has been used in 
-various research projects to process neuroimaging data \cite{Boubela, 
-ariel's paper, maybe the simulation one}. \tristan{in this paragraph you could refer
-to your pre-print for more information.} \tristan{I would summarize that in 1 or 2 sentences at the end of the previous
-paragraph.}
-
-The recommended method for launching Spark applications on HPC schedulers 
-involves batch requesting all the necessary resources and launching a standalone
-Spark cluster once the resources have been requested \TG{Add link to Compute Canada's instructions?}. This could significantly increase
-the scheduling time of Spark-based applications as a large amount of resources 
-may be necessary to process Big Data. In a pilot-scheduling model, rather than requesting all the resources 
-at once, a Spark cluster is launched with a subset of the resources and is expanded as 
-more resources get allocated. This, in turn, may reduce the overall processing of
-an application as less resources are requested at once and are therefore more likely
-to be scheduled faster.
-
-Pilot scheduling techniques have already been applied to scientific workflow engines, such as
-Pegasus, PSOM and RADICAL-Pilot~\cite{}. While there have been some efforts on
-implementing pilot scheduling approaches for Apache Spark\cite{jha and spark on pbs paper},
-research is limited and none of the them detail the effects of an overlay cluster
-on overall performance and robustness. 
-\tristan{This paragraph should write the scientific question of your paper more clearly: is it useful to 
-use a pilot job framework to run Spark workflows on HPC? Also, this comes a bit late in the intro, it would be good
-to precise the question early on, and the explain it in length as you do.}
-
-In this paper, we will analyse the effects of pilot scheduling strategies with an 
-overlay Spark cluster. Our evaluation will be broken down into two parts: 1)
-the added value of pilots, and 2), tolerance to failures. For determining the 
-added value of pilots, we will investigate the effect of queuing times on 
-pilot applications as compared to a batch requests. As much of the application's
-tolerance to failures are features brought on by the overlay Spark cluster, in
-Part 2 we will introduce failures that would be typical to pilot scheduling 
-approaches (e.g. node failures due to walltime expiration). This part will be 
-broken down into master failures, driver failures and checkpointing. As pilots 
-may need to be replenished during execution, we will also consider the effects of 
-job queues \todo{reword}.
-
-\tristan{Your text now mentions points 3 and 4 below, but doesn't talk about 1 and 2 and the fact that
-an overlay cluster already deals with that. Something that wasn't mentioned in my initial list was that pilot jobs
-also reduce the number of tasks seen by the cluster (this is also addressed by an overlay scheduler). It would be interesting
-to explain that many of the interests of pilot jobs are already addressed by overlay scheduling, hence the question whether
-pilot jobs are useful at all in this context.}
-\TG{The introduction is still very much geared toward the presentation of 
-the SPA software. Instead, I think it should introduce the question whether pilot jobs
-are useful at all to overlay schedulers, explain why they may be, and why they may not (there is already an overlay scheduler,
-and resource heterogeneity is limited). You could explain that the traditional benefits of pilot-job systems are:
+To summarize, our paper
+makes the following contributions:
 \begin{itemize}
-    \item Address worker node heterogeneity
-    \item Address worker node failures (software, incl walltime expirations, or hardware)
-    \item Address variable queuing times
-    \item Tolerant to resource requirement overestimation, in particular walltime and CPU
-    \item Increase workload distribution (use more nodes)
+\item We present a lightweight pilot-job framework to run Apache Spark
+applications on HPC clusters.
+\item We compare the makespan of a typical neuroimaging Big Data
+application with and without pilot jobs on two different HPC clusters and
+in different conditions.
+\item We describe a simple performance model to validate that the observed
+differences come from variations in queuing times rather than other
+factors.
 \end{itemize}
-Among these, (1) and to some extent (2) are already addressed by an overlay scheduler.
-
-The title should also reflect this.
-}
+Our methods, including infrastructure, job templates, application and
+performance model are described in Section~\ref{sec:methods}.
+Section~\ref{sec:results} presents our results which are discussed in
+Section~\ref{sec:discussion}. Section~\ref{sec:conclusion} concludes on the 
+relevant of pilot jobs for Apache Spark applications on HPC clusters.
 
@@
     of a scientific Spark application. For this, we created a synthetic application 
-    that would process the Big Brain~\cite{bigBrain}, a 76~G 3D histological
+    that would process the Big Brain~\cite{amunts2013bigbrain}, a 76~G 3D histological
     image of a human brain. The algorithm is a chain of map transformations that
@@
         \todo{incrementation with varying task durations}
-\section{Results}\label{sec:Results}
+\section{Results}\label{sec:results}
 \section{Discussion}\label{sec:discussion}
@@
 
+\section*{Acknowledgement}
+
+We warmly thank Compute Canada and its regional centers WestGrid and Calcul
+Qu\'ebec for providing the infrastructure used in our experiments.
+
+\bibliographystyle{IEEEtran} 
+\bibliography{biblio}
 \end{document}

### 2019-05-15 experiments/code/organize_logs.py
fixed organize logs

@@ -43,6 +43,9 @@ with open(json_f, 'r') as f:
                 for line in f:
                     if "'submissionId': '" in line:
                         driver_name = line.split("'submissionId': '")[1].split("', ")[0]
+                    elif "GET /v1/submissions/status/" in line:
+                        driver_name = line.split("GET /v1/submissions/status/")[1].split(' ')[0]
+
             if driver_name is not None:
                 print(driver_name)
                 try:

### 2019-05-16 paper.tex
added part of results and code to streamline generation of figures

--- 2019-05-08 paper.tex
@@
 \begin{document}
-\title{Evaluation of dynamic resource allocation strategies for overlay Apache
-       Spark clusters on HPC \tristan{``Overlay'' and ``Apache Spark'' are redundant}}
+\title{Evaluation of pilot scheduling strategies for Apache
+Spark clusters on HPC}
 \author{
@@
 \begin{abstract}
-    Big Data has become prominent throughout many scientific fields and, as a
-    result, scientific communities have sought out Big Data frameworks to 
+    Big Data is becoming prominent throughout many scientific fields and, as a
+    result, scientific communities are seeking Big Data frameworks to 
     accelerate the processing of their increasingly data-intensive pipelines.
-    \TG{I'd write the previous sentence in continuous present tense, since I think
-    this is an on-going process.}
     However, while scientific communities typically rely on High-Performance 
@@
     popular Big Data frameworks such as Hadoop and Spark were primarily designed
-    to be executed on dedicated commodity infrastructures. Differences found between
-    these infrastructures and the policies that surround them limit the advantages
-    of deploying Big Data pipelines on HPC infrastructure. For instance, HPC
-    clusters typically employ a batch job submission system requiring users to
-    specify details on number of cores, nodes, amount of memory and duration.
-    Big Data engines, such as Spark, have no built-in mechanism to interact with
-    HPC batch job submission systems. Morever, the details required by the HPC
-    schedulers \TG{be more specific than ``details'': do you mean ``scheduling policies'',
-    ``application resource requirerments?''
-    } are not necessarily initially known to users of Big Data frameworks.
-    \TG{The previous paragraph is a bit long: it could be summarized to 2-3
-    sentences.}
+    to be executed on dedicated commodity infrastructures. As Big Data frameworks
+    cannot leverage HPC schedulers directly, they must be executed on an overlay
+    cluster atop HPC allocations. This is problematic as application resource
+    requirements needed by the HPC scheduler may not be known by the user. 
+    Pilot scheduling strategies have been developed to address the limitations 
+    of traditional HPC batch job schedulers. Pilot schedulers, such as HTCondor and DIRAC, 
+    decouple resource
+    provisioning from task scheduling, thereby enabling efficient resource
+    utilisation through dynamic scheduling. This paper evaluates the benefits 
+    pilot-scheduling strategies over traditional batch submission
+    on HPC clusters with overlay Apache Spark clusters. We evaluate the
+    overall speedup brought on by employing pilot-scheduling strategies through
+    the application of four increasing resource configurations. Overall, we find
+    that there is little benefit to using pilot scheduling strategies, though it
+    can bring 2x when system queuing times are very slow. However, these
+    occurrences are rare. Generally pilots have approximately the same makespan
+    as batch. Despite makespan differences being found to be mostly due queueing 
+    times, pilots did not appear to have any advantage in this regard, potentially
+    due to system scheduling policies. Regardless, pilots may still be useful 
+    when application walltimes are underestimated. This remains to be investigated.
 
-    Pilot scheduling strategies have been developed to address the limitations 
-    of traditional HPC batch job schedulers \TG{You could give examples of pilot schedulers as you did with Spark and Hadoop}.
-    Pilot schedulers decouple resource
-    provisioning from task scheduling, thereby enabling efficient resource
-    utilization through dynamic scheduling. This paper evaluates the benefits 
-    pilot-scheduling strategies over traditional batch submission
-    on HPC clusters with overlay Apache Spark clusters. We first evaluate the
-    overall speedup brought on by employing pilot-scheduling strategies. We then
-    examine the robustness to master and driver failures, which may be frequent 
-    when underestimating pilot walltime. Spark built-in checkpointing is also 
-    investigated in relation to pilot expiration.
-
-    Our results show that\ldots TBD.
+   % 
+    %We then examine the robustness to master and driver failures, which may be frequent 
+    %when underestimating pilot walltime. Spark built-in checkpointing is also 
+    %investigated in relation to pilot expiration.
 \end{abstract}
@@
     queuing times. It is computed as follows:
-    $$
-    W = \frac{1}{M}\int_0^M{w(t)dt},
-    $$
+    \begin{equation}
+        W = \frac{1}{M}\int_0^M{w(t)dt}\label{eq:avgw}
+    \end{equation}
     where $w(t)$ is the number of workers available at time $t$. When the
@@
 
+    % maybe remove
     \subsection{Pipeline robustness}
@@
         \todo{scala standalone vs our pyspark workaround. kill masters in experiments}
-    \subsection{Checkpointing}
+    % maybe remove\subsection{Checkpointing}
         \todo{metric for determining how often to checkpoint based on cluster size}
-    \subsection{Job arrays}
+    % maybe remove \subsection{Job arrays}
         \todo{need to kill idle workers. may not want all workers to be running at once.}
-    \subsection{Example application}
+    % maybe remove \subsection{Example application}
         \todo{incrementation with varying task durations}
 \section{Results}\label{sec:Results}
+
+As can be seen in both figures~\ref{fig:makespansbeluga} and ~\ref{fig:makespanscedar}, pilots 
+generally did not bring any performance improvement to the execution of Spark workloads
+on Beluga and Cedar. In Beluga (Figure~\ref{fig:beluga1}), 16 pilots were always 
+slower than batch except for repetitions 4 and 12. The maximum speedup achieved
+was approximately 1.3~x for 8 pilots and 1.1~x for 16 pilots. In Figures~\ref{fig:beluga2}
+and ~\ref{fig:beluga3}, it is also the case that batch is generally faster than
+pilots for the majority of the cases, except a couple, in which the speedup brought
+by use of pilots is very limited. It is only in Configuration~4 (Figure~\ref{fig:beluga4})
+that we see a maximum speedup of 2.8, however, it generally remains that pilots are slower.
+
+The Cedar cluster (Figure~\ref{fig:makespanscedar}) also follows the same trend that
+the makespan of batch scheduled applications are generally faster than that of pilots,
+with a maximum speedup of 3, found both in Configuration~1 and 4. Moreover, in 
+Cedar, it was less clear whether to use 8 or 16 pilots as speedup between the two
+varies between the iterations. In general, both pilot configurations performed the
+same.
+
+In order to determine whether queueing time was the cause of the difference 
+in makespans between batch and pilots, we compared makespan with queuing time of 
+first pilot. As can be seen in Figure~\ref{fig:mqcedar}, variations in overall
+makespans is primiarily due to queueing time.
+
+In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
+are used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
+given an experiment configuration, where the line corresponds to the estimated makespan using 
+our model. Makespan for the model was calculated as:
+$$
+M = (\lceil 125/W \rceil\times d) + 20,
+$$
+where:
+\begin{itemize}
+    \item $M$ is the makespan of the application
+    \item $W$ is the average number of Spark workers throughout the execution
+    \item $d$ is the task delay associated with a given configuration
+    \item $20$ is the average measured incrementation duration for a BigBrain chunk
+\end{itemize}
+As can be seen in the figure, although pilots and batch had similar makespans
+for a given number of average workers, pilots were consistently slower than batch
+for the same number of workers. Both batch and pilots were consistent with the 
+model, however. In general, both pilots and batch had the same number of average
+workers.
+
+    \begin{figure*}
+        \centering
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_1_beluga}
+            \caption[]%
+            {{\small Configuration 1}}
+            \label{fig:beluga1}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_2_beluga}
+            \caption[]%
+            {{\small Configuration 2}}
+            \label{fig:beluga2}
+        \end{subfigure}
+        \vskip\baselineskip
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_3_beluga}
+            \caption[]%
+            {{\small Configuration 3}}
+            \label{fig:beluga3}
+        \end{subfigure}
+        \quad
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_4_beluga}
+            \caption[]%
+            {{\small Configuration 4}}
+            \label{fig:beluga4}
+        \end{subfigure}
+        \caption[]
+        {\small The application makespan on the Beluga cluster for all repetitions.}
+        \label{fig:makespansbeluga}
+    \end{figure*}
+
+    \begin{figure*}
+        \centering
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_1_cedar}
+            \caption[]%
+            {{\small Configuration 1}}
+            \label{fig:cedar1}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_2_cedar}
+            \caption[]%
+            {{\small Configuration 2}}
+            \label{fig:cedar2}
+        \end{subfigure}
+        \vskip\baselineskip
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_3_cedar}
+            \caption[]%
+            {{\small Configuration 3}}
+            \label{fig:cedar3}
+        \end{subfigure}
+        \quad
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_4_cedar}
+            \caption[]%
+            {{\small Configuration 4}}
+            \label{fig:cedar4}
+        \end{subfigure}
+        \caption[]
+        {\small The application makespan on the Cedar cluster for all repetitions.}
+        \label{fig:makespanscedar}
+    \end{figure*}
+
+      \begin{figure*}
+        \centering
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_4_cedar}
+            \caption[]%
+            {{\small Makespan}}
+            \label{fig:makespancedar}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/pq_4_cedar}
+            \caption[]%
+            {{\small Queuing time}}
+            \label{fig:queuecedar}
+        \end{subfigure}
+        \caption[]
+        {\small Side-by-side comparison of total makespan and queueuing time of
+        first pilots or batch application for Configuration 4 running on Cedar}
+        \label{fig:mqcedar}
+    \end{figure*}
+
+    \begin{figure*}
+        \centering
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/mw_beluga}
+            \caption[]%
+            {{\small Beluga}}
+            \label{fig:mwbeluga}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/mw_cedar}
+            \caption[]%
+            {{\small Cedar}}
+            \label{fig:mwcedar}
+        \end{subfigure}
+        \caption[]
+        {\small The relation between makespan and average number of workers as 
+        calculated using Equation~\ref{eq:avgw}. Trendline denotes the expected makespan
+        given the average number of workers given a certain configuration}
+        \label{fig:mwall}
+    \end{figure*}
+
 \section{Discussion}\label{sec:discussion}
 
-SPA vs other frameworks.
+%SPA vs other frameworks.
 
-Technical overhead of using pilots + an overlay cluster over a cluster.
+%Technical overhead of using pilots + an overlay cluster over a cluster.
+Pilots do not appear to provide any performance advantage over batch scheduling.
+This can be due to a few reasons. The queuing time for batch and pilots were similar, 
+which can be an indication that the required resources to run the batch requests were
+available, and our priority on the cluster was what affected the queuing time. This may, in turn,
+explain why batch was sometimes significantly slower than pilots -- queueing time was not
+only controlled by priority, but also by available resources.
+
+It was also found that having 16 pilots was generally slower than 8 for the same
+configuration. This was particularly the case on Beluga. We suspect this might be
+related to the number of Slurm jobs permitted for a user on the backfill queue. For
+all configurations, we ran batch, 8 pilots and 16 pilots concurrently. This would have
+resulted in 25 Slurm jobs trying the access the queue concurrently. On Beluga, this
+feature is configured to a total of 10 jobs. For 16 nodes, this would at best be
+only 62.5\% of the pilots being scheduled at a given time if all resources are available.
+The next 6 jobs would then have to wait 3 minutes before having an opportunity to enter the
+backfill. Cedar, on the other hand, permits 40 user jobs to be backfilled at a given time.
+Therefore, on Cedar, all concurrent Slurm jobs had an opportunity to be backfilled at
+the same time, which may explain why this behaviour is more apparent on B\'eluga. Furthermore,
+while both pilot configurations could be places in B\'eluga's low memory nodes, batch requests
+had to be place on B\'eluga's medium memory nodes. B\'eluga has 172 low memory nodes and
+516 mid memory nodes. As pilots would get priority for the low memory nodes first and 
+low memory nodes are less frequent that medium memory nodes, it is possible that this
+may have increased the overall queuing times of the pilots. However, Cedar's basic
+node has enough memory to for all batch and pilot configuration, therefore, this could not
+have affected queuing time for Cedar.
+
+Although queuing time is largely responsible for makespan variations, it is not
+always entirely responsible for the difference. Sometimes there are errors related
+to workers not registering properly with the master. These occurences lead to the 
+pipeline being processed entirely with a smaller number of available workers. As the task
+delay added considers the maximum amount of parallel workers, functioning with less total workers
+will significantly affect makespan. However, as seen in Figure~\ref{fig:mwall}
+pilots are slower even with the same amount of average workers. A potential reasoning for this
+may be the, worker registration delay and the time required to transfer data
+over after the application has already started. In batch scheduling, all workers a started in
+parallel and the data is transferred automatically at the beginning of the pipeline as the maximum amount
+of parallelism is available at driver start time. Pilots, on the other hand, may start workers in parallel,
+but not necessarily all will be started at the same time. This means that the entire 
+workflow may suffer the impacts of starting workers in sequence. Moreover, there would also
+be a delay with respect to transferring data from workers to newly added workers. Therefore,
+it is only natural that pilots would have a bit more overhead that batch given the same
+queuing time should pilots not all commence at once. Furthermore, the pilot scripts also
+have some starter overhead as the all start a master, a worker (registered to the main
+master) and attempt to start a driver. Batch need only start a single master and attempts to start
+a driver only once.
+
+The motivation to use pilots, however, is dynamic scheduling on HPC clusters. While we
+investigated the queuing time differences with respect to differences in available resources,
+the walltimes were kept static. Had walltime been underestimated to the point where the Slurm jobs
+would terminate prior to application completion, pilots would have been preferable to batch
+as batch scheduling has no mechanism to restart. Even if such a mechanism was in place, the entire
+cluster would be shutdown and need to restart, which has additional overheads. With pilots, as long as 
+there was at least one pilot alive, it would be possible to maintain the cluster and add additional pilots.
+Spark provides built-it fault-tolerance for not only the workers, but the master and driver as well.
 

### 2019-05-16 paper.tex

--- 2019-05-15 paper.tex
@@
 \begin{document}
+
 \title{Evaluation of pilot jobs for Apache Spark\\ applications on HPC clusters}
@@
     HPC clusters.
+
 \end{abstract}
@@
         \todo{scala standalone vs our pyspark workaround. kill masters in experiments}
-    \subsection{Checkpointing}
+    % maybe remove\subsection{Checkpointing}
         \todo{metric for determining how often to checkpoint based on cluster size}
-    \subsection{Job arrays}
+    % maybe remove \subsection{Job arrays}
         \todo{need to kill idle workers. may not want all workers to be running at once.}
-    \subsection{Example application}
+    % maybe remove \subsection{Example application}
         \todo{incrementation with varying task durations}
+
 \section{Results}\label{sec:results}
+
+As can be seen in both figures~\ref{fig:makespansbeluga} and ~\ref{fig:makespanscedar}, pilots 
+generally did not bring any performance improvement to the execution of Spark workloads
+on Beluga and Cedar. In Beluga (Figure~\ref{fig:beluga1}), 16 pilots were always 
+slower than batch except for repetitions 4 and 12. The maximum speedup achieved
+was approximately 1.3~x for 8 pilots and 1.1~x for 16 pilots. In Figures~\ref{fig:beluga2}
+and ~\ref{fig:beluga3}, it is also the case that batch is generally faster than
+pilots for the majority of the cases, except a couple, in which the speedup brought
+by use of pilots is very limited. It is only in Configuration~4 (Figure~\ref{fig:beluga4})
+that we see a maximum speedup of 2.8, however, it generally remains that pilots are slower.
+
+The Cedar cluster (Figure~\ref{fig:makespanscedar}) also follows the same trend that
+the makespan of batch scheduled applications are generally faster than that of pilots,
+with a maximum speedup of 3, found both in Configuration~1 and 4. Moreover, in 
+Cedar, it was less clear whether to use 8 or 16 pilots as speedup between the two
+varies between the iterations. In general, both pilot configurations performed the
+same.
+
+In order to determine whether queueing time was the cause of the difference 
+in makespans between batch and pilots, we compared makespan with queuing time of 
+first pilot. As can be seen in Figure~\ref{fig:mqcedar}, variations in overall
+makespans is primiarily due to queueing time.
+
+In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
+are used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
+given an experiment configuration, where the line corresponds to the estimated makespan using 
+our model. Makespan for the model was calculated as:
+$$
+M = (\lceil 125/W \rceil\times d) + 20,
+$$
+where:
+\begin{itemize}
+    \item $M$ is the makespan of the application
+    \item $W$ is the average number of Spark workers throughout the execution
+    \item $d$ is the task delay associated with a given configuration
+    \item $20$ is the average measured incrementation duration for a BigBrain chunk
+\end{itemize}
+As can be seen in the figure, although pilots and batch had similar makespans
+for a given number of average workers, pilots were consistently slower than batch
+for the same number of workers. Both batch and pilots were consistent with the 
+model, however. In general, both pilots and batch had the same number of average
+workers.
+
+    \begin{figure*}
+        \centering
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_1_beluga}
+            \caption[]%
+            {{\small Configuration 1}}
+            \label{fig:beluga1}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_2_beluga}
+            \caption[]%
+            {{\small Configuration 2}}
+            \label{fig:beluga2}
+        \end{subfigure}
+        \vskip\baselineskip
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_3_beluga}
+            \caption[]%
+            {{\small Configuration 3}}
+            \label{fig:beluga3}
+        \end{subfigure}
+        \quad
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_4_beluga}
+            \caption[]%
+            {{\small Configuration 4}}
+            \label{fig:beluga4}
+        \end{subfigure}
+        \caption[]
+        {\small The application makespan on the Beluga cluster for all repetitions.}
+        \label{fig:makespansbeluga}
+    \end{figure*}
+
+    \begin{figure*}
+        \centering
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_1_cedar}
+            \caption[]%
+            {{\small Configuration 1}}
+            \label{fig:cedar1}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_2_cedar}
+            \caption[]%
+            {{\small Configuration 2}}
+            \label{fig:cedar2}
+        \end{subfigure}
+        \vskip\baselineskip
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_3_cedar}
+            \caption[]%
+            {{\small Configuration 3}}
+            \label{fig:cedar3}
+        \end{subfigure}
+        \quad
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_4_cedar}
+            \caption[]%
+            {{\small Configuration 4}}
+            \label{fig:cedar4}
+        \end{subfigure}
+        \caption[]
+        {\small The application makespan on the Cedar cluster for all repetitions.}
+        \label{fig:makespanscedar}
+    \end{figure*}
+
+      \begin{figure*}
+        \centering
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/dedicated_4_cedar}
+            \caption[]%
+            {{\small Makespan}}
+            \label{fig:makespancedar}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/pq_4_cedar}
+            \caption[]%
+            {{\small Queuing time}}
+            \label{fig:queuecedar}
+        \end{subfigure}
+        \caption[]
+        {\small Side-by-side comparison of total makespan and queueuing time of
+        first pilots or batch application for Configuration 4 running on Cedar}
+        \label{fig:mqcedar}
+    \end{figure*}
+
+    \begin{figure*}
+        \centering
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/mw_beluga}
+            \caption[]%
+            {{\small Beluga}}
+            \label{fig:mwbeluga}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/mw_cedar}
+            \caption[]%
+            {{\small Cedar}}
+            \label{fig:mwcedar}
+        \end{subfigure}
+        \caption[]
+        {\small The relation between makespan and average number of workers as 
+        calculated using Equation~\ref{eq:avgw}. Trendline denotes the expected makespan
+        given the average number of workers given a certain configuration}
+        \label{fig:mwall}
+    \end{figure*}
+
 \section{Discussion}\label{sec:discussion}
 
-SPA vs other frameworks.
+%SPA vs other frameworks.
 
-Technical overhead of using pilots + an overlay cluster over a cluster.
+%Technical overhead of using pilots + an overlay cluster over a cluster.
+Pilots do not appear to provide any performance advantage over batch scheduling.
+This can be due to a few reasons. The queuing time for batch and pilots were similar, 
+which can be an indication that the required resources to run the batch requests were
+available, and our priority on the cluster was what affected the queuing time. This may, in turn,
+explain why batch was sometimes significantly slower than pilots -- queueing time was not
+only controlled by priority, but also by available resources.
+
+It was also found that having 16 pilots was generally slower than 8 for the same
+configuration. This was particularly the case on Beluga. We suspect this might be
+related to the number of Slurm jobs permitted for a user on the backfill queue. For
+all configurations, we ran batch, 8 pilots and 16 pilots concurrently. This would have
+resulted in 25 Slurm jobs trying the access the queue concurrently. On Beluga, this
+feature is configured to a total of 10 jobs. For 16 nodes, this would at best be
+only 62.5\% of the pilots being scheduled at a given time if all resources are available.
+The next 6 jobs would then have to wait 3 minutes before having an opportunity to enter the
+backfill. Cedar, on the other hand, permits 40 user jobs to be backfilled at a given time.
+Therefore, on Cedar, all concurrent Slurm jobs had an opportunity to be backfilled at
+the same time, which may explain why this behaviour is more apparent on B\'eluga. Furthermore,
+while both pilot configurations could be places in B\'eluga's low memory nodes, batch requests
+had to be place on B\'eluga's medium memory nodes. B\'eluga has 172 low memory nodes and
+516 mid memory nodes. As pilots would get priority for the low memory nodes first and 
+low memory nodes are less frequent that medium memory nodes, it is possible that this
+may have increased the overall queuing times of the pilots. However, Cedar's basic
+node has enough memory to for all batch and pilot configuration, therefore, this could not
+have affected queuing time for Cedar.
+
+Although queuing time is largely responsible for makespan variations, it is not
+always entirely responsible for the difference. Sometimes there are errors related
+to workers not registering properly with the master. These occurences lead to the 
+pipeline being processed entirely with a smaller number of available workers. As the task
+delay added considers the maximum amount of parallel workers, functioning with less total workers
+will significantly affect makespan. However, as seen in Figure~\ref{fig:mwall}
+pilots are slower even with the same amount of average workers. A potential reasoning for this
+may be the, worker registration delay and the time required to transfer data
+over after the application has already started. In batch scheduling, all workers a started in
+parallel and the data is transferred automatically at the beginning of the pipeline as the maximum amount
+of parallelism is available at driver start time. Pilots, on the other hand, may start workers in parallel,
+but not necessarily all will be started at the same time. This means that the entire 
+workflow may suffer the impacts of starting workers in sequence. Moreover, there would also
+be a delay with respect to transferring data from workers to newly added workers. Therefore,
+it is only natural that pilots would have a bit more overhead that batch given the same
+queuing time should pilots not all commence at once. Furthermore, the pilot scripts also
+have some starter overhead as the all start a master, a worker (registered to the main
+master) and attempt to start a driver. Batch need only start a single master and attempts to start
+a driver only once.
+
+The motivation to use pilots, however, is dynamic scheduling on HPC clusters. While we
+investigated the queuing time differences with respect to differences in available resources,
+the walltimes were kept static. Had walltime been underestimated to the point where the Slurm jobs
+would terminate prior to application completion, pilots would have been preferable to batch
+as batch scheduling has no mechanism to restart. Even if such a mechanism was in place, the entire
+cluster would be shutdown and need to restart, which has additional overheads. With pilots, as long as 
+there was at least one pilot alive, it would be possible to maintain the cluster and add additional pilots.
+Spark provides built-it fault-tolerance for not only the workers, but the master and driver as well.
 

### 2019-05-17 paper.tex

@@
 % Goal of the paper
-In this paper, we study the use of pilot jobs to deploy Apache
-Spark~\cite{zaharia2016apache} applications on shared HPC clusters. The method
-currently recommended for this purpose involves batch requesting all the
+In this paper, we study the use of pilot jobs to run scientific
+applications with Apache Spark~\cite{zaharia2016apache} on shared HPC
+clusters. The method currently recommended for this purpose involves batch
+requesting all the
 necessary resources and launching a standalone Spark cluster once the
@@
     analysis scripts are publicly available and can be found in our Spark
-    Pilot-job scheduler for HPC Applications (SPA) software at:
+    Pilot-job scheduler for HPC Applications (SPA) repository at:
     \href{https://github.com/big-data-lab-team/spa}{https://github.com/big-data-lab-team/spa}.
@@
     \subsection{Infrastructure}
-    All experiments were conducted on the Cedar and B\'eluga HPC computing clusters
-    made available by Compute Canada through WestGrid and Calcul Qu\'ebec \TG{add links}. Both 
-    clusters are accessible through the Slurm batch scheduler and Lustre parallel file system.
-    The Cedar cluster has a total of 1542 nodes
-    with a total of 58,416 CPU cores. Available memory on a Cedar node can range
-    from 125 to 3022~G. Standard nodes are equipped with either 2x Intel E5-2683 v4
-    Broadwell @ 2.1~Ghz (32 cores total) or 2x Intel Platinum 8160F Skylake @
-    2.1Ghz (48 cores total) CPUs and 
-    2 x 480~G SSD. All nodes and temporary storage on Cedar are connected by an 
+    All experiments were conducted on the Cedar and B\'eluga HPC computing
+    clusters made available by \href{https://www.computecanada.ca}{Compute
+    Canada} through \href{https://www.westgrid.ca}{WestGrid} and
+    \href{http://www.calculquebec.ca}{Calcul Qu\'ebec}. Both clusters are
+    accessible through the Slurm batch scheduler and Lustre parallel file
+    system~\cite{schwan2003lustre}. The Cedar cluster has a total of 1542
+    nodes with a total of 58,416 CPU cores. Available memory on a Cedar
+    node can range from 125 to 3022~G. Standard nodes are equipped with
+    either 2x Intel E5-2683 v4 Broadwell @ 2.1~Ghz (32 cores total) or 2x
+    Intel Platinum 8160F Skylake @ 2.1Ghz (48 cores total) CPUs and 2 x
+    480~G SSD. All nodes and temporary storage on Cedar are connected by an
     Intel OmniPath (version 1) with 100Gbit/s bandwidth.
 
-    B\'eluga, on the other hand, is a smaller cluster with 872 available nodes. 
-    Node memory can range between 92 to 752~G, with the most common node type having 
-    186G. All nodes contain 2 x Intel Gold 6148
-    Skylake @ 2.4~Ghz (40 cores/node) CPU and are connected to each other with a
-    100~Gb/s Mellanox Infiniband EDR network \TG{expand EDR}. Each non-GPU node type contains one 
-    480~G SSD.
+    B\'eluga, on the other hand, is a smaller cluster with 872 available
+    nodes. Node memory can range between 92 to 752~G, with the most common
+    node type having 186G. All nodes contain 2 x Intel Gold 6148 Skylake @
+    2.4~Ghz (40 cores/node) CPU and are connected to each other with a
+    100~Gb/s Mellanox Infiniband EDR network. Each non-GPU
+    node type contains one 480~G SSD.
+
+    \TG{A sub-section could be added here to describe the main features of Spark used:
+    client mode, cluster mode, master selection, supervise mode, etc}
 
@@
     inspired by the template provided by Compute Canada to launch Spark applications 
-    on Slurm \href{https://docs.computecanada.ca/wiki/Apache\_Spark/en}{available here}.
-    The template operates as follows: Certain resource requirements are requested by
+    on Slurm, \href{https://docs.computecanada.ca/wiki/Apache\_Spark/en}{available here}.
+    The template operates as follows: certain resource requirements are requested by
     the user (e.g. walltime, amount of memory per node, number of CPUs per task, number 
@@
     available memory on the node, to allow for offheap space. 
-    The Spark deploy-mode selected for the batch template is client-mode.
+    The Spark deploy-mode selected for the batch template is client-mode \TG{why?}.
 
@@
     but also allow us to make the driver fault-tolerant through the
-    \texttt{supervise} mode, which is only available in cluster deploy
-    \TG{A brief section on Spark to explain these modes and features would
-    be useful}. As with the masters, it is particularly important to have a
-    fault-tolerant driver in pilot-scheduling scenarios due to possible
-    walltime expiration \TG{It should be explained before that one of the interests of pilots
-    is to ``play'' with shorter walltimes}. \TG{It would also be nice to have runs where pilots 
-    have shorter walltimes, to see if queuing times are shorter too.}
+    \texttt{supervise} mode, which is only available in cluster deploy. As
+    with the masters, it is particularly important to have a fault-tolerant
+    driver in pilot-scheduling scenarios due to possible walltime
+    expiration.
     Should pilots be idle for a certain duration, the
     pilots will terminate themselves such as to not hog resources.
+    % \TG{It should be explained before that one of the interests
+    % of pilots is to ``play'' with shorter walltimes}. \TG{It would also be
+    % nice to have runs where pilots 
+    % have shorter walltimes, to see if queuing times are shorter too.}
 
@@
                         \State $chunk\gets chunk+1$                                              
-                        \State sleep $x$ \TG{Don't you sleep by x minus the duration of the previous step?}                                                        
+                        \State sleep $x$                               
                     \EndFor                                                                      
@@
     applications, we required a Spark application operating on a large dataset with
-    an important processing time to emulate a what would be the average requirements
+    an important processing time to emulate what would be the average requirements
     of a scientific Spark application. For this, we created a synthetic application 
@@
     \todo{walltime parameter of workflow}
-    \subsection{Model}
 
-    The makespan of an application can be defined as the total duration
+    \subsection{Performance model}
+
+    The makespan of an application is defined as the total duration
     between the submission time of the first application task, and the
-    completion time of the last application time. It includes any
+    completion time of the last application task. It includes any
     scheduling time, queueing time, data transfer time, and any other
@@
     Assuming a divisible load, i.e., the application can be divided in any
-    number of tasks, the makespan can be expressed using the following
+    number of tasks, the makespan can be written using the following
     expression, which holds for both batch and pilot execution modes:
@@
     \end{itemize}
-    The average number of workers $W$ allows us to obtain a more accurate makespan
-    figure as the average number of workers will decrease with increasing 
-    queuing times. It is computed as follows:
+    The average number of workers $W$ allows the model to take into account
+     variable queuing times. It is computed as follows:
     \begin{equation}
@@
     application is not subject to any scheduling or queuing time, the
-    average number of workers equal the number of workers requested.
+    average number of workers equals the number of workers requested.
 
@@
             \rowcolor{headcolor}                                                             
-            Configuration & RAM (GB) & Tasks \\                               
+            Configuration & RAM (GB) & Tasks \TG{Should be cores?}\\                               
             \hline                                                                           
@@
        not need to be readjusted for each configuration. Given utmost parallelism,
-       the sleep delay added would amount to 1 hour of processing time regardless
-       of configuration. The real processing time of the application, however, 
-       would differ between configurations as we did not account for task duration
-       in the sleep delay. However, task duration was not expected to signicantly impact
+       the sleep delay added would amount to a makespan of 1 hour regardless
+       of configuration. The actual processing time of the application, however, 
+       would differ between configurations as we did not account for task duration (incrementation time, line
+       8 in Algorithm~\ref{alg:incrementation})
+       in the sleep delay. However, task duration was not expected to significantly impact
        overall processing time, but nevertheless, we set application walltime to 
@@
        processing time that may be incurred in applications with lower levels of 
-       parallelism.
+       parallelism. \TG{this last sentence is a bit confusing}
 
@@
        different execution modes were launched. Furthermore, the order of the execution modes
-       for each configuration was randomized to ensure that SLURM job request
+       for each configuration was randomized to ensure that Slurm job request
        order could not have affected results.
@@
 
-As can be seen in both figures~\ref{fig:makespansbeluga} and ~\ref{fig:makespanscedar}, pilots 
+As can be seen in both Figures~\ref{fig:makespansbeluga} and~\ref{fig:makespanscedar}, pilots 
 generally did not bring any performance improvement to the execution of Spark workloads
-on Beluga and Cedar. In Beluga (Figure~\ref{fig:beluga1}), 16 pilots were always 
+on B\'eluga and Cedar. In B\'eluga (Figure~\ref{fig:beluga1}), 16 pilots were always 
 slower than batch except for repetitions 4 and 12. The maximum speedup achieved
@@
 by use of pilots is very limited. It is only in Configuration~4 (Figure~\ref{fig:beluga4})
-that we see a maximum speedup of 2.8, however, it generally remains that pilots are slower.
+that we see a maximum speedup of 2.8, however, it generally remains that pilots are slower. \TG{could you report the 
+average speed-up of pilots?}
 
@@
 \end{itemize}
+\TG{You should remove the $\lceil . \righceil$ as it is not part of the model and I'm not sure
+if it works with pilots. Also, if you're using numbers you should report their units (seconds here, for 20 and the makespan).
+It's unclear where 125 comes from. And why isn't it $\frac{125(d+20)}{W}$?}
 As can be seen in the figure, although pilots and batch had similar makespans
@@
 for the same number of workers. Both batch and pilots were consistent with the 
-model, however. In general, both pilots and batch had the same number of average
-workers.
+model, however \TG{there is a contradiction here: if both batch and pilots were consistent with the model, then
+they should have similar makespans.}. In general, both pilots and batch had the same number of average
+workers. \TG{You could also plot the difference in the number of workers, for each experiment.}
 
@@
 which can be an indication that the required resources to run the batch requests were
-available, and our priority on the cluster was what affected the queuing time. This may, in turn,
+available, and our priority on the cluster was what affected the queuing time \TG{I don't understand this, could you clarify?}. This may, in turn,
 explain why batch was sometimes significantly slower than pilots -- queueing time was not
@@
 It was also found that having 16 pilots was generally slower than 8 for the same
-configuration. This was particularly the case on Beluga. We suspect this might be
+configuration. This was particularly the case on B\'eluga. We suspect this might be
 related to the number of Slurm jobs permitted for a user on the backfill queue. For
 all configurations, we ran batch, 8 pilots and 16 pilots concurrently. This would have
-resulted in 25 Slurm jobs trying the access the queue concurrently. On Beluga, this
-feature is configured to a total of 10 jobs. For 16 nodes, this would at best be
-only 62.5\% of the pilots being scheduled at a given time if all resources are available.
+resulted in 25 Slurm jobs trying the access the queue concurrently. On
+B\'eluga, the max number of jobs in the backfill queue is configured to a
+total of 10 jobs. For 16 nodes, this would at best be only 62.5\% of the
+pilots being scheduled at a given time if all resources are available.
 The next 6 jobs would then have to wait 3 minutes before having an opportunity to enter the
@@
 the same time, which may explain why this behaviour is more apparent on B\'eluga. Furthermore,
-while both pilot configurations could be places in B\'eluga's low memory nodes, batch requests
+while both pilot configurations could be placed in B\'eluga's low memory nodes, batch requests
 had to be place on B\'eluga's medium memory nodes. B\'eluga has 172 low memory nodes and
@@
 always entirely responsible for the difference. Sometimes there are errors related
-to workers not registering properly with the master. These occurences lead to the 
-pipeline being processed entirely with a smaller number of available workers. As the task
+to workers not registering properly with the master. These occurrences lead to the 
+application being processed entirely with a smaller number of available workers. As the task
 delay added considers the maximum amount of parallel workers, functioning with less total workers
@@
 pilots are slower even with the same amount of average workers. A potential reasoning for this
-may be the, worker registration delay and the time required to transfer data
-over after the application has already started. In batch scheduling, all workers a started in
+may be the worker registration delay and the time required to transfer data
+over after the application has already started. In batch scheduling, all workers are started in
 parallel and the data is transferred automatically at the beginning of the pipeline as the maximum amount
@@
 but not necessarily all will be started at the same time. This means that the entire 
-workflow may suffer the impacts of starting workers in sequence. Moreover, there would also
+workflow may suffer the impacts of starting workers in sequence \TG{isn't it queuing time again?}. Moreover, there would also
 be a delay with respect to transferring data from workers to newly added workers. Therefore,
 it is only natural that pilots would have a bit more overhead that batch given the same
-queuing time should pilots not all commence at once. Furthermore, the pilot scripts also
-have some starter overhead as the all start a master, a worker (registered to the main
-master) and attempt to start a driver. Batch need only start a single master and attempts to start
+queuing time should pilots not all commence at once \TG{ok, but isn't the total amount of transfered data the same in both conditions?}.
+ Furthermore, the pilot scripts also
+have some starter overhead as the all \TG{?} start a master, a worker (registered to the main
+master) and attempt to start a driver. Batch needs only start a single master and attempts to start
 a driver only once.
@@
 the walltimes were kept static. Had walltime been underestimated to the point where the Slurm jobs
-would terminate prior to application completion, pilots would have been preferable to batch
+would terminate prior to application completion, pilots might have been preferable to batch
 as batch scheduling has no mechanism to restart. Even if such a mechanism was in place, the entire
@@
 Spark provides built-it fault-tolerance for not only the workers, but the master and driver as well.
+
+\TG{You could end the discussion by a comment on the current literature on pilot jobs. How are your results
+consistent or not with the literature?}
 

### 2019-05-17 paper.tex
updated figures and added text on spark

@@
     client mode, cluster mode, master selection, supervise mode, etc}
+
+    \section{Spark Configuration}
+    Spark has many configuration options available for dynamic resource provisioning.
+    Its standalone cluster provides two deployment modes: client and cluster. The client
+    deploy mode executes the driver within the \texttt{spark-submit} process as
+    a client to the cluster. The cluster deploy-mode, however, runs the driver within
+    a worker process. Such a configuration is practical when the driver program
+    is deployed on a machine not co-located in the cluster to reduce network 
+    latency between workers and driver. The cluster deploy mode also has the texttt{supervise}
+    feature which provides fault-tolerance to the driver. This is particularly useful for
+    pilot-based overlay clusters where the running driver may fail to walltime expiration.
+    Cluster deploy mode is not available for python applications using the standalone mode.
+
+    With HPC clusters, client mode is not necessarily possible on the login node as the
+    driver client may require more resources than permitted. Moreover, due to cluster
+    security policies, it may not be possible, or difficult to execute the driver from
+    an external computer, such as a personal laptop. Therefore, for both batch and pilot
+    scheduling, it is necessary to launch the driver from within the Slurm jobs.
+
+    Spark also provides master fault tolerance through single-node recovery, where a shared
+    filesystem is available. To enable this, a recovery folder must be set. Should the master
+    fail, a new master will be able to takeover based on the information available in the recovery
+    folder. Although we start a master on each pilot, we have not incorporated this recovery function
+    for our experiments. This feature is, however, important and should be incorporated in pilot schedulers
+    were any nodes can fail by design.
+
+    %%maybe add checkpointing
 
@@
 
-        \todo{scala standalone vs our pyspark workaround. kill masters in experiments}
     % maybe remove\subsection{Checkpointing}
-        \todo{metric for determining how often to checkpoint based on cluster size}
+    %    \todo{metric for determining how often to checkpoint based on cluster size}
     % maybe remove \subsection{Job arrays}
-        \todo{need to kill idle workers. may not want all workers to be running at once.}
+    %    \todo{need to kill idle workers. may not want all workers to be running at once.}
     % maybe remove \subsection{Example application}
-        \todo{incrementation with varying task durations}
+    %    \todo{incrementation with varying task durations}
 
@@
 average speed-up of pilots?}
-
-The Cedar cluster (Figure~\ref{fig:makespanscedar}) also follows the same trend that
-the makespan of batch scheduled applications are generally faster than that of pilots,
-with a maximum speedup of 3, found both in Configuration~1 and 4. Moreover, in 
-Cedar, it was less clear whether to use 8 or 16 pilots as speedup between the two
-varies between the iterations. In general, both pilot configurations performed the
-same.
-
-In order to determine whether queueing time was the cause of the difference 
-in makespans between batch and pilots, we compared makespan with queuing time of 
-first pilot. As can be seen in Figure~\ref{fig:mqcedar}, variations in overall
-makespans is primiarily due to queueing time.
-
-In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
-are used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
-given an experiment configuration, where the line corresponds to the estimated makespan using 
-our model. Makespan for the model was calculated as:
-$$
-M = (\lceil 125/W \rceil\times d) + 20,
-$$
-where:
-\begin{itemize}
-    \item $M$ is the makespan of the application
-    \item $W$ is the average number of Spark workers throughout the execution
-    \item $d$ is the task delay associated with a given configuration
-    \item $20$ is the average measured incrementation duration for a BigBrain chunk
-\end{itemize}
-\TG{You should remove the $\lceil . \righceil$ as it is not part of the model and I'm not sure
-if it works with pilots. Also, if you're using numbers you should report their units (seconds here, for 20 and the makespan).
-It's unclear where 125 comes from. And why isn't it $\frac{125(d+20)}{W}$?}
-As can be seen in the figure, although pilots and batch had similar makespans
-for a given number of average workers, pilots were consistently slower than batch
-for the same number of workers. Both batch and pilots were consistent with the 
-model, however \TG{there is a contradiction here: if both batch and pilots were consistent with the model, then
-they should have similar makespans.}. In general, both pilots and batch had the same number of average
-workers. \TG{You could also plot the difference in the number of workers, for each experiment.}
 
@@
 
+The Cedar cluster (Figure~\ref{fig:makespanscedar}) also follows the same trend that
+the makespan of batch scheduled applications are generally faster than that of pilots,
+with a maximum speedup of 3, found both in Configuration~1 and 4. Moreover, in 
+Cedar, it was less clear whether to use 8 or 16 pilots as speedup between the two
+varies between the iterations. In general, both pilot configurations performed the
+same.
+
     \begin{figure*}
@@
 
+In order to determine whether queueing time was the cause of the difference 
+in makespans between batch and pilots, we compared makespan with queuing time of 
+first pilot. As can be seen in Figure~\ref{fig:mqcedar}, variations in overall
+makespans is primiarily due to queueing time.
+
       \begin{figure*}
@@
     \end{figure*}
+
+In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
+are used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
+given an experiment configuration, where the line corresponds to the estimated makespan using 
+our model. Makespan for the model was calculated as:
+$$
+M = \frac{125\times(d + 20)}{W}\times 10,
+$$
+where:
+\begin{itemize}
+    \item $125$ is the number of chunks BigBrain was split into
+    \item $M$ is the makespan of the application
+    \item $W$ is the average number of Spark workers throughout the execution
+    \item $d$ is the task delay associated with a given configuration
+    \item $20$ is the average measured incrementation duration, in seconds, for a BigBrain chunk
+    \item $10$ is the number of iterations
+\end{itemize}
+As can be seen in the figure, although pilots and batch had similar makespans
+for a given number of average workers, pilots were consistently slower than batch
+for the same number of workers. Both batch and pilots were consistent with the 
+model, however \TG{there is a contradiction here: if both batch and pilots were consistent with the model, then
+they should have similar makespans.}. In general, both pilots and batch had the same number of average
+workers. \TG{You could also plot the difference in the number of workers, for each experiment.}
 
@@
 be a delay with respect to transferring data from workers to newly added workers. Therefore,
-it is only natural that pilots would have a bit more overhead that batch given the same
+it is only natural that pilots would have a bit more overhead than batch given the same
 queuing time should pilots not all commence at once \TG{ok, but isn't the total amount of transfered data the same in both conditions?}.
  Furthermore, the pilot scripts also
-have some starter overhead as the all \TG{?} start a master, a worker (registered to the main
+have some startup overhead as all pilots start a master, a worker (registered to the main
 master) and attempt to start a driver. Batch needs only start a single master and attempts to start

### 2019-05-18 paper.tex

@@
 \usepackage{algpseudocode}
+\usepackage{stmaryrd}
 \algnewcommand\algorithmicforeach{\textbf{for each}}
@@
         Val\'erie Hayot-Sasson and Tristan Glatard \\
-        Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada
+        Department of Computer Science and Software Engineering, Concordia University \\
+        Montr\'eal, Qu\'ebec, Canada
     }
@@
 jobs, rather than requesting all the resources at once, a Spark cluster is
-launched with a subset of the resources and is expanded as more resources
+launched with a subset of the resources, and is expanded as more resources
 get allocated. Consistently with previous examples of pilot job
@@
 \begin{itemize}
-\item We present a lightweight pilot-job framework to run Apache Spark
+\item We present SPA, a lightweight pilot-job framework to run Apache Spark
 applications on HPC clusters.
@@
 
-    \TG{A sub-section could be added here to describe the main features of Spark used:
-    client mode, cluster mode, master selection, supervise mode, etc}
+    It is important to note that these clusters were used in production,
+    that is, concurrently with other users. This allowed us to test the
+    added-value of pilot scheduling in different realistic conditions of queuing
+    times. At the time of our experiments, B\'eluga had recently been
+    commissioned, resulting in low usage and shorter queuing times overall.
+    Conversely, Cedar had been operating for a few years, resulting in
+    higher usage and longer queuing times. Our experiments were repeated 15
+    times in each configuration to capture queuing time variability.
 
-    \section{Spark Configuration}
+    \subsection{Spark Configuration}
+
+    \tristan{add a paragraph to explain master, worker, driver}
+
     Spark has many configuration options available for dynamic resource provisioning.
@@
     is deployed on a machine not co-located in the cluster to reduce network 
-    latency between workers and driver. The cluster deploy mode also has the texttt{supervise}
-    feature which provides fault-tolerance to the driver. This is particularly useful for
-    pilot-based overlay clusters where the running driver may fail to walltime expiration.
-    Cluster deploy mode is not available for python applications using the standalone mode.
+    latency between workers and driver. The cluster deploy mode also has the \texttt{supervise}
+    feature which provides fault-tolerance to the driver \tristan{, by means of checkpointing?}.
+    This is particularly useful for
+    pilot-based overlay clusters where the running driver may fail due to walltime expiration.
+    Cluster deploy mode is not available for Python applications using the standalone mode \tristan{standalone mode wasn't defined.
+    Rewrite positively saying that it's ony available in Scala?}.
 
@@
     an external computer, such as a personal laptop. Therefore, for both batch and pilot
-    scheduling, it is necessary to launch the driver from within the Slurm jobs.
+    scheduling, it is necessary to launch the driver from within the Slurm jobs, that is, in cluster mode.
 
@@
     for our experiments. This feature is, however, important and should be incorporated in pilot schedulers
-    were any nodes can fail by design.
+    were any nodes can fail by design. \tristan{Last two sentences should be moved to job templates.}
 
@@
     \subsection{Job Templates}
+
+    \tristan{This section could be renamed to ``SPA system design'', and you could add a paragraph explaining
+    that for the sake of the experiment, you reimplemented the simplest possible pilot system since existing ones (refs)
+    were either not found online, or difficult to use.}
 
@@
     each pilot will start its own Spark master and worker. Only one pilot,
-    the first pilot to attempt to launch the driver, will do so. The reason for 
+    the first pilot to attempt to launch the driver, will do so \tristan{will do what? launch the driver?}.
+    The reason for 
     which each pilot starts its own master is to ensure the fault tolerance of the 
@@
     % have shorter walltimes, to see if queuing times are shorter too.}
+    \TG{You could add a figure showing the driver, workers, master and stand-by masters, to make the system 
+    description clearer.}
 
@@
                 \EndInput                                                                    
-                \ForEach{$chunk \in C$}                                                      
+                \ForEach{$chunk$ in $C$}                                                      
                     \State read $chunk$                                        
-                    \For{$i \in [1, n]$}                                                         
+                    \For{$i \in \llbracket 1, n \rrbracket$}                                                         
                         \State $chunk\gets chunk+1$                                              
@@
     corresponds to an ideal case where no data or other overhead is
-    present: only queuing times are included.
+    present: only queuing times are included, through the integration in
+    Equation~\ref{eq:avgw}.
 
@@
 generally did not bring any performance improvement to the execution of Spark workloads
-on B\'eluga and Cedar. In B\'eluga (Figure~\ref{fig:beluga1}), 16 pilots were always 
+on B\'eluga and Cedar. In B\'eluga (Figure~\ref{fig:beluga1}) \tristan{why just in 1a and not 1b, ...?}, 16 pilots were always 
 slower than batch except for repetitions 4 and 12. The maximum speedup achieved
@@
 by use of pilots is very limited. It is only in Configuration~4 (Figure~\ref{fig:beluga4})
-that we see a maximum speedup of 2.8, however, it generally remains that pilots are slower. \TG{could you report the 
-average speed-up of pilots?}
+that we see a maximum speedup of 2.8, however, it generally remains that
+pilots are slightly slower. \TG{You should report the average speed-up of
+pilots: 8 pilots vs batch and 16 pilots vs batch.}
 
@@
 In order to determine whether queueing time was the cause of the difference 
-in makespans between batch and pilots, we compared makespan with queuing time of 
+in makespans between batch and pilots \tristan{rather than other overheads?}, we compared makespan with queuing time of 
 first pilot. As can be seen in Figure~\ref{fig:mqcedar}, variations in overall
-makespans is primiarily due to queueing time.
+makespans is primiarily due to queueing time. \tristan{I don't like these figures: if you want to show
+this, you should plot the makespan vs the queuing time of the first pilot, in an XY plot. But then the MW plot
+is better because it takes into account the queuing time of all pilots, not only the first one. I would remove this figure.}
 
@@
 given an experiment configuration, where the line corresponds to the estimated makespan using 
-our model. Makespan for the model was calculated as:
+our performance model. Makespan for the model was calculated as:
 $$
@@
     \item $125$ is the number of chunks BigBrain was split into
-    \item $M$ is the makespan of the application
-    \item $W$ is the average number of Spark workers throughout the execution
+    \item $M$ is the makespan of the application, in seconds
+    \item $W$ is the average number of Spark workers throughout the execution, computed as in Equation~\ref{eq:avgw}
     \item $d$ is the task delay associated with a given configuration
@@
 \end{itemize}
+\tristan{You should remove the ceil from the plot too.}
 As can be seen in the figure, although pilots and batch had similar makespans
@@
 
+\tristan{Comment on the difficulties introduce by adding another layer in an already complex system:
+HPC + Spark. It complexifies debugging.}
+
 \TG{You could end the discussion by a comment on the current literature on pilot jobs. How are your results

### 2019-05-19 paper.tex
added figures and text. need to complete discussion

@@
 
-    \tristan{add a paragraph to explain master, worker, driver}
+    There are three possible cluster managers available to use in Spark applications:
+    the Standalone cluster manager, YARN and Mesos. The Standalone cluster manager is the
+    most basic cluster manager available and is packaged directly with Spark. YARN is a
+    resource manager designed for the resource management of Hadoop applications. However,
+    it can be used with different types of applications as well. In contrast, Mesos was 
+    designed to manage the resources for a variety of applications. As such, it incorporates
+    strategies that are more efficient for use with different frameworks making it possible
+    to be used as an HPC scheduler. Due to Standalone's lightweight nature and our intent
+    to start Spark clusters per application, we focus on Standalone mode.
+
+    A Spark cluster is made up of three components: the cluster manager, workers and
+    driver. The cluster manager, also known as the Spark master in Standalone mode, is responsible for the
+    resource provisioning within the cluster. Unlike the workers and driver, it is not
+    application specific. The driver, which can start in the client where it is launched (i.e. client deploy mode)
+    or directly on a worker node (i.e. cluster deploy mode), must request resources from the master
+    to launch its tasks. The master the requests workers with the necessary resources available to start
+    executors. The driver can then communicate with the executors to schedule and launch its tasks. Both the driver and
+    the executors are application specific and connected directly to the master.
 
@@
     latency between workers and driver. The cluster deploy mode also has the \texttt{supervise}
-    feature which provides fault-tolerance to the driver \tristan{, by means of checkpointing?}.
+    feature which allows the driver to be restarted in case of failure.
     This is particularly useful for
     pilot-based overlay clusters where the running driver may fail due to walltime expiration.
-    Cluster deploy mode is not available for Python applications using the standalone mode \tristan{standalone mode wasn't defined.
-    Rewrite positively saying that it's ony available in Scala?}.
+    Using the Standalone cluster manager, client mode is supported in all APIs, whereas cluster mode is
+    only available for applications using the Scala, Java and R APIs.
 
@@
     fail, a new master will be able to takeover based on the information available in the recovery
-    folder. Although we start a master on each pilot, we have not incorporated this recovery function
-    for our experiments. This feature is, however, important and should be incorporated in pilot schedulers
-    were any nodes can fail by design. \tristan{Last two sentences should be moved to job templates.}
+    folder.
 
-    %%maybe add checkpointing
+    In long running applications, it is necessary to be able to recovery the application execution from
+    last successful state. Spark provides a checkpointing mechanism in order to ensure that the application
+    is fault-tolerant to any kind of non application-related failure, such as a system failure.
 
-    \subsection{Job Templates}
+    \subsection{SPA system design}
 
-    \tristan{This section could be renamed to ``SPA system design'', and you could add a paragraph explaining
-    that for the sake of the experiment, you reimplemented the simplest possible pilot system since existing ones (refs)
-    were either not found online, or difficult to use.}
+    There is not shortage of scripts available online for starting a Spark cluster on HPC~\footnote{\url{https://github.com/NIH-HPC/spark-slurm},}\footnote{\url{https://www.sherlock.stanford.edu/docs/software/using/spark/},}\footnote{\url{https://sparkhpc.readthedocs.io},}\footnote{\url{https://www.osc.edu/~troy/pbstools/man/pbs-spark-submit}}.
+    However, all these scripts start up a cluster within a single batch call. At most, there are two calls, one to startup the cluster,
+    and the second, to start the application execution. None of these use any kind of pilot-scheduling approach, nor do they discuss the
+    effects of queuing time and how they might want to adjust it in the case where the driver is started separately
+    from the batch job to start the cluster. Even in the case of RADICAL-Pilot, as per some available tutorials for using the pilot-job application with
+    Spark~\footnote{\url{https://github.com/radical-cybertools/pilot-streaming/blob/master/examples/Pilot-Streaming-GettingStarted.ipynb},}\footnote{\url{https://github.com/radical-cybertools/MIDAS-tutorial/blob/master/pilot/Pilot-Spark.ipynb}}, it appears that the Spark cluster is started within a single pilot and the application is submitted, in a separate process, to that pilot.
+    Due to the fact that existing solutions mainly rely on a single batch call to startup the Spark
+    cluster and do not appear to be able to submit multiple smaller jobs to improve queuing times, we have decided to implement our own solution.
 
@@
     available memory on the node, to allow for offheap space. 
-    The Spark deploy-mode selected for the batch template is client-mode \TG{why?}.
+    The Spark deploy mode selected for the batch template is client mode. We selected this
+    deploy mode over cluster mode as driver recovery would not be required in batch in the
+    case of walltime expiration, as all the allocated resources will expire at the same time.
+    Moreover, many scientific applications are written in Python, which cannot use cluster mode using
+    the Standalone resource manager. Should pilot scheduling using cluster mode significantly reduce
+    queuing time compared to batch scheduling using client mode, it would provide enough justification to extend
+    Spark's Standalone scheduler to include cluster mode for PySpark applications.
 
     The pilot submission template is similar to that of the batch template, however,
-    each pilot will start its own Spark master and worker. Only one pilot,
-    the first pilot to attempt to launch the driver, will do so \tristan{will do what? launch the driver?}.
+    each pilot will start its own Spark master and worker. However, there will only be one
+    pilot which will start the driver process. The pilot selected to start the driver is the first one
+    that attempts to do so by way of lockfile.
     The reason for 
@@
     description clearer.}
+
+    Although we start a master on each pilot, we have not incorporated the master recovery function
+    for our experiments. This feature is, however, important and should be incorporated in pilot schedulers
+    were any nodes can fail by design.
 
@@
 
-    \todo{walltime parameter of workflow}
-
     \subsection{Performance model}
@@
 
-    \todo{perhaps expand a bit here \TG{I have a formal proof for eq (1) but I don't think
-    it's required here.}}
-
     \subsection{Added value of pilot scheduling}
@@
             \centering                                                                       
-            \begin{tabular}{c|c|c}                                                             
+            \begin{tabular}{c|c|c|c}                                                             
             \rowcolor{headcolor}                                                             
-            Configuration & RAM (GB) & Tasks \TG{Should be cores?}\\                               
+            Configuration & RAM (GB) & Tasks & Cores per task\\                               
             \hline                                                                           
-            1 & 112 & 16 \\                                               
-            2 & 224 & 32 \\                                               
-            3 & 336 & 48 \\
-            4 & 448 & 64 \\
+            1 & 112 & 16 & 1\\                                               
+            2 & 224 & 32 & 1\\                                               
+            3 & 336 & 48 & 1\\
+            4 & 448 & 64 & 1\\
             \end{tabular}                                                                    
@@
        8 in Algorithm~\ref{alg:incrementation})
-       in the sleep delay. However, task duration was not expected to significantly impact
-       overall processing time, but nevertheless, we set application walltime to 
-       2 hours and 30 minutes for all configurations to account for additional
-       processing time that may be incurred in applications with lower levels of 
-       parallelism. \TG{this last sentence is a bit confusing}
+       in the sleep delay. Due to slight variability in task durations between experiments, in 
+       addition to variability in application parallelism due to queuing times of each pilot Slurm job, 
+       we have set the walltime to be 2h30 to ensure experiment completion.
 
@@
 generally did not bring any performance improvement to the execution of Spark workloads
-on B\'eluga and Cedar. In B\'eluga (Figure~\ref{fig:beluga1}) \tristan{why just in 1a and not 1b, ...?}, 16 pilots were always 
-slower than batch except for repetitions 4 and 12. The maximum speedup achieved
-was approximately 1.3~x for 8 pilots and 1.1~x for 16 pilots. In Figures~\ref{fig:beluga2}
-and ~\ref{fig:beluga3}, it is also the case that batch is generally faster than
-pilots for the majority of the cases, except a couple, in which the speedup brought
-by use of pilots is very limited. It is only in Configuration~4 (Figure~\ref{fig:beluga4})
-that we see a maximum speedup of 2.8, however, it generally remains that
-pilots are slightly slower. \TG{You should report the average speed-up of
-pilots: 8 pilots vs batch and 16 pilots vs batch.}
+on either system. In B\'eluga and Cedar, 16 pilots were generally slower than batch except for a 
+few occurrencs. When 16 pilots were faster, it was generally either very slight, or, as expected,
+when batch queuing times were very large. The maximum speedup achieved
+was approximately 5.6~x for 8 and 16 pilots on B\'eluga and 2.6~x on Cedar. 
+It is only in Configuration~3 (Figure~\ref{fig:beluga4})
+that we spot any large speedups, however, it generally remains that
+pilots are slightly slower, even in Configuration~3. The same can be said about Cedar, however,
+it occured with Configuration~4 instead of 3. On average, pilots were up to 0.11~x slower on B\'eluga and
+0.17~x slower on Cedar (see Table~\ref{table:speedup}), both of which occurred when using 16 pilots.
+The only times when pilots were found to be, on average, faster, was in Configuration~3 on B\'eluga
+and Configuration~4 on Cedar where the speedup was found to reach a max of 1.46 for 8 pilots on 
+B\'eluga.
 
@@
 
-The Cedar cluster (Figure~\ref{fig:makespanscedar}) also follows the same trend that
-the makespan of batch scheduled applications are generally faster than that of pilots,
-with a maximum speedup of 3, found both in Configuration~1 and 4. Moreover, in 
-Cedar, it was less clear whether to use 8 or 16 pilots as speedup between the two
-varies between the iterations. In general, both pilot configurations performed the
-same.
-
     \begin{figure*}
@@
 
-In order to determine whether queueing time was the cause of the difference 
-in makespans between batch and pilots \tristan{rather than other overheads?}, we compared makespan with queuing time of 
-first pilot. As can be seen in Figure~\ref{fig:mqcedar}, variations in overall
-makespans is primiarily due to queueing time. \tristan{I don't like these figures: if you want to show
-this, you should plot the makespan vs the queuing time of the first pilot, in an XY plot. But then the MW plot
-is better because it takes into account the queuing time of all pilots, not only the first one. I would remove this figure.}
-
-      \begin{figure*}
-        \centering
-        \begin{subfigure}[b]{0.475\textwidth}
-            \centering
-            \includegraphics[width=\textwidth]{figures/dedicated_4_cedar}
-            \caption[]%
-            {{\small Makespan}}
-            \label{fig:makespancedar}
-        \end{subfigure}
-        \hfill
-        \begin{subfigure}[b]{0.475\textwidth}
-            \centering
-            \includegraphics[width=\textwidth]{figures/pq_4_cedar}
-            \caption[]%
-            {{\small Queuing time}}
-            \label{fig:queuecedar}
-        \end{subfigure}
-        \caption[]
-        {\small Side-by-side comparison of total makespan and queueuing time of
-        first pilots or batch application for Configuration 4 running on Cedar}
-        \label{fig:mqcedar}
-    \end{figure*}
-
 In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
@@
 \end{itemize}
-\tristan{You should remove the ceil from the plot too.}
-As can be seen in the figure, although pilots and batch had similar makespans
-for a given number of average workers, pilots were consistently slower than batch
-for the same number of workers. Both batch and pilots were consistent with the 
-model, however \TG{there is a contradiction here: if both batch and pilots were consistent with the model, then
-they should have similar makespans.}. In general, both pilots and batch had the same number of average
-workers. \TG{You could also plot the difference in the number of workers, for each experiment.}
+As can be seen in the Figure~\ref{fig:mwall}, although pilots and batch had similar makespans
+and followed the general trend denoted by the model, pilots were consistently slower than batch
+for the same number of average workers. When comparing the average number of worker difference
+(Figures\ref{fig:nworkersbeluga} and \ref{fig:nworkerscedar}), it can be seen that pilot average
+number of workers start to exceed batch and the requirements increase. However, employing 16 pilots
+is generally always more costly than 8 pilots.
+
+    \begin{figure*}
+        \centering
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/nworkers_1_beluga}
+            \caption[]%
+            {{\small Configuration 1}}
+            \label{fig:nwbeluga1}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/nworkers_2_beluga}
+            \caption[]%
+            {{\small Configuration 2}}
+            \label{fig:nwbeluga2}
+        \end{subfigure}
+        \vskip\baselineskip
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/nworkers_3_beluga}
+            \caption[]%
+            {{\small Configuration 3}}
+            \label{fig:nwbeluga3}
+        \end{subfigure}
+        \quad
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/nworkers_4_beluga}
+            \caption[]%
+            {{\small Configuration 4}}
+            \label{fig:nwbeluga4}
+        \end{subfigure}
+        \caption[]
+        {\small The average worker difference on B\'eluga between batch and pilots for all configurations and repetitions}
+        \label{fig:nworkersbeluga}
+    \end{figure*}
+    \begin{figure*}
+        \centering
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/nworkers_1_cedar}
+            \caption[]%
+            {{\small Configuration 1}}
+            \label{fig:nwcedar1}
+        \end{subfigure}
+        \hfill
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/nworkers_2_cedar}
+            \caption[]%
+            {{\small Configuration 2}}
+            \label{fig:nwcedar2}
+        \end{subfigure}
+        \vskip\baselineskip
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/nworkers_3_cedar}
+            \caption[]%
+            {{\small Configuration 3}}
+            \label{fig:nwcedar3}
+        \end{subfigure}
+        \quad
+        \begin{subfigure}[b]{0.475\textwidth}
+            \centering
+            \includegraphics[width=\textwidth]{figures/nworkers_4_cedar}
+            \caption[]%
+            {{\small Configuration 4}}
+            \label{fig:nwcedar4}
+        \end{subfigure}
+        \caption[]
+        {\small The average worker difference on Cedar between batch and pilots for all configurations and repetitions}
+        \label{fig:makespanscedar}
+    \end{figure*}
+
+\begin{table}                                                                    
+    \centering                                                                       
+    \begin{tabular}{c|c|c|c|c}                                                             
+        {} & \multicolumn{2}{c}{B\'eluga} & \multicolumn{2}{c}{Cedar}\\
+    \rowcolor{headcolor}                                                             
+    Configuration & 8 pilots & 16 pilots & 8 pilots & 16 pilots\\                               
+    \hline                                                                           
+    1 & 0.949 & 0.932 & 0.724 & 0.874\\                                               
+    2 & 0.988 & 0.916 & 0.836 & 0.826\\                                               
+    3 & 1.458 & 1.369 & 0.941 & 0.964\\
+    4 & 0.970 & 0.891 & 1.049 & 1.068\\
+    \end{tabular}                                                                    
+    \setlength{\belowcaptionskip}{-10pt}                                             
+    \caption{Speedup of pilots}                                                    
+    \label{table:speedup}                                                            
+\end{table}
 
@@
 This can be due to a few reasons. The queuing time for batch and pilots were similar, 
-which can be an indication that the required resources to run the batch requests were
-available, and our priority on the cluster was what affected the queuing time \TG{I don't understand this, could you clarify?}. This may, in turn,
-explain why batch was sometimes significantly slower than pilots -- queueing time was not
-only controlled by priority, but also by available resources.
+which can be an indication that we were just lacking the right level of priority to 
+have been scheduled immediately. In the instances where batch was found to be significantly slower,
+it might have been then case that we had obtained the necessary priority level but there
+were not enough resources available to be scheduled.
 

### 2019-05-19 experiments/notebooks/generate_figures.py

@@ -109,9 +109,49 @@ def repetition_fig(data, n_dedicated, system="Beluga", save=None, ylim=None):
     if ylim is not None:
         ax.set_ylim(0, ylim)
     elif system == "beluga":
-        ax.set_ylim(0, 15000)
+        ax.set_ylim(0, 28000)
     else:
-        ax.set_ylim(0, 15000) # 60000
+        ax.set_ylim(0, 28000) # 60000
+
+    ax.set_xticks(index + bar_width / 2)
+    ax.set_xticklabels((i + 1 for i in range(n_groups)))
+    ax.legend()
+    
+    if save is None:
+        ax.set_title('{0} - {1} dedicated'.format(system, n_dedicated))
+        plt.show()
+    else:
+        plt.savefig(save)
+
+
+def nworkers_fig(data, n_dedicated, system="Beluga", save=None, ylim=None):
+    n_groups = len(data['batch'])
+    print(n_groups)
+
+    fig, ax = plt.subplots()
+
+    index = np.arange(n_groups)
+    bar_width = 0.30
+    opacity = 0.4
+
+    pilot8_diff = [elem - data['8pilots'][i] for i,elem in enumerate(data['batch']) if elem != 0 or data['8pilots'][i] != 0]
+    pilot16_diff = [elem - data['16pilots'][i] for i,elem in enumerate(data['batch']) if elem != 0 or data['16pilots'][i] != 0]
+    rects2 = ax.bar(index, pilot8_diff, bar_width,
+                    alpha=opacity, color='b', label='batch - 8 pilots')
+
+    rects3 = ax.bar(index + bar_width, pilot16_diff, bar_width,
+                    alpha=opacity, color='green', label='batch - 16 pilots')
+
+
+    ax.set_xlabel('Repetition')
+    ax.set_ylabel('Average number of worker difference (Batch - Pilot)')
+
+    if ylim is not None:
+        ax.set_ylim(0, ylim)
+    elif system == "beluga":
+        ax.set_ylim(-60, 50)
+    else:
+        ax.set_ylim(-60, 50) # 60000
 
     ax.set_xticks(index + bar_width / 2)
     ax.set_xticklabels((i + 1 for i in range(n_groups)))
@@ -200,8 +240,8 @@ def basic_model(xy1, xy2, xy3, xy4, num_pilots, xlabel, ylabel, system="Beluga",
[…]

### 2019-05-20 paper.tex

@@
     % maybe remove
-    \subsection{Pipeline robustness}
-        While nodes failures can occur for various reasons, underestimation of 
-        walltime may lead to early node termination in pilot scheduling systems.
-        Therefore, it is crucial for the overlay cluster within the pilot system
-        to be robust to node failures. In this series of experiments, we investigate
-        Spark's fault-tolerance with respect to worker, master and driver termination.
+    %\subsection{Pipeline robustness}
+    %    While nodes failures can occur for various reasons, underestimation of 
+    %    walltime may lead to early node termination in pilot scheduling systems.
+    %    Therefore, it is crucial for the overlay cluster within the pilot system
+    %    to be robust to node failures. In this series of experiments, we investigate
+    %    Spark's fault-tolerance with respect to worker, master and driver termination.
 
-        In our configuration, the driver is running in cluster deploy-mode, resulting
-        in it being executed directly on a worker. This feature alone should not make
-        it fault-tolerant to failures. However, using this feature coupled with the 
-        \texttt{--supervise} flag, will.
+    %   In our configuration, the driver is running in cluster deploy-mode, resulting
+    %    in it being executed directly on a worker. This feature alone should not make
+    %    it fault-tolerant to failures. However, using this feature coupled with the 
+    %    \texttt{--supervise} flag, will.
 
@@
 always entirely responsible for the difference. Sometimes there are errors related
-to workers not registering properly with the master. These occurrences lead to the 
-application being processed entirely with a smaller number of available workers. As the task
+to executors not properly starting. These occurrences lead to the 
+application being processed entirely with a smaller number of available executors. As the task
 delay added considers the maximum amount of parallel workers, functioning with less total workers
@@
 of parallelism is available at driver start time. Pilots, on the other hand, may start workers in parallel,
-but not necessarily all will be started at the same time. This means that the entire 
-workflow may suffer the impacts of starting workers in sequence \TG{isn't it queuing time again?}. Moreover, there would also
+but not necessarily all will be started at the same time (i.e. delays in starting the workers once resources
+have been allocated). This means that the entire 
+workflow may suffer the impacts of starting workers in sequence rather than in parallel. Moreover, there would also
 be a delay with respect to transferring data from workers to newly added workers. Therefore,
 it is only natural that pilots would have a bit more overhead than batch given the same
-queuing time should pilots not all commence at once \TG{ok, but isn't the total amount of transfered data the same in both conditions?}.
+queuing time should pilots not all commence at once, despite transferring the same amount of data.
  Furthermore, the pilot scripts also
@@
 there was at least one pilot alive, it would be possible to maintain the cluster and add additional pilots.
-Spark provides built-it fault-tolerance for not only the workers, but the master and driver as well.
+Spark provides built-it fault-tolerance for not only the workers, but the master and driver as well. However,
+it is also likely that with faster queuing times, all resources would be allocated at the same time. If walltime
+were to be underestimated, in this case, checkpointing and restarting the entire application from the last
+checkpoint would be the only option. This too has limitations as the last checkpoint may not be recent, requiring
+necessitating significant recomputation. It
+might even be desirable to have many small resource requests in such a scenario, as with an increase in pilots,
+it becomes less likely for all of them to be scheduled at once, as can be seen with 16 pilots. This would allow
+for more chances for the application to reach completiong without failing entirely.
 
-\tristan{Comment on the difficulties introduce by adding another layer in an already complex system:
-HPC + Spark. It complexifies debugging.}
+Many failures can be found, particularly with regards to the pilots. Many of these have been
+a result of Spark failing to start properly. The driver either would not return a submission ID after
+being started, or would remain in \texttt{SUBMITTED} status until the walltime expiration, having 
+not processed any data. As these error become fewer with increased resources, it is possible they are
+entirely related to lack of resources. Furthermore, it may also explain why batch submission always passes.
+Not only are all the workers that were requested available when the pilot begins running, batch has one extra
+executor available thanks to the driver process running directly on the process which had created it, rather than
+a worker process. Conversely, it is not only the pilot applications that fail, the batch applications
+do experience some failures as well. It can be seen in Figure~\ref{fig:makespansbeluga} that the batch application failed
+twice: once in Configuration~2 and the other in Configuration~4. In both cases, the reason for failure was the same: lack of
+available resources despite all executors having been assigned. These failures might be due to running on a cluster. Either
+the resources were down, or other users might have been using them, resulting in the resources being unavailable to our application.
+
+Running a Spark cluster atop of a Slurm allocation complexifies debugging, particularly if the application runs in cluster mode.
+Spark worker logs, unless saved to network storage, are unavailable. Saving them to network storage also effectively slows down
+the application as it is a slower storage than RAM and local disk. As a result, the Spark worker UI is also inaccessible. When running
+in client mode, the driver output ends up it standard output/error. However, in cluster mode, the driver output is found stored within
+the worker logs, which means the driver logs are also inaccessible without worker logs being written to network storage. Some HPC clusters
+may have a file limit for user, meaning that the user must be sure that these logs coupled without other stored data does not exceed the file
+limit.
 
@@
 \section{Conclusion}\label{sec:conclusion}
+Pilots do not provide significant advantage over batch systems using the same amount of resources. This is affected
+by the overhead of dynamically adjusting the cluster size and the fact that queuing time of the batch applications took
+as long as the pilots. These results, however, are to an extent cluster-dependent. It is expected that as the number
+of resources requested increases, the speedup provided by pilots will increase. Furthermore, users executing on a smaller cluster 
+may experience greater speedups using the same resource configuration.
+
+Pilots are meant to allow for underestimation of resources, particularly walltime, which would not
+be known to the user prior to execution. In its current state, \texttt{SPA} is not able to 
+evaluate that functionality. Further experiments on master fault-tolerance, driver recovery, and checkpointing
+will need to be conducted to determine if pilot jobs are even a viable solution for Spark applications running on 
+HPC.
+
+Currently, as client mode is not available in Spark Standalone clusters for PySpark applications,
+using pilots on HPC would be limited to non-Python applications until there is a solution for 
+Standalone mode.
 

### 2019-05-20 paper.tex
Edits and comments

@@
 \usepackage{stmaryrd}
+
 \algnewcommand\algorithmicforeach{\textbf{for each}}
@@
     commodity infrastructures. This paper evaluates the benefits of pilot
-    jobs over traditional batch submission on HPC clusters with overlay
-    Apache Spark clusters. We first evaluate the overall speedup brought on
-    by employing pilot jobs. We then examine the robustness to master and
-    driver failures, which may be frequent when underestimating pilot
-    walltime. Spark built-in checkpointing is also investigated in relation
-    to pilot expiration. Surprisingly, our results show that the speed-up
-    provided by pilot jobs over batch scheduling remains moderate (\TG{X} on
+    jobs over traditional batch submission for Apache Spark on HPC
+    clusters. Surprisingly, our results show that the speed-up provided by
+    pilot jobs over batch scheduling is moderate to inexistent (0.98 on
     average) despite the presence of long queuing times. In addition, pilot
@@
     and deployment. We conclude that traditional batch scheduling should
-    remain the default strategies to deploy Apache Spark applications on
+    remain the default strategy to deploy Apache Spark applications on
     HPC clusters.
@@
 heterogeneity, variable task queuing times, fine task granularity, and node
-failures on distributed computing infrastructures~\cite{jakubsmodel}. Made popular by software
+failures on distributed computing infrastructures. Made popular by software
 projects such as Condor~\cite{thain2005distributed} and DIRAC~\cite{casajus2010dirac}, they
@@
 requesting all the
-necessary resources and launching a standalone Spark cluster once the
+necessary resources and launching a Standalone Spark cluster once the
 resources have been allocated. This is for instance the method used in
-Compute Canada (see details \href{https://docs.computecanada.ca/wiki/Apache_Spark/en}{here}),
+Compute Canada,
  our national computing infrastructure. With pilot
@@
 Yet Another Resource Negotiator (YARN~\cite{apache13yet}), and Mesos~\cite{hindman2011mesos}. As a
-result of the sustained increase in data volumes, Apache Spark and other
+result of the sustained growth in data volumes, Apache Spark and other
 Big Data engines are increasingly used for scientific applications,
@@
     nodes with a total of 58,416 CPU cores. Available memory on a Cedar
-    node can range from 125 to 3022~G. Standard nodes are equipped with
+    node can range from 125 to 3022~GB. Standard nodes are equipped with
     either 2x Intel E5-2683 v4 Broadwell @ 2.1~Ghz (32 cores total) or 2x
     Intel Platinum 8160F Skylake @ 2.1Ghz (48 cores total) CPUs and 2 x
-    480~G SSD. All nodes and temporary storage on Cedar are connected by an
+    480~GB SSD. All nodes and temporary storage on Cedar are connected by an
     Intel OmniPath (version 1) with 100Gbit/s bandwidth.
@@
     B\'eluga, on the other hand, is a smaller cluster with 872 available
-    nodes. Node memory can range between 92 to 752~G, with the most common
+    nodes. Node memory can range between 92 to 752~GB, with the most common
     node type having 186G. All nodes contain 2 x Intel Gold 6148 Skylake @
@@
     100~Gb/s Mellanox Infiniband EDR network. Each non-GPU
-    node type contains one 480~G SSD.
+    node type contains one 480~GB SSD.
 
@@
     Conversely, Cedar had been operating for a few years, resulting in
-    higher usage and longer queuing times. Our experiments were repeated 15
+    higher usage and longer queuing times. We repeated our experiments multiple 
     times in each configuration to capture queuing time variability.
@@
     application specific. The driver, which can start in the client where it is launched (i.e. client deploy mode)
-    or directly on a worker node (i.e. cluster deploy mode), must request resources from the master
-    to launch its tasks. The master the requests workers with the necessary resources available to start
+    or directly on a worker node (i.e. cluster deploy mode) \TG{Deploy modes are explained in the next paragraph,
+    I'd remove them from here}, must request resources from the master
+    to launch its tasks. The master dispatches the requests to workers with the necessary resources available to start
     executors. The driver can then communicate with the executors to schedule and launch its tasks. Both the driver and
@@
 
-    There is not shortage of scripts available online for starting a Spark cluster on HPC~\footnote{\url{https://github.com/NIH-HPC/spark-slurm},}\footnote{\url{https://www.sherlock.stanford.edu/docs/software/using/spark/},}\footnote{\url{https://sparkhpc.readthedocs.io},}\footnote{\url{https://www.osc.edu/~troy/pbstools/man/pbs-spark-submit}}.
+    There is not shortage of scripts available online for starting a Spark cluster on HPC, 
+    see for instance \href{https://github.com/NIH-HPC/spark-slurm}{spark-slurm},
+    \href{https://www.sherlock.stanford.edu/docs/software/using/spark}{scripts} provided
+     by Stanford's Sherlock cluster, \href{https://sparkhpc.readthedocs.io}{sparkhpc},
+     or \href{https://www.osc.edu/~troy/pbstools/man/pbs-spark-submit}{pbs-spark-submit}.
     However, all these scripts start up a cluster within a single batch call. At most, there are two calls, one to startup the cluster,
     and the second, to start the application execution. None of these use any kind of pilot-scheduling approach, nor do they discuss the
-    effects of queuing time and how they might want to adjust it in the case where the driver is started separately
-    from the batch job to start the cluster. Even in the case of RADICAL-Pilot, as per some available tutorials for using the pilot-job application with
-    Spark~\footnote{\url{https://github.com/radical-cybertools/pilot-streaming/blob/master/examples/Pilot-Streaming-GettingStarted.ipynb},}\footnote{\url{https://github.com/radical-cybertools/MIDAS-tutorial/blob/master/pilot/Pilot-Spark.ipynb}}, it appears that the Spark cluster is started within a single pilot and the application is submitted, in a separate process, to that pilot.
-    Due to the fact that existing solutions mainly rely on a single batch call to startup the Spark
-    cluster and do not appear to be able to submit multiple smaller jobs to improve queuing times, we have decided to implement our own solution.
+    effects of queuing time and how they might want to adjust it in the
+    case where the driver is started separately from the batch job to start
+    the cluster. Even in the case of RADICAL-Pilot, as per some available
+    tutorials for using the pilot-job application with
+    Spark (see
+    \href{https://github.com/radical-cybertools/pilot-streaming/blob/master/examples/Pilot-Streaming-GettingStarted.ipynb}{here}
+    and \href{https://github.com/radical-cybertools/MIDAS-tutorial/blob/master/pilot/Pilot-Spark.ipynb}{here}),
+    it appears that the Spark cluster is started within a single pilot and
+    the application is submitted, in a separate process, to that pilot. Due
+    to the fact that existing solutions mainly rely on a single batch call
+    to startup the Spark cluster and do not appear to be able to submit
+    multiple smaller jobs to improve queuing times, we have decided to
+    implement our own solution for the sake of our experiments.
 
-    Two different job templates were used to implement the two main
+    Two different job templates were developed to implement the two main
     conditions compared in our experiments: the batch submission
@@
     \textit{SPA}. The templates are used in conjunction with JSON configuration 
-    file and passed to the SlurmPy~\cite{slurmpy} library within \textit{SPA}.
+    file and passed to the \href{https://github.com/brentp/slurmpy}{SlurmPy} library within \textit{SPA}.
     The \textit{SPA} application, all the while ensures that all is preconfigured
@@
     of a scientific Spark application. For this, we created a synthetic application 
-    that would process the Big Brain~\cite{amunts2013bigbrain}, a 76~G 3D histological
+    that would process the Big Brain~\cite{amunts2013bigbrain}, a 76~GB 3D histological
     image of a human brain. The algorithm is a chain of map transformations that
@@
     Investigating the fault-tolerance provided by the overlay cluster is
-    crucial in determining the suitability of pilot scheduling with overlay clusters.
+    crucial in determining the suitability of pilot scheduling with overlay clusters \TG{You should water this down if 
+    no fault tolerance experiment is presented}.
     Fault-tolerance of the driver is only possible in cluster deploy mode, however,
-    when using Spark's standalone scheduler, this mode is not available for Python 
+    when using Spark's Standalone scheduler, this mode is not available for Python 
     applications. It is for this reason that our synthetic application is written
@@
         \item $C$ is the total CPU time of the application
-        \item $W$ is the average number of Spark workers throughout the execution
+        \item $W$ is the \emph{average} number of Spark workers throughout the execution
     \end{itemize}
@@
           \rowcolor{headcolor}                                                           
-          execution mode & nodes/job & RAM (G) & CPUs per task & tasks/node & walltime & task delay (s) \\                             
+          Execution mode & Nodes/job & RAM (GB) & CPUs per task & Tasks/node & Walltime & Task delay (s) \\                             
           \hline
@@
        For batch, we requested 1 to 4 dedicated nodes, depending on resource 
-       configuration, each with 112~G of RAM, 1 core per task and 16 tasks per 
+       configuration, each with 112~GB of RAM, 1 core per task and 16 tasks per 
        node. On the other hand, for all configurations, we ran our experiments
@@
 
-       For each configuration, all three execution modes were executed in parallel
-       to ensure that the status of the overall cluster was the same when the 
-       different execution modes were launched. Furthermore, the order of the execution modes
-       for each configuration was randomized to ensure that Slurm job request
-       order could not have affected results.
+       For each configuration, all three execution modes (batch, 8 pilots,
+       16 pilots) were executed in parallel to ensure that the status of
+       the HPC cluster was the same when the different execution modes
+       were launched. Furthermore, the submission order of the execution
+       modes for each configuration was randomized to ensure that Slurm job
+       request order could not have affected results.
 
@@
 and Configuration~4 on Cedar where the speedup was found to reach a max of 1.46 for 8 pilots on 
-B\'eluga.
+B\'eluga. The missing bars in Figures~\ref{fig:makespansbeluga} and~\ref{fig:makespanscedar} correspond to 
+failed executions.
 
@@
     \item $W$ is the average number of Spark workers throughout the execution, computed as in Equation~\ref{eq:avgw}
-    \item $d$ is the task delay associated with a given configuration
+    \item $d$ is the task delay, in seconds, associated with a given configuration
     \item $20$ is the average measured incrementation duration, in seconds, for a BigBrain chunk
@@
 \end{itemize}
-As can be seen in the Figure~\ref{fig:mwall}, although pilots and batch had similar makespans
-and followed the general trend denoted by the model, pilots were consistently slower than batch
-for the same number of average workers. When comparing the average number of worker difference
-(Figures\ref{fig:nworkersbeluga} and \ref{fig:nworkerscedar}), it can be seen that pilot average
-number of workers start to exceed batch and the requirements increase. However, employing 16 pilots
-is generally always more costly than 8 pilots.
+As can be seen in the Figure~\ref{fig:mwall}, both pilots and batch 
+followed the general trend denoted by the model, which confirms that 
+the system was behaving without major external sources of
+overhead. However, pilots were consistently slower than batch for the same
+number of average workers. This is particularly visible in Configuration 4
+on B\'eluga (Fig. \ref{fig:mwbeluga}), and in Configuration 2, 3 and 4 on
+Cedar (Fig. \ref{fig:mwcedar}). It means that some sort of overhead
+impacted pilots but not batch. It can also be noticed that both batch and
+pilots occasionally deviated from the model line, in particular
+Configurations 2 and 3 in B\'eluga: this is due to the fact that our
+application is not a divisible load, as the model assumes. Extending the
+model beyond divisible loads is easy enough for batch and confirms that the
+observed deviations from the model line come from this assumption (see gray
+line in Fig.~\ref{fig:mwall} \TG{can you add the step line in gray? I
+realize that this is not the best explanation, but I don't have the
+expression for pilots. Also, I think this comment belongs here as it directly impacts
+the reading of the results, but it might as well be moved to the discussion if you want to keep
+this section short and sweet.}). The extended expression for pilots is more complex
+though.
+
+When
+comparing the average number of worker difference
+(Figures~\ref{fig:nworkersbeluga} and~\ref{fig:nworkerscedar}), it can be
+seen that (1) repetitions where pilots were faster than batch correspond
+to repetitions where pilots had more workers, and vice versa -- this
+confirms that performance differences are mainly coming from queuing time
+differences, and (2) pilot average number of workers start to exceed batch
+as the application requirements increase, that is, for Configuration 3 and
+4. However, employing 16 pilots
+is generally always more costly than 8 pilots \TG{The last sentence looks off-topic to me but maye I didn't understand.}.
 
@@
         \caption[]
-        {\small The average worker difference on B\'eluga between batch and pilots for all configurations and repetitions}
+        {\small  The difference in average workers on B\'eluga between batch and pilots for all configurations and repetitions. Positive values mean
+        that batch had more workers than pilots, that is, pilots did not improve queuing times.}
         \label{fig:nworkersbeluga}
@@
         \caption[]
-        {\small The average worker difference on Cedar between batch and pilots for all configurations and repetitions}
-        \label{fig:makespanscedar}
+        {\small The difference in average workers on Cedar between batch and pilots for all configurations and repetitions. Positive values mean
+        that batch had more workers than pilots, that is, pilots did not improve queuing times.}
+        \label{fig:nworkerscedar}
     \end{figure*}
@@
     \setlength{\belowcaptionskip}{-10pt}                                             
-    \caption{Speedup of pilots}                                                    
+    \caption{Speedup of pilots compared to batch}                                                    
     \label{table:speedup}                                                            
@@
 %Technical overhead of using pilots + an overlay cluster over a cluster.
-Pilots do not appear to provide any performance advantage over batch scheduling.
+Pilots do not appear to provide any performance advantage over batch
+scheduling to deploy Apache Spark overlay clusters on HPC.
 This can be due to a few reasons. The queuing time for batch and pilots were similar, 
@@
 delay added considers the maximum amount of parallel workers, functioning with less total workers
-will significantly affect makespan. However, as seen in Figure~\ref{fig:mwall}
+will significantly affect makespan. However, as seen in Figure~\ref{fig:mwall},
 pilots are slower even with the same amount of average workers. A potential reasoning for this
@@
 it becomes less likely for all of them to be scheduled at once, as can be seen with 16 pilots. This would allow
-for more chances for the application to reach completiong without failing entirely.
+for more chances for the application to reach completion without failing entirely.
 
-Many failures can be found, particularly with regards to the pilots. Many of these have been
-a result of Spark failing to start properly. The driver either would not return a submission ID after
-being started, or would remain in \texttt{SUBMITTED} status until the walltime expiration, having 
-not processed any data. As these error become fewer with increased resources, it is possible they are
-entirely related to lack of resources. Furthermore, it may also explain why batch submission always passes.
-Not only are all the workers that were requested available when the pilot begins running, batch has one extra
-executor available thanks to the driver process running directly on the process which had created it, rather than
-a worker process. Conversely, it is not only the pilot applications that fail, the batch applications
-do experience some failures as well. It can be seen in Figure~\ref{fig:makespansbeluga} that the batch application failed
-twice: once in Configuration~2 and the other in Configuration~4. In both cases, the reason for failure was the same: lack of
-available resources despite all executors having been assigned. These failures might be due to running on a cluster. Either
-the resources were down, or other users might have been using them, resulting in the resources being unavailable to our application.
+Many failures can be found, particularly with regards to the pilots. Many
+of these have been a result of Spark failing to start properly. The driver
+either would not return a submission ID after being started, or would
+remain in \texttt{SUBMITTED} status until the walltime expiration, having
+not processed any data. As these error become fewer with increased
+resources, it is possible they are entirely related to lack of resources.
+Furthermore, it may also explain why batch submission always passes. Not
+only are all the workers that were requested available when the pilot
+begins running, but batch also has one extra executor available thanks to the driver
+process running directly on the process which had created it, rather than on a
+worker process. Conversely, it is not only the pilot applications that
+fail, the batch applications do experience some failures as well. It can be
+seen in Figure~\ref{fig:makespansbeluga} that the batch application failed
+twice: once in Configuration~2 and the other in Configuration~4. In both
+cases, the reason for failure was the same: lack of available resources
+despite all executors having been assigned. These failures might be due to
+running on a shared HPC cluster. Either the resources were down, or other users might
+have been using them, resulting in the resources being unavailable to our
+application.
 
@@
 \section{Conclusion}\label{sec:conclusion}
-Pilots do not provide significant advantage over batch systems using the same amount of resources. This is affected
-by the overhead of dynamically adjusting the cluster size and the fact that queuing time of the batch applications took
-as long as the pilots. These results, however, are to an extent cluster-dependent. It is expected that as the number
-of resources requested increases, the speedup provided by pilots will increase. Furthermore, users executing on a smaller cluster 
-may experience greater speedups using the same resource configuration.
+Pilots do not provide significant advantage over batch systems to run
+Apache Spark applications on HPC clusters. This is affected by the overhead of dynamically
+adjusting the cluster size and the fact that queuing time of the batch
+applications took as long as the pilots. These results, however, are to an
+extent cluster-dependent. It is expected that as the number of resources
+requested increases, the speedup provided by pilots will increase.
+Furthermore, users executing on a smaller cluster may experience greater
+speedups using the same resource configuration. Our resource requirements
+and cluster size, however, match typical configurations currently available
+to scientific communities.
 
-Pilots are meant to allow for underestimation of resources, particularly walltime, which would not
-be known to the user prior to execution. In its current state, \texttt{SPA} is not able to 
-evaluate that functionality. Further experiments on master fault-tolerance, driver recovery, and checkpointing
-will need to be conducted to determine if pilot jobs are even a viable solution for Spark applications running on 
-HPC.
+Although obtained with Apache Spark, this result is likely to generalize to
+other implementations of overlay clusters. Conversely, our experiments do
+\emph{not} imply that pilot jobs are not useful on HPC clusters in the
+absence of overlay clusters. Most pilot job implementations in fact include
+an overlay scheduler, which allows for various types of optimization
+including dealing with fine task granularity or enforcing data locality.
+Pilot jobs naturally remain useful in that respect.
 
-Currently, as client mode is not available in Spark Standalone clusters for PySpark applications,
-using pilots on HPC would be limited to non-Python applications until there is a solution for 
-Standalone mode.
+Pilots are meant to allow for underestimation of resources, particularly
+walltime, which would not be known to the user prior to execution. In its
+current state, \texttt{SPA} is not able to evaluate that functionality.
+Further experiments on master fault-tolerance, driver recovery, and
+checkpointing will need to be conducted to determine if pilot jobs are even
+a viable solution for Spark applications running on HPC.
+
+Currently, as client mode is not available in Spark Standalone clusters for
+PySpark applications, using pilots on HPC would be limited to non-Python
+applications until there is a solution for Standalone mode.
 

### 2019-05-20 paper.tex
final modifications

@@
 necessary resources and launching a Standalone Spark cluster once the
-resources have been allocated. This is for instance the method used in
+resources have been allocated. This is, for instance, the method used on
 Compute Canada,
@@
 estimates. While there have been some efforts on implementing pilot jobs
-for Apache Spark~\cite{jha-spark-pbs}, research is limited and
+for Apache Spark~\cite{luckow2016hadoop}, research is limited and
 none of the them detail their quantitative effect.
@@
 including in neuroimaging, our primary field of
-interest~\cite{boubela2016big,mehta2017comparative,maybethesimulationone,freeman2014mapping}.
+interest~\cite{boubela2016big,mehta2017comparative,freeman2014mapping}.
 
@@
 in mind, and, with the exception of Dask~\cite{rocklin2015dask}, do not support batch
-HPC schedulers such as PBS, SGE, Slurm and TORQUE~\cite{schedulers}, which
+HPC schedulers such as the \href{http://www.univa.com/products/}{Oracle Grid Engine (OGE)},
+\href{https://slurm.schedmd.com/}{Slurm} and \href{https://www.adaptivecomputing.com/products/torque/}{TORQUE}, which
 are commonly available to scientists. Therefore, to run Big Data
@@
 through batch schedulers. Our experiments quantify the effect of pilot jobs
-when combined with such an overlay cluster. Our evaluation will be broken
-down into two parts:
-1) the added value of pilots, and 2), tolerance to failures.
+when combined with such an overlay cluster.
 
@@
     driver. The cluster manager, also known as the Spark master in Standalone mode, is responsible for the
-    resource provisioning within the cluster. Unlike the workers and driver, it is not
-    application specific. The driver, which can start in the client where it is launched (i.e. client deploy mode)
-    or directly on a worker node (i.e. cluster deploy mode) \TG{Deploy modes are explained in the next paragraph,
-    I'd remove them from here}, must request resources from the master
-    to launch its tasks. The master dispatches the requests to workers with the necessary resources available to start
-    executors. The driver can then communicate with the executors to schedule and launch its tasks. Both the driver and
+    resource provisioning within the cluster.  As it is the resource manager for a given cluster,
+    the cluster manager is not application specific. The driver requests resources from the master
+    to launch its tasks. The master then dispatches the requests to workers with
+    the necessary resources available to start
+    executors. The driver may subsequently communicate with the executors to schedule and launch its tasks. Both the driver and
     the executors are application specific and connected directly to the master.
@@
     driver client may require more resources than permitted. Moreover, due to cluster
-    security policies, it may not be possible, or difficult to execute the driver from
+    security policies, it may be difficult or impossible, to execute the driver from
     an external computer, such as a personal laptop. Therefore, for both batch and pilot
-    scheduling, it is necessary to launch the driver from within the Slurm jobs, that is, in cluster mode.
+    scheduling, it is necessary to launch the driver from within the Slurm jobs. That is, in cluster mode.
 
-    Spark also provides master fault tolerance through single-node recovery, where a shared
+    Spark also provides master fault tolerance through single-node recovery, wherever a shared
     filesystem is available. To enable this, a recovery folder must be set. Should the master
@@
 
-    In long running applications, it is necessary to be able to recovery the application execution from
+    In long running applications, it is necessary to be able to recover application execution from
     last successful state. Spark provides a checkpointing mechanism in order to ensure that the application
@@
 
-    There is not shortage of scripts available online for starting a Spark cluster on HPC, 
+    There is no shortage of scripts available online for starting a Spark cluster on HPC, 
     see for instance \href{https://github.com/NIH-HPC/spark-slurm}{spark-slurm},
@@
      or \href{https://www.osc.edu/~troy/pbstools/man/pbs-spark-submit}{pbs-spark-submit}.
-    However, all these scripts start up a cluster within a single batch call. At most, there are two calls, one to startup the cluster,
-    and the second, to start the application execution. None of these use any kind of pilot-scheduling approach, nor do they discuss the
+    However, all these scripts start up a cluster within a single batch call. At most, there are two calls, one for
+    cluster startup,
+    and the other, to launch the application. None of these use any kind of pilot-scheduling approach, nor do they discuss the
     effects of queuing time and how they might want to adjust it in the
@@
 
-    The pilot submission template is similar to that of the batch template, however,
+    The pilot submission template is similar to that of the batch template, although,
     each pilot will start its own Spark master and worker. However, there will only be one
@@
     which each pilot starts its own master is to ensure the fault tolerance of the 
-    masters. In this sort of configuration, should the active master be killed, one
-    of the stand-by masters will takeover and the application will be able to 
-    resume. Such a configuration is particularly favourable in pilot scheduling 
+    masters. In this configuration, should the active master be killed, one
+    of the stand-by masters can takeover and the application may be able to 
+    resume if single-node recovery is set. Such a configuration is particularly favourable in pilot scheduling 
     scenarios as node failures may be more frequent due to walltime expiration.
@@
     % have shorter walltimes, to see if queuing times are shorter too.}
-    \TG{You could add a figure showing the driver, workers, master and stand-by masters, to make the system 
-    description clearer.}
 
@@
 
-    Investigating the fault-tolerance provided by the overlay cluster is
-    crucial in determining the suitability of pilot scheduling with overlay clusters \TG{You should water this down if 
-    no fault tolerance experiment is presented}.
+    Spark cluster fault-tolerance is important in determining the suitability of
+    pilot-jobs for Spark applications on HPC. Executing jobs on a shared cluster may
+    result in a variety of failures. Pilot jobs may be more likely to fail as a result of
+    underestimation of resources (e.g. walltimes) and therefore should be fault-tolerant
+    to master, worker and driver failures. Although built-in fault-tolerance was not
+    investigated in this paper, we want to ensure that the environment is set up in such
+    a way that Spark's fault-tolerance configuration could easily be set should our 
+    experiments return favourable results for pilot-jobs.
     Fault-tolerance of the driver is only possible in cluster deploy mode, however,
@@
 
-       There were 15 repetitions in total for each configuration, and the order
+       There were 10 repetitions in total for each configuration, and the order
        in which the configurations were launched was also randomized. This was to 
@@
        Different clusters may have different resource configurations, number of 
-       users and scheduling policies. Therefore, we executed all 15 repetitions on
+       users and scheduling policies. Therefore, we executed all 10 repetitions on
        both Cedar and B\'eluga to determine how much our results differ between
@@
 was approximately 5.6~x for 8 and 16 pilots on B\'eluga and 2.6~x on Cedar. 
-It is only in Configuration~3 (Figure~\ref{fig:beluga4})
+It is only in Configuration~3 (Figure~\ref{fig:beluga4}), on B\'eluga,
 that we spot any large speedups, however, it generally remains that
@@
 In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
-are used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
+were used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
 given an experiment configuration, where the line corresponds to the estimated makespan using 
@@
 observed deviations from the model line come from this assumption (see gray
-line in Fig.~\ref{fig:mwall} \TG{can you add the step line in gray? I
-realize that this is not the best explanation, but I don't have the
-expression for pilots. Also, I think this comment belongs here as it directly impacts
-the reading of the results, but it might as well be moved to the discussion if you want to keep
-this section short and sweet.}). The extended expression for pilots is more complex
+line in Fig.~\ref{fig:mwall}. The extended expression for pilots is more complex
 though.
@@
 as the application requirements increase, that is, for Configuration 3 and
-4. However, employing 16 pilots
-is generally always more costly than 8 pilots \TG{The last sentence looks off-topic to me but maye I didn't understand.}.
+4.
 
 \begin{figure*}
+
         \centering
@@
     \setlength{\belowcaptionskip}{-10pt}                                             
-    \caption{Speedup of pilots compared to batch}                                                    
+    \caption{Average speedup of pilots for each configuration}                                                    
     \label{table:speedup}                                                            
@@
 Pilots do not appear to provide any performance advantage over batch
-scheduling to deploy Apache Spark overlay clusters on HPC.
+scheduling when deploying Apache Spark overlay clusters on HPC.
 This can be due to a few reasons. The queuing time for batch and pilots were similar, 
@@
 have been scheduled immediately. In the instances where batch was found to be significantly slower,
-it might have been then case that we had obtained the necessary priority level but there
+it might have been case that we had obtained the necessary priority level but there
 were not enough resources available to be scheduled.
@@
 were to be underestimated, in this case, checkpointing and restarting the entire application from the last
-checkpoint would be the only option. This too has limitations as the last checkpoint may not be recent, requiring
-necessitating significant recomputation. It
-might even be desirable to have many small resource requests in such a scenario, as with an increase in pilots,
-it becomes less likely for all of them to be scheduled at once, as can be seen with 16 pilots. This would allow
+checkpoint would be the only option. This too has limitations as the last checkpoint may not be recent,
+necessitating significant amounts of recomputation. It
+might even be desirable to have many small resource requests in such a scenario, as it becomes less likely for all pilots to start at once as they become more numerous, as can be seen with 16 pilots when compared to 8.
+This would allow
 for more chances for the application to reach completion without failing entirely.
 
-Many failures can be found, particularly with regards to the pilots. Many
+Many failures can be found, particularly with regards to the pilots. Most
 of these have been a result of Spark failing to start properly. The driver
@@
 remain in \texttt{SUBMITTED} status until the walltime expiration, having
-not processed any data. As these error become fewer with increased
+not processed any data. As these errors become fewer with increased
 resources, it is possible they are entirely related to lack of resources.
-Furthermore, it may also explain why batch submission always passes. Not
+Furthermore, it may also explain why batch submission is more frequently successful. Not
 only are all the workers that were requested available when the pilot
@@
 Running a Spark cluster atop of a Slurm allocation complexifies debugging, particularly if the application runs in cluster mode.
-Spark worker logs, unless saved to network storage, are unavailable. Saving them to network storage also effectively slows down
-the application as it is a slower storage than RAM and local disk. As a result, the Spark worker UI is also inaccessible. When running
-in client mode, the driver output ends up it standard output/error. However, in cluster mode, the driver output is found stored within
-the worker logs, which means the driver logs are also inaccessible without worker logs being written to network storage. Some HPC clusters
-may have a file limit for user, meaning that the user must be sure that these logs coupled without other stored data does not exceed the file
+Spark worker logs, unless saved to network storage, become unavailable after program execution.
+Although the worker UI is available during program execution, a user must manually create an SSH tunnel
+to each worker node to gain access to them.
+An option is to save the to a shared network storage for future access. This effectively slows down
+the application as the network FS is typically a slower storage than local disk and RAM.
+
+When running in client mode, the driver output ends up it standard output/error. 
+However, in cluster mode, the driver output is found stored within
+the worker logs, which means the driver logs are also inaccessible without 
+worker logs being written to network storage. Some HPC clusters
+may have a file limit for users, meaning that the user must be sure that these 
+logs coupled without other stored data does not exceed the file
 limit.
-
-\TG{You could end the discussion by a comment on the current literature on pilot jobs. How are your results
-consistent or not with the literature?}
 

### 2019-05-21 paper.tex
A few fixes

@@
 observed deviations from the model line come from this assumption (see gray
-line in Fig.~\ref{fig:mwall}. The extended expression for pilots is more complex
+line in Fig.~\ref{fig:mwall}). The extended expression for pilots is more complex
 though.
@@
 An option is to save the to a shared network storage for future access. This effectively slows down
-the application as the network FS is typically a slower storage than local disk and RAM.
+the application as the network file system is typically a slower storage than local disk and RAM.
 

### 2019-05-21 experiments/notebooks/generate_figures.py
fixed figure generation script

@@ -197,6 +197,7 @@ def makespan_box(d1, d2, d3, d4, num_pilots, system="Beluga", save=None):
 def basic_model(xy1, xy2, xy3, xy4, num_pilots, xlabel, ylabel, system="Beluga", save=None):   
     fig, ax = plt.subplots()
     execution_mode = "Batch" if type(num_pilots) == str and "batch" in num_pilots else "Pilots" 
+    
     ax.scatter(xy1[0], xy1[1], c="#6600cc", alpha=0.5, label="{} 1 dedicated".format(execution_mode))                                                  
     ax.scatter(xy2[0], xy2[1], c="#ff0000", alpha=0.5, label="{} 2 dedicated".format(execution_mode))
     ax.scatter(xy3[0], xy3[1], c="#ffa500", alpha=0.5, label="{} 3 dedicated".format(execution_mode))
@@ -225,6 +226,8 @@ def basic_model(xy1, xy2, xy3, xy4, num_pilots, xlabel, ylabel, system="Beluga",
                                      label='16 Pilots')
         lgnd = [ pilot8_symbol, pilot16_symbol ]
 
+    model_symbol = mlines.Line2D([], [], color="gray", alpha=0.5, linestyle='-',
+                              label='Batch model, non divisible load')
     batch_symbol = mlines.Line2D([], [], color="black", alpha=0.5, marker='o', linestyle='None',
                                  label=execution_mode)
     d1_symbol = mlines.Line2D([], [], color="purple", alpha=0.5, linestyle='-',
@@ -237,7 +240,7 @@ def basic_model(xy1, xy2, xy3, xy4, num_pilots, xlabel, ylabel, system="Beluga",
                               label='Configuration 4')
 
     lgnd.insert(0, batch_symbol)
-    lgnd.extend([d1_symbol, d2_symbol, d3_symbol, d4_symbol])
+    lgnd.extend([model_symbol, d1_symbol, d2_symbol, d3_symbol, d4_symbol])
     ax.legend(handles=lgnd)
     
     get_c = lambda x, y, z=False : (10*(x+20)*125)/y if z else 10*(x+20)*np.ceil(125/y)
@@ -249,7 +252,13 @@ def basic_model(xy1, xy2, xy3, xy4, num_pilots, xlabel, ylabel, system="Beluga",
     ax.plot(n_workers, get_c(90, n_workers, ceil_results), '-', c='red', alpha=a)
[…]

### 2019-05-21 paper.tex
final fixes

@@
 were used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
-given an experiment configuration, where the line corresponds to the estimated makespan using 
+given an experiment configuration, where the coloured line corresponds to the estimated makespan using 
 our performance model. Makespan for the model was calculated as:

### 2019-05-21 experiments/notebooks/generate_figures.py

@@ -227,7 +227,7 @@ def basic_model(xy1, xy2, xy3, xy4, num_pilots, xlabel, ylabel, system="Beluga",
         lgnd = [ pilot8_symbol, pilot16_symbol ]
 
     model_symbol = mlines.Line2D([], [], color="gray", alpha=0.5, linestyle='-',
-                              label='Batch model, non divisible load')
+                              label='Batch model, non DL')
     batch_symbol = mlines.Line2D([], [], color="black", alpha=0.5, marker='o', linestyle='None',
                                  label=execution_mode)
     d1_symbol = mlines.Line2D([], [], color="purple", alpha=0.5, linestyle='-',
@@ -570,20 +570,22 @@ def main():
             if elem != 0 and data[1][i] != 0:
                 w.append(elem)
                 m.append(data[1][i])
-        data = (w, m)
+        return (w, m)
 
-    keep_successful(batchmw_1d) 
-    keep_successful(batchmw_2d) 
-    keep_successful(batchmw_3d) 
-    keep_successful(batchmw_4d) 
-    keep_successful(pilots8mw_1d) 
-    keep_successful(pilots8mw_2d) 
-    keep_successful(pilots8mw_3d) 
-    keep_successful(pilots8mw_4d) 
-    keep_successful(pilots16mw_1d) 
-    keep_successful(pilots16mw_2d) 
-    keep_successful(pilots16mw_3d) 
-    keep_successful(pilots16mw_4d) 
+    batchmw_1d = keep_successful(batchmw_1d) 
+    batchmw_2d = keep_successful(batchmw_2d) 
+    batchmw_3d = keep_successful(batchmw_3d) 
+    batchmw_4d = keep_successful(batchmw_4d) 
+    pilots8mw_1d = keep_successful(pilots8mw_1d) 
+    pilots8mw_2d = keep_successful(pilots8mw_2d) 
+    pilots8mw_3d = keep_successful(pilots8mw_3d) 
+    pilots8mw_4d = keep_successful(pilots8mw_4d) 
+    pilots16mw_1d = keep_successful(pilots16mw_1d) 
+    pilots16mw_2d = keep_successful(pilots16mw_2d) 
+    pilots16mw_3d = keep_successful(pilots16mw_3d) 
+    pilots16mw_4d = keep_successful(pilots16mw_4d) 
+
+    print(pilots8mw_1d)
 
     basic_model(batchmw_1d + pilots8mw_1d,
                 batchmw_2d + pilots8mw_2d,

### 2019-08-31 paper.tex

@@
     clusters. Surprisingly, our results show that the speed-up provided by
-    pilot jobs over batch scheduling is moderate to inexistent (0.98 on
+    pilot jobs over batch scheduling is moderate to non-existent (0.98 on
     average) despite the presence of long queuing times. In addition, pilot
-    jobs provide an extra layer of scheduling that complexifies debugging
+    jobs provide an extra layer of scheduling that complicates debugging
     and deployment. We conclude that traditional batch scheduling should
@@
 resources can thus be created, shielding applications from the underlying
-queuing times, failures, and other idiosyncracies. A variety of frameworks
+queuing times, failures, and other idiosyncrasies. A variety of frameworks
 now rely on pilot jobs, including RADICAL-Pilot~\cite{merzky2015radical},
@@
 applications with Apache Spark~\cite{zaharia2016apache} on shared HPC
-clusters. The method currently recommended for this purpose involves batch
+clusters. The currently recommended method for this purpose involves batch
 requesting all the
@@
 deployments, we hypothesize that this strategy would reduce queuing times
-by (1) fragmenting resource requirements, and (2) allowing shorter walltime
+by (1) fragmenting resource requirements, and (2) allowing shorter wall-time
 estimates. While there have been some efforts on implementing pilot jobs
@@
 to execute iterative workflows -- something not easily doable in older
-frameworks. Schedulers for Spark include its built-in standalone schedule,
+frameworks. Schedulers for Spark include its built-in Standalone scheduler,
 Yet Another Resource Negotiator (YARN~\cite{apache13yet}), and Mesos~\cite{hindman2011mesos}. As a
@@
     nodes. Node memory can range between 92 to 752~GB, with the most common
-    node type having 186G. All nodes contain 2 x Intel Gold 6148 Skylake @
+    node type having 186~GB. All nodes contain 2 x Intel Gold 6148 Skylake @
     2.4~Ghz (40 cores/node) CPU and are connected to each other with a
@@
 
-    It is important to note that these clusters were used in production,
-    that is, concurrently with other users. This allowed us to test the
+    It is important to note that these clusters were used in production.
+    That is, concurrently with other users. This allowed us to test the
     added-value of pilot scheduling in different realistic conditions of queuing
@@
 
-    There are three possible cluster managers available to use in Spark applications:
+    There are three possible cluster managers available to use with Spark applications:
     the Standalone cluster manager, YARN and Mesos. The Standalone cluster manager is the
-    most basic cluster manager available and is packaged directly with Spark. YARN is a
-    resource manager designed for the resource management of Hadoop applications. However,
+    most basic cluster manager available and is packaged directly with Spark. The 
+    YARN resource manager is designed for the resource management of Hadoop applications, however,
     it can be used with different types of applications as well. In contrast, Mesos was 
@@
     to be used as an HPC scheduler. Due to Standalone's lightweight nature and our intent
-    to start Spark clusters per application, we focus on Standalone mode.
+    to start Spark clusters per application, we utilized the Standalone cluster manager.
 
@@
     is deployed on a machine not co-located in the cluster to reduce network 
-    latency between workers and driver. The cluster deploy mode also has the \texttt{supervise}
+    latency between workers and the driver. The cluster deploy mode also has the \texttt{supervise}
     feature which allows the driver to be restarted in case of failure.
     This is particularly useful for
-    pilot-based overlay clusters where the running driver may fail due to walltime expiration.
+    pilot-based overlay clusters where the running driver may fail due to wall-time expiration.
     Using the Standalone cluster manager, client mode is supported in all APIs, whereas cluster mode is
@@
     The template operates as follows: certain resource requirements are requested by
-    the user (e.g. walltime, amount of memory per node, number of CPUs per task, number 
+    the user (e.g. wall-time, amount of memory per node, number of CPUs per task, number 
     of nodes and number of tasks per node). Once these resources are allocated, a 
@@
     deploy mode over cluster mode as driver recovery would not be required in batch in the
-    case of walltime expiration, as all the allocated resources will expire at the same time.
+    case of wall-time expiration, as all the allocated resources will expire at the same time.
     Moreover, many scientific applications are written in Python, which cannot use cluster mode using
@@
     pilot which will start the driver process. The pilot selected to start the driver is the first one
-    that attempts to do so by way of lockfile.
+    that attempts to do so by means of a lockfile.
     The reason for 
@@
     resume if single-node recovery is set. Such a configuration is particularly favourable in pilot scheduling 
-    scenarios as node failures may be more frequent due to walltime expiration.
+    scenarios as node failures may be more frequent due to wall-time expiration.
     Additionally, the Spark deploy mode of the driver was selected to be cluster deploy mode.
@@
     with the masters, it is particularly important to have a fault-tolerant
-    driver in pilot-scheduling scenarios due to possible walltime
+    driver in pilot-scheduling scenarios due to possible wall-time
     expiration.
     Should pilots be idle for a certain duration, the
-    pilots will terminate themselves such as to not hog resources.
+    pilots will terminate themselves in order to not hog resources.
     % \TG{It should be explained before that one of the interests
-    % of pilots is to ``play'' with shorter walltimes}. \TG{It would also be
+    % of pilots is to ``play'' with shorter wall-times}. \TG{It would also be
     % nice to have runs where pilots 
-    % have shorter walltimes, to see if queuing times are shorter too.}
+    % have shorter wall-times, to see if queuing times are shorter too.}
 
     Although we start a master on each pilot, we have not incorporated the master recovery function
-    for our experiments. This feature is, however, important and should be incorporated in pilot schedulers
-    were any nodes can fail by design.
+    for our experiments. However, this feature is important and should be incorporated in pilot schedulers
+    where any nodes can fail by design.
 
@@
     in cluster deploy, which pilot uses, the driver occupies a worker, whereas in
-    client deploy, the driver is separate from any worker.
+    client deploy, the driver is a distinct process.
 
@@
     file and passed to the \href{https://github.com/brentp/slurmpy}{SlurmPy} library within \textit{SPA}.
-    The \textit{SPA} application, all the while ensures that all is preconfigured
+    The \textit{SPA} application ensures that all is preconfigured
     correctly before passing it to SlurmPy. It also ensures that enough pilots are
@@
         \end{algorithm}
-    To determine the added value of pilot scheduling over batch scheduling of Spark
+    To determine the added value of pilot over batch scheduling of Spark
     applications, we required a Spark application operating on a large dataset with
@@
     of a scientific Spark application. For this, we created a synthetic application 
-    that would process the Big Brain~\cite{amunts2013bigbrain}, a 76~GB 3D histological
-    image of a human brain. The algorithm is a chain of map transformations that
-    at each transformation increment the voxels of the image by 1 (see Algorithm~\ref{alg:incrementation}).
-    We chose such a synthetic algorithm
+    that would process the 40~$\mu$m BigBrain~\cite{amunts2013bigbrain}, a 76~GB 3D histological
+    image of a human brain. The algorithm is a chain of map transformations that,
+    at each transformation, increment the voxels of the image by 1 (see Algorithm~\ref{alg:incrementation}).
+    We selected a synthetic algorithm
     as the focus of our experiments is pilot scheduling and not the application in
-    itself. Furthermore, this algorithm enabled us to have control over the task duration
-    which was representative of scientific applications. Additionally, it was important
+    and of itself. Furthermore, this algorithm enabled us to have control over the task duration and make it representative of scientific applications. Additionally, it was important
     that the overall application duration did not vary between the different levels of
@@
     result in a variety of failures. Pilot jobs may be more likely to fail as a result of
-    underestimation of resources (e.g. walltimes) and therefore should be fault-tolerant
+    underestimation of resources (e.g. wall-times), and therefore, should be fault-tolerant
     to master, worker and driver failures. Although built-in fault-tolerance was not
@@
     applications. It is for this reason that our synthetic application is written
-    in Scala. Nevertheless, cluster deploy mode is possible for Python application
+    in Scala. Nevertheless, cluster deploy mode is available for PySpark applications
     using YARN or Mesos schedulers.
@@
         \item $M$ is the makespan of the application
-        \item $C$ is the total CPU time of the application
+        \item $C$ is the total wall-time of the application
         \item $W$ is the \emph{average} number of Spark workers throughout the execution
@@
 
-    Therefore, assuming a fixed total CPU time, the relation
+    Therefore, assuming a fixed total wall-time, the relation
     between batch and pilot jobs can be represented as:
@@
         resource queue longer than multiple pilot requests that ultimately use up
-        the same amount of resources as each individual pilot requests less resources at
+        the same amount of resources, as each individual pilot requests fewer resources at
         a time. We therefore used four resource configurations to investigate this
@@
 
-       As the focus of our experiments here is to measure the impact pilot scheduling
-       has on queueing times, we wanted to ensure that our walltime estimates would
+       As the focus of our experiments is to measure the impact pilot scheduling
+       has on queueing times, we wanted to ensure that our wall-time estimates would
        be consistent for each experimental condition. Therefore, we adjusted task 
-       delay based on the maximum level of parallelism to ensure that walltimes would
+       delay based on the maximum level of parallelism to ensure that wall-times would
        not need to be readjusted for each configuration. Given utmost parallelism,
@@
        addition to variability in application parallelism due to queuing times of each pilot Slurm job, 
-       we have set the walltime to be 2h30 to ensure experiment completion.
+       we set the Slurm allocation wall-time to be 2h30 to ensure experiment completion.
 
@@
     %    While nodes failures can occur for various reasons, underestimation of 
-    %    walltime may lead to early node termination in pilot scheduling systems.
+    %    wall-time may lead to early node termination in pilot scheduling systems.
     %    Therefore, it is crucial for the overlay cluster within the pilot system
@@
 on either system. In B\'eluga and Cedar, 16 pilots were generally slower than batch except for a 
-few occurrencs. When 16 pilots were faster, it was generally either very slight, or, as expected,
+few occurrences. When 16 pilots were faster, it was generally either very slight, or, as expected,
 when batch queuing times were very large. The maximum speedup achieved
 was approximately 5.6~x for 8 and 16 pilots on B\'eluga and 2.6~x on Cedar. 
-It is only in Configuration~3 (Figure~\ref{fig:beluga4}), on B\'eluga,
+It is only in Configuration~3 (Figure~\ref{fig:beluga4}) on B\'eluga
 that we spot any large speedups, however, it generally remains that
-pilots are slightly slower, even in Configuration~3. The same can be said about Cedar, however,
-it occured with Configuration~4 instead of 3. On average, pilots were up to 0.11~x slower on B\'eluga and
+pilots are slightly slower, even in this configuration. The same can be said about Cedar, however,
+it occurred with Configuration~4 instead of 3. On average, pilots were up to 0.11~x slower on B\'eluga and
 0.17~x slower on Cedar (see Table~\ref{table:speedup}), both of which occurred when using 16 pilots.
@@
 model beyond divisible loads is easy enough for batch and confirms that the
-observed deviations from the model line come from this assumption (see gray
+observed deviations from the model line come from this assumption (see grey
 line in Fig.~\ref{fig:mwall}). The extended expression for pilots is more complex
@@
 node has enough memory to for all batch and pilot configuration, therefore, this could not
-have affected queuing time for Cedar.
+have affected queuing times in Cedar.
 
 Although queuing time is largely responsible for makespan variations, it is not
-always entirely responsible for the difference. Sometimes there are errors related
+always entirely responsible for the differences. Sometimes there are errors related
 to executors not properly starting. These occurrences lead to the 
 application being processed entirely with a smaller number of available executors. As the task
-delay added considers the maximum amount of parallel workers, functioning with less total workers
+delay added assumes the maximum number of parallel workers, functioning with less total workers
 will significantly affect makespan. However, as seen in Figure~\ref{fig:mwall},
@@
 over after the application has already started. In batch scheduling, all workers are started in
-parallel and the data is transferred automatically at the beginning of the pipeline as the maximum amount
+parallel and the data is transferred automatically at the beginning of the pipeline, as the maximum amount
 of parallelism is available at driver start time. Pilots, on the other hand, may start workers in parallel,
@@
 investigated the queuing time differences with respect to differences in available resources,
-the walltimes were kept static. Had walltime been underestimated to the point where the Slurm jobs
+the wall-times were kept static. Had wall-time been underestimated to the point where the Slurm jobs
 would terminate prior to application completion, pilots might have been preferable to batch
@@
 cluster would be shutdown and need to restart, which has additional overheads. With pilots, as long as 
-there was at least one pilot alive, it would be possible to maintain the cluster and add additional pilots.
-Spark provides built-it fault-tolerance for not only the workers, but the master and driver as well. However,
-it is also likely that with faster queuing times, all resources would be allocated at the same time. If walltime
+there is at least one pilot alive, it would be possible to maintain the cluster and add additional pilots.
+Spark provides built-in fault-tolerance for not only the workers, but the master and driver as well. However,
+it is also likely that with faster queuing times, all resources would be allocated at the same time. If wall-time
 were to be underestimated, in this case, checkpointing and restarting the entire application from the last
@@
 either would not return a submission ID after being started, or would
-remain in \texttt{SUBMITTED} status until the walltime expiration, having
+remain in \texttt{SUBMITTED} status until the wall-time expiration, having
 not processed any data. As these errors become fewer with increased
@@
 
-Running a Spark cluster atop of a Slurm allocation complexifies debugging, particularly if the application runs in cluster mode.
+Running a Spark cluster atop of a Slurm allocation complicates debugging, particularly if the application runs in cluster mode.
 Spark worker logs, unless saved to network storage, become unavailable after program execution.
@@
 An option is to save the to a shared network storage for future access. This effectively slows down
-the application as the network file system is typically a slower storage than local disk and RAM.
+the application as the network file system is typically a slower storage than local storage.
 
-When running in client mode, the driver output ends up it standard output/error. 
+When running in client mode, the driver output ends up in standard output/error. 
 However, in cluster mode, the driver output is found stored within
@@
 Pilots are meant to allow for underestimation of resources, particularly
-walltime, which would not be known to the user prior to execution. In its
+wall-time, which would not be known to the user prior to execution. In its
 current state, \texttt{SPA} is not able to evaluate that functionality.

### 2019-08-31 paper.tex

@@
 \begin{abstract}
-    Big Data has become prominent throughout many scientific fields and, as
+    Big Data has become prominent throughout many scientific fields, and as
     a result, scientific communities have sought out Big Data frameworks to
@@
     driver client may require more resources than permitted. Moreover, due to cluster
-    security policies, it may be difficult or impossible, to execute the driver from
+    security policies, it may be difficult or impossible to execute the driver from
     an external computer, such as a personal laptop. Therefore, for both batch and pilot
@@
 
-    Spark also provides master fault tolerance through single-node recovery, wherever a shared
+    Spark also provides master fault tolerance through single-node recovery wherever a shared
     filesystem is available. To enable this, a recovery folder must be set. Should the master
@@
         \item $M$ is the makespan of the application
-        \item $C$ is the total wall-time of the application
+        \item $C$ is the total execution time of the application
         \item $W$ is the \emph{average} number of Spark workers throughout the execution
@@
 
-    Therefore, assuming a fixed total wall-time, the relation
+    Therefore, assuming a fixed total execution time, the relation
     between batch and pilot jobs can be represented as:
@@
 while both pilot configurations could be placed in B\'eluga's low memory nodes, batch requests
-had to be place on B\'eluga's medium memory nodes. B\'eluga has 172 low memory nodes and
+had to be placed on B\'eluga's medium memory nodes. B\'eluga has 172 low memory nodes and
 516 mid memory nodes. As pilots would get priority for the low memory nodes first and 
@@
 investigated the queuing time differences with respect to differences in available resources,
-the wall-times were kept static. Had wall-time been underestimated to the point where the Slurm jobs
+the wall-times were kept static. Had wall-times been underestimated to the point where the Slurm jobs
 would terminate prior to application completion, pilots might have been preferable to batch
@@
 Spark provides built-in fault-tolerance for not only the workers, but the master and driver as well. However,
-it is also likely that with faster queuing times, all resources would be allocated at the same time. If wall-time
+it is also likely that with faster queuing times, all resources would be allocated at the same time. If wall-times
 were to be underestimated, in this case, checkpointing and restarting the entire application from the last
@@
 worker process. Conversely, it is not only the pilot applications that
-fail, the batch applications do experience some failures as well. It can be
+fail; batch applications do experience some failures as well. It can be
 seen in Figure~\ref{fig:makespansbeluga} that the batch application failed
